---
title: "Psychometrics with the IRW: course map"
---

```{r}
source("lessons/_course.R")
course <- read_course()
check_course(course)
# Links on this page are grey for stubs, the site's link colour otherwise.
map_link <- function(l) {
  link <- lesson_link(l, from = "root")
  if (l$status == "stub") paste0(link, "{.lesson-stub}") else link
}
```

```{=html}
<div style="border:3px solid #c2410c;border-radius:8px;background:#fff7ed;padding:1.2rem 1.5rem;margin:1.2rem 0 1.8rem 0">
  <div style="font-size:2.2rem;font-weight:800;letter-spacing:0.12em;color:#c2410c;line-height:1">DRAFT</div>
  <div style="margin-top:0.6rem;font-size:1.05rem">
    This course is under construction and is not ready for teaching. Most lessons are
    still outlines or stubs, the few drafted ones will change, and nothing here has been
    checked by anyone but its authors. Please don't cite, assign, or circulate it yet.
    Comments are very welcome: <a href="mailto:bdomingue@stanford.edu">bdomingue@stanford.edu</a>.
  </div>
</div>
```

These are stand-alone lessons on psychometrics, each built around open data from the
[Item Response Warehouse](https://itemresponsewarehouse.org). You can teach them in
sequence as a full course or pick out individual lessons. The lessons come in three
tranches:

- **Preliminary** lessons set up the course.
- **Core** lessons make a first course in psychometrics.
- **Extensions** go further.

Each lesson lists what it assumes and what builds on it. The lessons grew out of
EDUC 252 (Introduction to Psychometrics) at Stanford. The source (lessons, code and data notes) is on
[GitHub](https://github.com/ben-domingue/irw-course); comments and issues are welcome there.

The lessons were written largely by Claude (Anthropic), an AI model, working from the
EDUC 252 materials and under Ben Domingue's direction. Ben reviews each lesson before
it is marked done; until then, read its text, code and interpretations as a draft.

## Why psychometrics?

Much of what we want to know about people hinges on attributes we cannot observe directly:
how a child's executive functioning changes with age, whether an intervention reduces
anxiety, how reading develops across school. What we observe instead are responses to
items: a tap on the correct arrow, "most days" on a survey, a wrong answer to question 7.
Psychometrics is the discipline that turns those responses into measurements and tells us
how far to trust them.

Consider a ruler. It measures a child's height regardless of that child's background; a
different ruler gives a reading with the same meaning; and a well-chosen ruler measures
quickly across the whole range of heights we care about. Psychological measures rarely do
all of this, but the ruler suggests what to aim for
([Domingue, 2026](https://doi.org/10.1146/annurev-devpsych-111124-041520)):

- Measurements should be insensitive to differences in attributes other than the one being
  measured.
- Measurements from different instruments should be comparable.
- Measurement should deliver precise information about the attribute as quickly as
  possible.

None of these comes for free. Suppose we measure executive functioning with a tablet task.
Children with less tablet experience may look worse for reasons that have nothing to do with
executive functioning. The task that suits a four-year-old is not the one that suits a
ten-year-old, yet we want to talk about growth across those ages. And time with a child is
short, so every item has to earn its place. Psychometrics has tools for each problem:
checking whether items work the same way across groups
([Differential item functioning](lessons/dif.qmd)), putting scores from different forms on a
common scale ([Linking and equating](lessons/equating.qmd)), and choosing items to fit the
respondent ([Item banks and adaptive testing](lessons/item-banks-cat.qmd)).

The common thread is a probabilistic model for how people respond to items. Such a model can
be falsified, so we can interrogate whether it works in a given setting; when it does, it
gives a compact description of how an instrument functions, one that can be reused with new
people, new forms and new ages.

Psychological measurement is hard, and whether it deserves the name "measurement" at all has
been argued over for most of a century ([What is measurement?](lessons/measurement.qmd)).
Interval scales do not come easily ([Scale properties](lessons/scale-properties.qmd)). But
the limits of our measures are limits on what we can learn about the attributes themselves.
Psychometrics is how we find out where those limits are, and how we push them back.


## Lessons by module

Lessons in blue have a draft to read. Lessons in grey are still stubs: a placeholder page
with no content yet.

```{=html}
<style>
.module-grid { display: grid; grid-template-columns: repeat(2, 1fr); gap: 1rem; margin-bottom: 1.5rem; }
@media (max-width: 768px) { .module-grid { grid-template-columns: 1fr; } }
.module-card { border: 1px solid #dee2e6; border-left: 4px solid #2780e3; border-radius: 6px; padding: 1rem 1.3rem; background: #fff; }
.module-card h3 { font-size: 1.05rem; margin-top: 0; }
.module-card ol { margin-bottom: 0; padding-left: 1.2rem; }
.lesson-tranche { font-size: 0.72rem; font-weight: 600; text-transform: uppercase; letter-spacing: 0.05em; color: #b45309; }
.lesson-status { font-size: 0.72rem; font-weight: 600; text-transform: uppercase; letter-spacing: 0.05em; color: #6c757d; }
/* Stubs have a page but no content yet: grey them out so drafts stand out. */
a.lesson-stub { color: #adb5bd; }
a.lesson-stub:hover { color: #6c757d; }
</style>
```

```{r}
#| output: asis
cat('::: {.module-grid}\n')
for (m in course$modules) {
  ls <- Filter(function(l) l$module == m$id, course$lessons)
  cat('::: {.module-card}\n### ', m$title, '\n\n', sep = '')
  for (i in seq_along(ls)) {
    st <- paste0(if (ls[[i]]$tranche != 'core') sprintf(' [%s]{.lesson-tranche}', ls[[i]]$tranche) else '',
                 if (ls[[i]]$status != "done") sprintf(' [%s]{.lesson-status}', ls[[i]]$status) else '')
    cat(i, '. ', map_link(ls[[i]]), st, '\n', sep = '')
  }
  cat('\n:::\n\n')
}
cat(':::\n')
```

## Build your own course

Build a course from the lesson bank: drag lessons into your course (or use Add),
put them in any order, and group them under unit or week headings. Start from a
preset or from nothing. Nothing is blocked. The notes flag a lesson that comes
before a lesson it assumes, or that assumes one you left out, and offer a fix you can
take or ignore. Open "Dependencies and threads" under a lesson in the bank to see
what it needs, what it leads to, and the threads it starts and picks up. The
lessons, prerequisites and threads all come from `lessons.yml`.

```{r}
# The builder reads lessons.yml through `course`; nothing in it is hand-maintained.
# Passed as a JSON string so one-element lists stay arrays.
ojs_define(cb_course = as.character(jsonlite::toJSON(list(
  modules = lapply(course$modules, function(m) list(id = m$id, title = m$title)),
  lessons = lapply(course$lessons, function(l) list(
    id = l$id, title = l$title, module = l$module, tranche = l$tranche,
    status = l$status, prereqs = I(as.character(unlist(l$prereqs))))),
  paths = lapply(course$paths, function(p) list(
    id = p$id, title = p$title, lessons = I(as.character(unlist(p$lessons))))),
  threads = lapply(course$threads, function(t) list(
    id = t$id, idea = t$idea, introduced = t$introduced,
    returns = lapply(t$returns, function(r) list(lesson = r$lesson, how = r$how))))
), auto_unbox = TRUE)))
```

```{ojs}
//| echo: false
//| output: false
import {courseBuilder} from "./lessons/widgets/course-builder.js"
Sortable = require("sortablejs@1")  // drag and drop, with touch
```

```{ojs}
//| echo: false
courseBuilder(JSON.parse(cb_course), {Sortable})
```

## Paths through the course

A path is one order for teaching the lessons. The builder's "A first course" and
"EDUC 252" presets start from these lessons. Add your own path to `lessons.yml`.

```{r}
#| output: asis
for (p in course$paths) {
  cat("### ", p$title, "\n\n", sep = "")
  for (i in seq_along(p$lessons))
    cat(i, ". ", map_link(lesson_by_id(p$lessons[[i]], course)), "\n", sep = "")
  cat("\n")
}
```
