---
title: "Psychometrics with the IRW: course map"
---

```{r}
source("lessons/_course.R")
course <- read_course()
check_course(course)
# Links on this page are grey for stubs, the site's link colour otherwise.
map_link <- function(l) {
  link <- lesson_link(l, from = "root")
  if (l$status == "stub") paste0(link, "{.lesson-stub}") else link
}
```

```{=html}
<div style="border:3px solid #c2410c;border-radius:8px;background:#fff7ed;padding:1.2rem 1.5rem;margin:1.2rem 0 1.8rem 0">
  <div style="font-size:2.2rem;font-weight:800;letter-spacing:0.12em;color:#c2410c;line-height:1">DRAFT</div>
  <div style="margin-top:0.6rem;font-size:1.05rem">
    This course is under construction and is not ready for teaching. Most lessons are
    still outlines or stubs, the few drafted ones will change, and nothing here has been
    checked by anyone but its authors. Please don't cite, assign, or circulate it yet.
    Comments are very welcome: <a href="mailto:bdomingue@stanford.edu">bdomingue@stanford.edu</a>.
  </div>
</div>
```

These are stand-alone lessons on psychometrics, each built around open data from the
[Item Response Warehouse](https://itemresponsewarehouse.org). You can teach them in
sequence as a full course or pick out individual lessons. The lessons come in three
tranches: *preliminary* lessons set up the course, *core* lessons make a first course in
psychometrics, and *extensions* go further. Each lesson lists what it
assumes and what builds on it. The lessons grew out of EDUC 252 (Introduction to
Psychometrics) at Stanford.

The lessons were written largely by Claude (Anthropic), an AI model, working from the
EDUC 252 materials and under Ben Domingue's direction. Ben reviews each lesson before
it is marked done; until then, read its text, code and interpretations as a draft.


## Lessons by module

Lessons in blue have a draft to read. Lessons in grey are still stubs: a placeholder page
with no content yet.

```{=html}
<style>
.module-grid { display: grid; grid-template-columns: repeat(2, 1fr); gap: 1rem; margin-bottom: 1.5rem; }
@media (max-width: 768px) { .module-grid { grid-template-columns: 1fr; } }
.module-card { border: 1px solid #dee2e6; border-left: 4px solid #2780e3; border-radius: 6px; padding: 1rem 1.3rem; background: #fff; }
.module-card h3 { font-size: 1.05rem; margin-top: 0; }
.module-card ol { margin-bottom: 0; padding-left: 1.2rem; }
.lesson-tranche { font-size: 0.72rem; font-weight: 600; text-transform: uppercase; letter-spacing: 0.05em; color: #b45309; }
.lesson-status { font-size: 0.72rem; font-weight: 600; text-transform: uppercase; letter-spacing: 0.05em; color: #6c757d; }
/* Stubs have a page but no content yet: grey them out so drafts stand out. */
a.lesson-stub { color: #adb5bd; }
a.lesson-stub:hover { color: #6c757d; }
</style>
```

```{r}
#| output: asis
cat('::: {.module-grid}\n')
for (m in course$modules) {
  ls <- Filter(function(l) l$module == m$id, course$lessons)
  cat('::: {.module-card}\n### ', m$title, '\n\n', sep = '')
  for (i in seq_along(ls)) {
    st <- paste0(if (ls[[i]]$tranche != 'core') sprintf(' [%s]{.lesson-tranche}', ls[[i]]$tranche) else '',
                 if (ls[[i]]$status != "done") sprintf(' [%s]{.lesson-status}', ls[[i]]$status) else '')
    cat(i, '. ', map_link(ls[[i]]), st, '\n', sep = '')
  }
  cat('\n:::\n\n')
}
cat(':::\n')
```

## How the lessons connect

An arrow from A to B means B assumes A. Extensions (beyond a first course) have a dashed
border. Click a lesson to open it.

```{r}
#| output: asis
node <- setNames(sprintf("L%02d", seq_along(course$lessons)),
                 vapply(course$lessons, `[[`, "", "id"))
# One colour per module (no subgraphs: they force a wide, unreadable layout).
pal <- c("#dbeafe", "#dcfce7", "#ccfbf1", "#fef9c3", "#fce7f3", "#ede9fe", "#ffedd5", "#e0f2fe")
lines <- "flowchart TB"
for (i in seq_along(course$modules)) {
  m <- course$modules[[i]]
  lines <- c(lines, sprintf("  classDef %s fill:%s,stroke:#6c757d,color:#212529", m$id, pal[i]))
  for (l in Filter(function(l) l$module == m$id, course$lessons))
    lines <- c(lines, sprintf('  %s["%s"]:::%s', node[[l$id]], gsub('"', "'", l$title), m$id))
  # Extensions: dashed border.
  lines <- c(lines, sprintf("  classDef %s_opt fill:%s,stroke:#6c757d,stroke-dasharray:5 3,color:#495057", m$id, pal[i]))
}
for (l in Filter(function(l) l$tranche == 'extension', course$lessons))
  lines <- c(lines, sprintf("  class %s %s_opt", node[[l$id]], l$module))
for (l in course$lessons) for (p in unlist(l$prereqs))
  lines <- c(lines, sprintf("  %s --> %s", node[[p]], node[[l$id]]))
for (l in course$lessons)
  lines <- c(lines, sprintf('  click %s href "lessons/%s.html"', node[[l$id]], l$id))
cat("```{mermaid}\n", paste(lines, collapse = "\n"), "\n```\n\n", sep = "")
cat(paste(sprintf('<span style="background:%s;border:1px solid #6c757d;border-radius:3px;padding:1px 8px;margin-right:6px;font-size:0.85rem">%s</span>',
                  pal[seq_along(course$modules)], vapply(course$modules, `[[`, "", "title")), collapse = " "), "\n")
```

## Paths through the course

A path is one order for teaching the lessons. Add your own to `lessons.yml`.

```{r}
#| output: asis
for (p in course$paths) {
  cat("### ", p$title, "\n\n", sep = "")
  for (i in seq_along(p$lessons))
    cat(i, ". ", map_link(lesson_by_id(p$lessons[[i]], course)), "\n", sep = "")
  cat("\n")
}
```
