---
title: "Psychometrics with the IRW: course map"
---

```{r}
source("lessons/_course.R")
course <- read_course()
check_course(course)
```

These are stand-alone lessons on psychometrics, each built around open data from the
[Item Response Warehouse](https://itemresponsewarehouse.org). You can teach them in
sequence as a full course or pick out individual lessons. Each lesson lists what it
assumes and what builds on it. The lessons grew out of EDUC 252 (Introduction to
Psychometrics) at Stanford.

::: {.callout-warning appearance="simple"}
**Work in progress.** The lesson list and prerequisites are a first draft, and most
lessons are still stubs.
:::

## Lessons by module

```{=html}
<style>
.module-grid { display: grid; grid-template-columns: repeat(2, 1fr); gap: 1rem; margin-bottom: 1.5rem; }
@media (max-width: 768px) { .module-grid { grid-template-columns: 1fr; } }
.module-card { border: 1px solid #dee2e6; border-left: 4px solid #2780e3; border-radius: 6px; padding: 1rem 1.3rem; background: #fff; }
.module-card h3 { font-size: 1.05rem; margin-top: 0; }
.module-card ol { margin-bottom: 0; padding-left: 1.2rem; }
.lesson-status { font-size: 0.72rem; font-weight: 600; text-transform: uppercase; letter-spacing: 0.05em; color: #6c757d; }
</style>
```

```{r}
#| output: asis
cat('::: {.module-grid}\n')
for (m in course$modules) {
  ls <- Filter(function(l) l$module == m$id, course$lessons)
  cat('::: {.module-card}\n### ', m$title, '\n\n', sep = '')
  for (i in seq_along(ls)) {
    st <- if (ls[[i]]$status != "done") sprintf(' [%s]{.lesson-status}', ls[[i]]$status) else ''
    cat(i, '. ', lesson_link(ls[[i]], from = 'root'), st, '\n', sep = '')
  }
  cat('\n:::\n\n')
}
cat(':::\n')
```

## How the lessons connect

An arrow from A to B means B assumes A. Click a lesson to open it.

```{r}
#| output: asis
node <- setNames(sprintf("L%02d", seq_along(course$lessons)),
                 vapply(course$lessons, `[[`, "", "id"))
# One colour per module (no subgraphs: they force a wide, unreadable layout).
pal <- c("#dbeafe", "#dcfce7", "#fef9c3", "#fce7f3", "#ede9fe", "#ffedd5", "#e0f2fe")
lines <- "flowchart TB"
for (i in seq_along(course$modules)) {
  m <- course$modules[[i]]
  lines <- c(lines, sprintf("  classDef %s fill:%s,stroke:#6c757d,color:#212529", m$id, pal[i]))
  for (l in Filter(function(l) l$module == m$id, course$lessons))
    lines <- c(lines, sprintf('  %s["%s"]:::%s', node[[l$id]], gsub('"', "'", l$title), m$id))
}
for (l in course$lessons) for (p in unlist(l$prereqs))
  lines <- c(lines, sprintf("  %s --> %s", node[[p]], node[[l$id]]))
for (l in course$lessons)
  lines <- c(lines, sprintf('  click %s href "lessons/%s.html"', node[[l$id]], l$id))
cat("```{mermaid}\n", paste(lines, collapse = "\n"), "\n```\n\n", sep = "")
cat(paste(sprintf('<span style="background:%s;border:1px solid #6c757d;border-radius:3px;padding:1px 8px;margin-right:6px;font-size:0.85rem">%s</span>',
                  pal[seq_along(course$modules)], vapply(course$modules, `[[`, "", "title")), collapse = " "), "\n")
```

## Paths through the course

A path is one order for teaching the lessons. Add your own to `lessons.yml`.

```{r}
#| output: asis
for (p in course$paths) {
  cat("### ", p$title, "\n\n", sep = "")
  for (i in seq_along(p$lessons))
    cat(i, ". ", lesson_link(lesson_by_id(p$lessons[[i]], course), from = "root"), "\n", sep = "")
  cat("\n")
}
```
