---
title: "Estimating abilities: MLE and EAP"
---

```{r}
#| output: asis
source("_course.R")
lesson_header("ability-estimation")
```

```{webr}
#| context: setup
options(mc.cores = 1)
```

```{ojs}
//| echo: false
import {grid, loglik, palette} from "./widgets/irt.js"
import {wsB, logLikR, mleR, seR, wleR, mapR, posteriorR} from "./widgets/ability-estimation.js"
import {quiz, predict} from "./widgets/quiz.js"
```

## What this is for

A respondent answers all ten items of a vocabulary test correctly. What is their ability? The likelihood, which has given us every estimate so far, says it is larger than any number you care to name. Abilities come to us only through an estimator, and the estimators disagree most for the respondents a test measures worst, at the top and the bottom. With the item parameters known, this lesson starts from maximum likelihood, shows where it fails, brings in a prior to get the expected a posteriori (EAP) estimate, and fits four estimators to the Wordsum vocabulary test, where 12% of respondents have no maximum-likelihood estimate at all.

## Goals

By the end of this lesson you should be able to:

- write down the likelihood for one respondent's $\theta$ with the item parameters known, find its maximum, and give its standard error;
- explain why an all-correct or all-wrong pattern has no maximum-likelihood estimate, and what each alternative estimator does about it;
- compute an EAP estimate on a grid of nodes, and explain why it is pulled toward the mean and by how much;
- say what the MLE, WLE, MAP and EAP each assume, and choose among them for reporting scores to individuals and for comparing groups;
- show that under the Rasch model every one of these estimates is a function of the sum score, and that under the 2PL it isn't.

## Core ideas

### One respondent, one unknown

::: {.callout-tip appearance="simple"}
**Recall** from [The Rasch model](rasch.qmd): `mirt`, like most software, estimates the item parameters first and then estimates abilities treating them as known.
:::

This lesson is the second step. With the item parameters known, a respondent's answers are independent Bernoulli trials whose probabilities depend on one unknown, $\theta$: a logistic regression with the respondent's items as the observations ([Lord, 1980](https://doi.org/10.4324/9780203056615)). For the 2PL,
$$\ell(\theta) = \sum_{i} \Big[x_i \log P_i(\theta) + (1 - x_i)\log\big(1 - P_i(\theta)\big)\Big], \qquad \frac{\partial \ell}{\partial \theta} = \sum_i a_i\big(x_i - P_i(\theta)\big),$$
and the Rasch model sets every $a_i = 1$.

::: {.callout-tip appearance="simple"}
**Recall** from [The likelihood and logistic regression](likelihood.qmd): at the maximum the derivative of the log likelihood (the score equation) is zero, and the curvature gives the standard error. With an intercept, the score equation says the predicted number of successes equals the observed number.
:::

Here $\theta$ is that intercept. Under the Rasch model the maximum-likelihood estimate (MLE) solves $\sum_i P_i(\hat\theta) = r$: the expected number correct equals the observed number. Software solves it by Newton–Raphson, stepping from a guess by the ratio of the derivative to the curvature ([Baker & Kim, 2004](https://doi.org/10.1201/9781482276725)).

::: {.callout-tip appearance="simple"}
**Recall** from [Information, precision, and short forms](information.qmd): the negative curvature of the log likelihood is the test information, $I(\theta) = \sum_i a_i^2 P_i(1 - P_i)$, and the conditional standard error of measurement is $SE(\theta) = 1/\sqrt{I(\theta)}$. For one respondent's estimate, evaluate it at their $\hat\theta$.
:::

So the standard error of the MLE is $1/\sqrt{I(\hat\theta)}$, a large-sample approximation that holds up better in the middle of a test than at its ends.

#### Try it: the likelihood of a response pattern

Three Rasch items ($b$ = −1, 0, 1.5). Choose which the respondent gets right; the bold curve is that pattern's likelihood, the pale ones the other seven.

```{ojs}
//| echo: false
lp_b = [-1, 0, 1.5]
lp_labels = lp_b.map((b, i) => `Item ${i + 1} (b = ${b})`.replace("-", "−"))
viewof lp_x = Inputs.checkbox(lp_labels, {label: "Correct on", value: [lp_labels[0], lp_labels[1]]})
```

```{ojs}
//| echo: false
lp_pattern = lp_labels.map(l => lp_x.includes(l) ? 1 : 0)
lp_r = lp_pattern.reduce((s, v) => s + v, 0)
lp_all = [0, 1, 2, 3, 4, 5, 6, 7].map(m => lp_b.map((_, i) => (m >> i) & 1))
lp_th = grid(-5, 5, 301)
lp_curves = lp_all.flatMap(x => lp_th.map(t => ({theta: t, L: Math.exp(loglik(t, x, lp_b)), pat: x.join(""),
  mine: x.join("") === lp_pattern.join("")})))
lp_mle = mleR(lp_r, lp_b)
Plot.plot({
  width: 700, height: 300, x: {label: "Ability θ", domain: [-5, 5]}, y: {label: "Likelihood", domain: [0, 1]},
  marks: [
    Plot.ruleX(lp_b, {stroke: palette.rule}),
    Plot.line(lp_curves.filter(d => !d.mine), {x: "theta", y: "L", z: "pat", stroke: palette.light, strokeWidth: 1.2}),
    Plot.line(lp_curves.filter(d => d.mine), {x: "theta", y: "L", stroke: palette.main, strokeWidth: 3}),
    ...(Number.isFinite(lp_mle) ? [Plot.dot([{theta: lp_mle, L: Math.exp(loglik(lp_mle, lp_pattern, lp_b))}], {x: "theta", y: "L", r: 6, fill: palette.contrast})] : [])
  ]
})
```

```{ojs}
//| echo: false
md`Pattern **${lp_pattern.join("-")}** (sum score ${lp_r}). ${Number.isFinite(lp_mle) ?
  `The likelihood peaks at θ = **${lp_mle.toFixed(2)}**, where the information is ${(1 / seR(lp_mle, lp_b) ** 2).toFixed(2)} and the standard error 1 / √${(1 / seR(lp_mle, lp_b) ** 2).toFixed(2)} = **${seR(lp_mle, lp_b).toFixed(2)}**.` :
  `The likelihood keeps ${lp_r === 0 ? "rising as θ falls" : "rising as θ grows"}: it has no peak, so there is no maximum-likelihood estimate.`} Item difficulties are the grey lines.`
```

The default pattern, 1-1-0, peaks at 1.02, with a standard error of 1.37: three items locate a respondent only roughly. The other two-right patterns peak at the same place, though lower (we come back to this at the end). Now tick all three: the likelihood climbs toward 1 and never turns over.

### Perfect patterns have no maximum

Why does the all-correct pattern never peak? Each item's probability of a correct answer rises with $\theta$, so their product does too. The score equation asks the expected number correct to equal the number of items, which it approaches but never reaches. The MLE is $+\infty$, and for an all-wrong pattern $-\infty$.

::: {.callout-tip appearance="simple"}
**Recall** from [The likelihood and logistic regression](likelihood.qmd), its Go deeper callout: if a predictor perfectly separates successes from failures, the logistic regression MLE doesn't exist ([Albert & Anderson, 1984](https://doi.org/10.1093/biomet/71.1.1)). It named a respondent who answers every item correctly as the item-response version.
:::

It is the coin with ten heads in ten flips: $\hat p = 1$, whose log odds is infinite. `mirt` reports $\pm$`Inf`; a general optimizer given bounds, such as `optimize()` on $[-10, 10]$, returns the bound, a number that looks like an estimate (the simulation below shows it). How often it happens depends on the test and the respondents: short tests, and items far from many respondents, produce many.

```{ojs}
//| echo: false
quiz({
  q: "A respondent answers all three items in the widget correctly. You give them a fourth item, much harder ($b = 3$), and they get it wrong. What happens to their maximum-likelihood estimate?",
  options: [
    {text: "It becomes finite", correct: true, why: "With 3 of 4 right, the score equation $\\sum_i P_i(\\theta) = 3$ has a solution: the expected score climbs from 0 to 4 and passes 3 on the way. One wrong answer is enough for the likelihood to turn over. (Here the estimate is 2.49.)"},
    {text: "It stays at $+\\infty$, because the respondent still got every easy item right", correct: false, why: "Under the Rasch model only the sum score matters, and 3 of 4 is not a perfect score. The wrong answer makes very large values of $\\theta$ less likely, so the likelihood peaks."},
    {text: "It becomes finite, and it must be below 3, the new item's difficulty", correct: false, why: "It does become finite (2.49 here), but nothing ties it to the new item's $b$. Had all four items sat at $b = 3$, 3 of 4 would give $3 + \\log 3 \\approx 4.1$."}
  ]
})
```

::: {.callout-note collapse="true"}
## Go deeper: why perfect patterns have no MLE, and what keeps the alternatives finite

**Assumptions:** one respondent answers $I$ dichotomous items under the 2PL (the Rasch model is $a_i = 1$), with the item parameters known and responses locally independent given $\theta$. Write $P_i = \text{logit}^{-1}\big(a_i(\theta - b_i)\big)$.

**No maximum.** If every $x_i = 1$, the log likelihood is $\ell(\theta) = \sum_i \log P_i(\theta)$, and
$$\frac{d\ell}{d\theta} = \sum_i a_i\,\big(1 - P_i(\theta)\big) > 0 \quad\text{for every } \theta.$$
The log likelihood is strictly increasing, bounded above by 0, and approaches 0 as $\theta \to \infty$. It has a least upper bound but no maximum. The all-wrong pattern is the mirror image. This is separation in the sense of [Albert and Anderson (1984)](https://doi.org/10.1093/biomet/71.1.1): with $\theta$ as the only parameter, a pattern with no variation is perfectly separated.

**The WLE stays finite.** [Warm (1989)](https://doi.org/10.1007/bf02294627) adds a term to the score equation that removes the first-order bias of the MLE. For the Rasch model it is
$$\sum_i (x_i - P_i) + \frac{J(\theta)}{2I(\theta)} = 0, \qquad J(\theta) = \sum_i P_i(1 - P_i)(1 - 2P_i),\quad I(\theta) = \sum_i P_i(1 - P_i).$$
For a perfect score the first term is $\sum_i (1 - P_i)$, which falls to 0 as $\theta$ grows, while the second tends to $-1/2$: as every $P_i \to 1$, $1 - 2P_i \to -1$ and $J \to -I$. As $\theta$ falls, the first term tends to $I$, the number of items, and the second to $+1/2$. The left-hand side is positive at the bottom of the scale and negative at the top, so it has a root: the WLE is finite.

**The EAP and MAP stay finite.** With a normal prior, the log posterior is $\ell(\theta) - (\theta - \mu)^2/(2\sigma^2)$ plus a constant. The first term is bounded above by 0 and the second falls without limit, so the log posterior has a maximum (the MAP). The posterior is a bounded function times a normal density, so it has a finite mean (the EAP).

**Why it matters later.** Joint maximum likelihood, which estimates items and respondents together, has to set perfect patterns aside before it starts ([Estimating item parameters](item-estimation.qmd)). An adaptive test estimates $\theta$ after one or two items, when all-correct patterns are the rule ([Item banks and adaptive testing](item-banks-cat.qmd)). And a reported score has to say something about everyone at the ceiling ([What does a score mean?](score-meaning.qmd)).
:::

### Bring in a prior: the EAP estimate

The likelihood lesson set the Bayesian reading aside for this lesson. In it, $\theta$ gets a **prior** distribution, $g(\theta)$, and the data turn it into a **posterior**,
$$p(\theta \mid \mathbf{x}) = \frac{L(\theta \mid \mathbf{x})\,g(\theta)}{\int L(t \mid \mathbf{x})\,g(t)\,dt}.$$
For ability, the prior has a natural source: the distribution of $\theta$ in the respondent's population, which `mirt` estimates along with the items (a normal with mean 0 and an SD fixed at 1 or estimated). The **expected a posteriori** (EAP) estimate is the posterior mean, and its standard error is the posterior SD ([Bock & Mislevy, 1982](https://doi.org/10.1177/014662168200600405)). The integral has no closed form, and it doesn't need one: evaluate likelihood × prior at a grid of nodes $t_1, \dots, t_Q$ and take a weighted average,
$$\hat\theta_{\text{EAP}} = \frac{\sum_q t_q\,L(t_q)\,g(t_q)}{\sum_q L(t_q)\,g(t_q)}.$$
Nothing iterates and nothing fails: a normal prior times a bounded likelihood always has a mean, even when the likelihood never peaks. The posterior's peak is the **maximum a posteriori** (MAP) estimate, also called Bayes modal ([Mislevy, 1986](https://doi.org/10.1007/bf02293979)).

#### Try it: prior × likelihood = posterior

The ten Wordsum items analysed below, with their Rasch difficulties. Choose a sum score and the prior's SD (the fitted value is 1.88); each curve is scaled to a peak of 1.

```{ojs}
//| echo: false
viewof pp_r = Inputs.range([0, 10], {value: 9, step: 1, label: "Sum score (of 10)"})
viewof pp_sd = Inputs.range([0.3, 4], {value: 1.88, step: 0.01, label: "Prior SD"})
```

```{ojs}
//| echo: false
pp = posteriorR(pp_r, wsB, 0, pp_sd, grid(-7, 7, 561))
pp_ll = pp.nodes.map(t => logLikR(t, pp_r, wsB))
pp_llmax = Math.max(...pp_ll)
pp_pmax = Math.max(...pp.post)
pp_data = pp.nodes.flatMap((t, k) => [
  {theta: t, y: Math.exp(pp_ll[k] - pp_llmax), curve: "Likelihood"},
  {theta: t, y: Math.exp(-0.5 * (t / pp_sd) ** 2), curve: "Prior"},
  {theta: t, y: pp.post[k] / pp_pmax, curve: "Posterior"}])
pp_mle = mleR(pp_r, wsB)
pp_map = mapR(pp_r, wsB, 0, pp_sd)
Plot.plot({
  width: 700, height: 300, x: {label: "Ability θ", domain: [-7, 7]}, y: {label: "Scaled to a peak of 1", domain: [0, 1.05]},
  color: {legend: true, domain: ["Prior", "Likelihood", "Posterior"], range: [palette.guide, palette.light, palette.main]},
  marks: [
    Plot.line(pp_data, {x: "theta", y: "y", stroke: "curve", strokeWidth: 2.5}),
    ...(Number.isFinite(pp_mle) ? [Plot.ruleX([pp_mle], {stroke: palette.light, strokeDasharray: "4,3"})] : []),
    Plot.ruleX([pp.eap], {stroke: palette.contrast, strokeWidth: 2})
  ]
})
```

```{ojs}
//| echo: false
md`Sum score ${pp_r} of 10. MLE: **${Number.isFinite(pp_mle) ? pp_mle.toFixed(2) : (pp_mle > 0 ? "+∞" : "−∞")}**${Number.isFinite(pp_mle) ? ` (SE ${seR(pp_mle, wsB).toFixed(2)}; dashed line)` : " (no peak)"}. MAP: **${pp_map.toFixed(2)}**. EAP: **${pp.eap.toFixed(2)}** (posterior SD ${pp.psd.toFixed(2)}; solid line).`
```

At the defaults, a score of 9 gives an MLE of 2.55, a MAP of 1.95 and an EAP of 2.08: the prior pulls the estimate about half a logit toward 0. At a score of 10 the likelihood never falls, yet the posterior has a peak (2.83) and a mean (3.05). With a prior SD of 1, the EAP for a score of 9 is 1.34; with 4, it is 2.66, a little beyond the MLE, because the posterior is skewed toward the open end. At a score of 6, the EAP stays within about 0.1 of the MLE (0.09) at any prior SD. That pull toward the prior mean is **shrinkage**, largest where the likelihood is flattest: at the ends of the scale, and on short tests.

#### Try it: shrinkage by score, and by test length

Estimates for every score on the Wordsum items. "Test length" repeats the ten items (40 items is four copies of each); the horizontal axis is the proportion correct.

```{ojs}
//| echo: false
viewof sh_k = Inputs.radio(new Map([["10 items", 1], ["20 items", 2], ["40 items", 4]]), {value: 1, label: "Test length"})
viewof sh_sd = Inputs.range([0.5, 4], {value: 1.88, step: 0.01, label: "Prior SD"})
```

```{ojs}
//| echo: false
sh_b = Array.from({length: sh_k}, () => wsB).flat()
sh_n = sh_b.length
sh_rows = Array.from({length: sh_n + 1}, (_, r) => ({r, prop: r / sh_n, mle: mleR(r, sh_b), wle: wleR(r, sh_b), eap: posteriorR(r, sh_b, 0, sh_sd).eap}))
sh_long = sh_rows.flatMap(d => [
  ...(Number.isFinite(d.mle) ? [{prop: d.prop, est: d.mle, which: "MLE"}] : []),
  {prop: d.prop, est: d.wle, which: "WLE"}, {prop: d.prop, est: d.eap, which: "EAP"}])
Plot.plot({
  width: 700, height: 320, x: {label: "Proportion correct", domain: [0, 1]}, y: {label: "Estimate of θ", domain: [-6, 6]},
  color: {legend: true, domain: ["MLE", "WLE", "EAP"], range: [palette.light, palette.guide, palette.main]},
  marks: [
    Plot.ruleY([0], {stroke: palette.rule}),
    Plot.line(sh_long, {x: "prop", y: "est", stroke: "which", strokeWidth: 2}),
    Plot.dot(sh_long, {x: "prop", y: "est", fill: "which", r: sh_k === 1 ? 4 : 2.5})
  ]
})
```

```{ojs}
//| echo: false
sh_90 = sh_rows[Math.round(0.9 * sh_n)]
md`At 90% correct (${sh_90.r} of ${sh_n}): MLE **${sh_90.mle.toFixed(2)}**, WLE ${sh_90.wle.toFixed(2)}, EAP **${sh_90.eap.toFixed(2)}**. At 100%: no MLE, WLE ${sh_rows[sh_n].wle.toFixed(2)}, EAP ${sh_rows[sh_n].eap.toFixed(2)}.`
```

The MLE at 90% correct is 2.55 whatever the length, because repeating every item multiplies both sides of the score equation by the same number. The EAP moves toward it as the test grows, from 2.08 with 10 items to 2.42 with 40, as the likelihood sharpens. The WLE sits nearer the MLE, and gives a perfect score a finite value (3.59 with 10 items).

Shrinkage carries into anything computed from estimated abilities, residuals included. If you've done [Model fit and out-of-sample prediction](fit-prediction.qmd), it is why outfit computed from `mirt`'s default EAP abilities averaged about 0.9 on simulated 20-item data where the model was true.

```{ojs}
//| echo: false
quiz({
  q: "Two respondents each answer 90% of their items correctly, one on a 10-item test and one on a 40-item version with four copies of each item. Both are scored by EAP with the same prior. Whose estimate is pulled further toward the prior mean?",
  options: [
    {text: "The respondent on the 10-item test", correct: true, why: "With fewer items the likelihood is flatter, so the prior carries more weight in the posterior. In the widget, the EAP at 90% correct is 2.08 with 10 items and 2.42 with 40, against an MLE of 2.55 for both."},
    {text: "The respondent on the 40-item test, since more items give the prior more to work with", correct: false, why: "The prior is the same in both cases; what changes is the likelihood. Forty items give a sharper likelihood, which outweighs the prior."},
    {text: "Neither: the same proportion correct gives the same estimate", correct: false, why: "That's true of the MLE here, whose score equation scales with the number of items. The EAP also depends on how sharp the likelihood is."}
  ]
})
```

### Other estimators, and plausible values

Four estimators come up in practice ([Kim & Nicewander, 1993](https://doi.org/10.1007/bf02294829), compare them on conventional tests):

- **MLE**: the $\theta$ that makes the responses most probable. It assumes nothing about the population. It is biased in short tests, which [Lord (1983)](https://doi.org/10.1007/BF02294018) derives along with a correction, and it doesn't exist for perfect patterns.
- **WLE**: [Warm's (1989)](https://doi.org/10.1007/bf02294627) weighted likelihood estimate adds a term to the score equation that removes the MLE's first-order bias, with the same large-sample variance. It uses no prior and is finite for every pattern.
- **MAP**: the posterior mode, found by Newton–Raphson with the prior adding a term to the score equation. Always finite.
- **EAP**: the posterior mean, computed on a grid. Always finite, and if the prior describes the population it has the smallest mean squared error averaged over that population, since squared error is what a posterior mean minimizes. Shrinkage is the other side of that coin.

No individual estimate carries the spread of $\theta$ itself: EAPs vary too little and MLEs too much, so variances, correlations and regressions built on them are off by an amount that depends on the test. Large-scale assessments such as NAEP handle this with **plausible values** ([Mislevy, 1991](https://doi.org/10.1007/bf02294457); [Mislevy, Beaton, Kaplan & Sheehan, 1992](https://doi.org/10.1111/j.1745-3984.1992.tb00371.x)): several random draws from each respondent's posterior, under a prior that is a regression of $\theta$ on background variables. Each set of draws has the spread and the group differences of $\theta$, the analysis is run on each set, and the results are combined as in multiple imputation. They are a tool for population quantities, not scores to report for individuals ([Wu, 2005](https://doi.org/10.1016/j.stueduc.2005.05.005)). [What does a score mean?](score-meaning.qmd) takes them up in full.

### Sufficiency, revisited

::: {.callout-tip appearance="simple"}
**Recall** from [The Rasch model](rasch.qmd): under the Rasch model the sum score $r$ is sufficient for $\theta$. The likelihood factorizes into $e^{r\theta}$, times a term free of $\theta$, over a term free of the pattern.
:::

Every estimator here uses the responses only through the likelihood, so under the Rasch model every one depends on them only through $r$. The MLE's and WLE's score equations contain the data only as $r$; in the EAP and MAP, the factor that depends on the pattern cancels from the posterior. Ten Wordsum items have eleven possible EAPs, one per sum score, which is how the widgets above could list every estimate.

::: {.callout-tip appearance="simple"}
**Recall** from [From the 1PL to the 4PL](1pl-to-4pl.qmd): under the 2PL the likelihood depends on the pattern only through the weighted score $\sum_i a_i x_i$, which is therefore sufficient; the score equation is $\sum_i a_i (x_i - P_i) = 0$.
:::

The same argument now runs through the weighted score: respondents with the same sum score but different items right get different estimates, ordered by $\sum_i a_i x_i$. The real-data section counts how many.

```{ojs}
//| echo: false
quiz({
  q: "Under a 2PL, two respondents each answer 5 of 10 items correctly, and one gets a higher EAP than the other. What can you say about their responses?",
  options: [
    {text: "The one with the higher EAP answered items with larger total slope correctly", correct: true, why: "Under the 2PL the posterior depends on the pattern only through $\\sum_i a_i x_i$, and the EAP increases with it. Same sum score, different weighted score."},
    {text: "The one with the higher EAP answered harder items correctly", correct: false, why: "Difficulty enters the likelihood only in a term without $\\theta$, which cancels in the posterior. What matters is the slopes of the items answered correctly."},
    {text: "Nothing: EAPs contain random error", correct: false, why: "An EAP is a fixed function of the pattern, the item parameters and the prior. Two patterns with the same weighted score get the same EAP."}
  ]
})
```

## Simulate

Which estimator lands closest to the truth? The code below simulates Rasch responses to ten items (difficulties from −2 to 2) from 5,000 respondents in two groups whose true means differ by 1 (overall mean 0, SD 1). It estimates every $\theta$ four ways, with the difficulties known and a normal(0, 1) prior, and reports bias and root mean squared error (RMSE) by true $\theta$, the SD of the estimates and the group difference. It is base R and runs in a second.

With the seed as written, 112 of the 5,000 respondents (2.2%) have no finite MLE, and `optimize()` on a perfect score returns 10, its upper bound. The MLE and WLE are close to unbiased across the middle; the EAP and MAP are pulled toward 0, by about 0.16 within one SD of the mean, 0.5 between one and two SDs out, and 1 beyond. In return, the EAP's RMSE in the middle is 0.52, against 0.78 to 0.80 for the MLE, and overall 0.61 against 0.79 (leaving out the 112 with no MLE). At the ends the order reverses. The SD of the MLEs is 1.26 and of the EAPs 0.80, against a true 1, and the group difference follows: 0.98 (MLE), 1.03 (WLE) and 0.65 (EAP). Change `length.out = 10` to 20 or 40: the EAP's group difference rises to 0.80 and then 0.89 as the likelihood takes over from the prior.

```{webr}
{{< include code/ability-estimation-sim.R >}}
```

[Download this code](code/ability-estimation-sim.R) to run it in your own R session.

## With real data

```{r}
#| include: false
knitr::read_chunk("code/ability-estimation-irw.R")
options(digits = 7)  # R's default, so the page prints what a fresh R session prints
```

Our main example is [c19prc_uk_mcbride_2021_wordsum](https://itemresponsewarehouse.org/tables/c19prc_uk_mcbride_2021_wordsum/): the Wordsum vocabulary test as answered by 2,058 UK adults in the sixth wave of the COVID-19 Psychological Research Consortium study ([Bentall et al., 2022](https://doi.org/10.5255/UKDA-SN-855552); the study's design is described by [McBride et al., 2021](https://doi.org/10.1002/mpr.1861)). Wordsum is the ten-item vocabulary test of the US General Social Survey, drawn from a test by [Thorndike (1942)](https://doi.org/10.1037/h0060053); [Cor, Haertel, Krosnick and Malhotra (2012)](https://doi.org/10.1016/j.ssresearch.2012.05.007) analyse it with item response theory. For each target word the respondent chooses the closest in meaning from five words, or "don't know". Its job here is to be a short test on which many respondents reach the floor or the ceiling. A response is 1 for the right word and 0 for anything else, including "don't know". [Download the code](code/ability-estimation-irw.R) to run every chunk yourself; it needs `mirt` ([Chalmers, 2012](https://doi.org/10.18637/jss.v048.i06)).

```{r fetch-ws}
#| echo: true
#| code-fold: show
```

There is one wave, every respondent answered all ten items, and the items run from 27% to 84% correct.

```{r fit-ws}
#| echo: true
#| code-fold: show
```

The difficulties fall in two groups, six easy items from −2.51 to −1.10 and four hard ones from 0.88 to 1.55, with nothing between. The SD of ability, 1.88, is the prior's SD for the EAP and MAP estimates below.

```{ojs}
//| echo: false
predict({
  q: "What share of the 2,058 Wordsum respondents have no finite maximum-likelihood estimate?",
  choices: ["Under 1%", "About 5%", "About 12%", "About 25%"],
  answer: "About 12%",
  reveal: "About 12%: the 95 respondents who answered no item correctly and the 156 who answered all ten, 251 in all (12.2%). With ten items and abilities spread with an SD of 1.88, many respondents sit beyond the easiest or the hardest item. Check the output below."
})
```

```{r nomle-ws}
#| echo: true
#| code-fold: show
```

The 251 respondents with a score of 0 or 10 are 12.2% of the sample, and `mirt` gives them $-\infty$ and $+\infty$. The ten-item simulation above, with items over a similar range but an ability SD of 1, had 2.2%: the spread of the respondents matters as much as where the items are.

```{r estimators-ws}
#| echo: true
#| code-fold: show
```

Each estimator gives one value per sum score (the largest spread within a score is 0 to four decimals). In the middle they nearly agree: a score of 5 gives −0.57 (MLE), −0.60 (WLE), −0.48 (MAP) and −0.47 (EAP). They part at the ends. A score of 9 gives 2.55, 2.26, 1.95 and 2.07, and a score of 10 gives $+\infty$, 3.59, 2.83 and 3.03. The MLE's standard error is 0.79 at a score of 4 and 1.13 at 9; the posterior SDs are smaller everywhere, because the prior adds information.

```{r se-ws}
#| echo: true
#| code-fold: show
```

`mirt`'s MLE standard errors are $1/\sqrt{I(\hat\theta)}$, the test information evaluated at each estimate (`all.equal` returns `TRUE`).

```{r shrink-ws}
#| echo: true
#| code-fold: show
```

The EAPs have an SD of 1.69, below the model's 1.88; the WLEs, at 1.96, are above it. The shortfall isn't lost: the variance of the EAPs (2.86) plus the average posterior variance (0.67) is 3.53, the model's variance (3.54) to rounding. What the responses can't locate stays in the posterior SD.

```{r twopl-ws}
#| echo: true
#| code-fold: show
```

BIC ([Schwarz, 1978](https://doi.org/10.1214/aos/1176344136)) prefers the 2PL to the Rasch model (19,398 against 19,887), and the slopes run from 1.09 to 5.20. The 282 respondents with a score of 5 gave 40 different response patterns, which have 40 different weighted scores and get 40 different EAPs, from −1.05 to −0.04, in exactly the order of their weighted scores (rank correlation 1). Under the Rasch model all 282 shared one EAP, −0.47: a full logit of difference among respondents with the same number right comes from the choice of model.

### Which estimate to report

The choice rests on what we're willing to assume about the respondent and on what the numbers will be used for.

- The MLE and WLE assume nothing about the population: the estimate comes from the respondent's own responses, which suits comparisons across populations. The price is noise: on Wordsum the MLE's standard errors run from 0.79 to 1.13, and 12% of respondents have no MLE.
- The EAP and MAP assume the respondent comes from the population the prior describes. If so, they are closer to the truth on average (in the simulation, an RMSE of 0.61 against 0.79). The price is shrinkage toward that population's mean, largest at the ends and on short tests, which carries into spreads and group differences.

When I report a score to an individual from a test like this one, I report the EAP with its posterior SD, and I say that it is pulled toward the mean of the population the prior describes. It always exists, it is the most accurate on average for the people the test was given to, and its standard error grows where the test runs out of items. What I don't do is average EAPs to compare groups: the simulation's group difference of 1 came back as 0.65. The MLEs and WLEs came closer there, but they vary too much for spreads and correlations, and the MLE loses everyone at the floor and the ceiling. For group questions I'd put the groups into the model, or use plausible values. And if the Rasch model fits and everyone took the same items, [The Rasch model](rasch.qmd) has already argued for the sum score: every estimator in the table above is a relabelling of it.

## Problems

1. **Three items by hand.** For three Rasch items with $b = -1, 0, 1.5$, write down the likelihood of the pattern 1-1-0 and plot it. At what $\theta$ does 1-1-0 become more likely than 1-0-0, and why is the answer one of the difficulties? (PS4#1)
2. **The Rasch score equation.** Show that under the Rasch model the MLE solves $\sum_i P_i(\theta) = r$, so it depends on the pattern only through the sum score. Then show that the EAP does too, by writing the posterior with the factorization from the Rasch lesson's Go deeper callout.
3. **Change the prior.** Rescore Wordsum by EAP with a prior SD of 1 instead of 1.88 (`fscores(m, method = "EAP", mean = 0, cov = 1)`). Which respondents move most? What happens to the SD of the EAPs, and why?
4. **Classroom averages.** A school reports each student's EAP from a 15-item test and compares classroom means, using one normal prior for all students. What goes wrong with the comparison, and in which direction? What would you report instead, and to whom?
5. **Does EAP reduce error?** Extend the simulation: 1,000 respondents at each true $\theta$ from −3 to 3 in steps of 0.5. Where does the EAP's RMSE beat the MLE's, and where not? Then score normal(0, 1) respondents with a prior of mean 1. What changes? (PS5#2)
6. **Challenge (open): the floor and the ceiling.** Twelve percent of Wordsum respondents score 0 or 10. A survey analyst wants vocabulary as a predictor in a regression of, say, trust in science. What should they do with those 251 respondents: drop them, use WLEs, EAPs or plausible values, or fit the regression inside the item response model? Does it depend on the question? I don't think there is a single answer; make the case for yours.

```{r}
#| output: asis
ask_claude("ability-estimation", "It covers estimating one respondent's ability with the item parameters known: maximum likelihood by the score equation and its standard error from information, why all-correct and all-wrong patterns have no MLE, EAP estimation with a prior on a grid of nodes and the shrinkage it brings, MAP and Warm's weighted likelihood estimator, plausible values in brief, and why under the Rasch model every estimator is a function of the sum score while under the 2PL it follows the weighted score, with mirt analyses of the Wordsum vocabulary test.")
```

## Going further

- Bock, R. D., & Mislevy, R. J. (1982). Adaptive EAP estimation of ability in a microcomputer environment. *Applied Psychological Measurement*, 6(4), 431–444. <https://doi.org/10.1177/014662168200600405>. The EAP estimate on a grid of nodes, and its posterior SD as a standard error.
- Warm, T. A. (1989). Weighted likelihood estimation of ability in item response theory. *Psychometrika*, 54(3), 427–450. <https://doi.org/10.1007/bf02294627>. The WLE, compared with the MLE and Bayes modal estimates.
- Kim, J. K., & Nicewander, W. A. (1993). Ability estimation for conventional tests. *Psychometrika*, 58(4), 587–599. <https://doi.org/10.1007/bf02294829>. Five estimators compared for bias, standard error and reliability.
- Lord, F. M. (1983). Unbiased estimators of ability parameters, of their variance, and of their parallel-forms reliability. *Psychometrika*, 48(2), 233–245. <https://doi.org/10.1007/BF02294018>. The bias of the MLE, and estimators without it.
- Mislevy, R. J. (1986). Bayes modal estimation in item response models. *Psychometrika*, 51(2), 177–195. <https://doi.org/10.1007/bf02293979>. Priors in item response models.
- Mislevy, R. J. (1991). Randomization-based inference about latent variables from complex samples. *Psychometrika*, 56(2), 177–196. <https://doi.org/10.1007/bf02294457>. Plausible values as multiple imputation, with NAEP data.
- Mislevy, R. J., Beaton, A. E., Kaplan, B., & Sheehan, K. M. (1992). Estimating population characteristics from sparse matrix samples of item responses. *Journal of Educational Measurement*, 29(2), 133–161. <https://doi.org/10.1111/j.1745-3984.1992.tb00371.x>. Plausible values, from the concepts to NAEP.
- Wu, M. (2005). The role of plausible values in large-scale surveys. *Studies in Educational Evaluation*, 31(2–3), 114–128. <https://doi.org/10.1016/j.stueduc.2005.05.005>. Plausible values and individual estimates, side by side.
- Baker, F. B., & Kim, S.-H. (2004). *Item response theory: Parameter estimation techniques* (2nd ed.). Marcel Dekker. <https://doi.org/10.1201/9781482276725>. Newton–Raphson for abilities, and the rest of the machinery, step by step.
- Lord, F. M. (1980). *Applications of item response theory to practical testing problems*. Erlbaum. <https://doi.org/10.4324/9780203056615>. The likelihood for ability with the items known.
- Albert, A., & Anderson, J. A. (1984). On the existence of maximum likelihood estimates in logistic regression models. *Biometrika*, 71(1), 1–10. <https://doi.org/10.1093/biomet/71.1.1>. Separation, of which perfect patterns are an instance.
- Andersen, E. B. (1977). Sufficient statistics and latent trait models. *Psychometrika*, 42(1), 69–81. <https://doi.org/10.1007/BF02293746>. Why only the Rasch model makes the sum score sufficient.
- Cor, M. K., Haertel, E., Krosnick, J. A., & Malhotra, N. (2012). Improving ability measurement in surveys by following the principles of IRT: The Wordsum vocabulary test in the General Social Survey. *Social Science Research*, 41(5), 1003–1016. <https://doi.org/10.1016/j.ssresearch.2012.05.007>. Wordsum analysed with item response theory.
- Thorndike, R. L. (1942). Two screening tests of verbal intelligence. *Journal of Applied Psychology*, 26(2), 128–135. <https://doi.org/10.1037/h0060053>. Where the Wordsum items come from.
- McBride, O., Murphy, J., Shevlin, M., Gibson-Miller, J., Hartman, T. K., Hyland, P., et al. (2021). Monitoring the psychological, social, and economic impact of the COVID-19 pandemic in the population: Context, design and conduct of the longitudinal COVID-19 Psychological Research Consortium (C19PRC) Study. *International Journal of Methods in Psychiatric Research*, 30(1), e1861. <https://doi.org/10.1002/mpr.1861>. The study that gave Wordsum to these respondents.
- Schwarz, G. (1978). Estimating the dimension of a model. *The Annals of Statistics*, 6(2), 461–464. <https://doi.org/10.1214/aos/1176344136>. The BIC.
- Chalmers, R. P. (2012). mirt: A multidimensional item response theory package for the R environment. *Journal of Statistical Software*, 48(6). <https://doi.org/10.18637/jss.v048.i06>. `fscores()` computes every estimator in this lesson.

```{r}
#| output: asis
data_sources("ability-estimation")
```

## For instructors

- **Source:** EDUC 252 class 4 (slides c4, 44–48) and class 5 (slides c5, 64–66), problem sets 3 (#2), 4 (#1) and 5 (#2), and the code in `ben-domingue/252`: `ps3/1_abilities.R`, `ps5/eap_versus_mle.R`.
- **Timing:** the core ideas and widgets take about 60 minutes; the Wordsum analysis works as a 25-minute lab. Every fit takes a few seconds in local R.
- **Changed from 252:** the 252 code finds the MLE with `optim()` on a bounded interval, which returns the bound for a perfect pattern; the simulation here solves the score equation, counts the perfect patterns, and adds the WLE, the MAP and a group difference to PS5#2's comparison of MLE and EAP. The real data (Wordsum) are new.
- **Held back:** where the item parameters come from, and the EAP's nodes and weights as the E step of EM ([Estimating item parameters](item-estimation.qmd)); EAP in adaptive testing ([Item banks and adaptive testing](item-banks-cat.qmd)); priors on item parameters ([Guessing and priors](guessing-priors.qmd)); plausible values in full, and which estimate to put on a reported score scale ([What does a score mean?](score-meaning.qmd)). Out-of-sample prediction with MLE and EAP abilities (PS5#2b) is left as an extension of problem 5.
- **Solutions:** held back in `solutions/ability-estimation.qmd` (not published).
