---
title: "Dimensionality and multidimensional IRT"
---

```{r}
#| output: asis
source("_course.R")
lesson_header("dimensionality")
```

## What this is for

*To be written.*

## Goals

## Core ideas

## Worked example

## Problems

## Going further

- Reckase, M. D. (2016). Logistic multidimensional models. In W. J. van der Linden (Ed.), *Handbook of item response theory: Vol. 1. Models* (pp. 189–209). Chapman and Hall/CRC. <https://doi.org/10.1201/9781315374512>. A handbook chapter on multidimensional logistic item response models.
- Cai, L. (2016). Two-tier item factor analysis modeling. In W. J. van der Linden (Ed.), *Handbook of item response theory: Vol. 1. Models* (pp. 421–435). Chapman and Hall/CRC. <https://doi.org/10.1201/9781315374512>. A handbook chapter on two-tier item factor models, which include the bifactor model.
- Gibbons, R. D., & Cai, L. (2017). Dimensionality analysis. In W. J. van der Linden (Ed.), *Handbook of item response theory: Vol. 3. Applications* (pp. 47–66). Chapman and Hall/CRC. <https://doi.org/10.1201/9781315117430>. A handbook chapter on assessing the dimensionality of item response data.

```{r}
#| output: asis
data_sources("dimensionality")
```

## For instructors

