---
title: "What is an item? Explanatory item response models"
---

```{r}
#| output: asis
source("_course.R")
lesson_header("explanatory-irt")
```

## What this is for

*To be written.*

## Goals

## Core ideas

## Worked example

## Problems

## Going further

- Janssen, R. (2016). Linear logistic models. In W. J. van der Linden (Ed.), *Handbook of item response theory: Vol. 1. Models* (pp. 211–224). Chapman and Hall/CRC. <https://doi.org/10.1201/9781315374512>. A handbook chapter on the linear logistic test model and related models.
- De Boeck, P., & Wilson, M. R. (2016). Explanatory response models. In W. J. van der Linden (Ed.), *Handbook of item response theory: Vol. 1. Models* (pp. 565–580). Chapman and Hall/CRC. <https://doi.org/10.1201/9781315374512>. A handbook chapter on explanatory item response models.
- Rabe-Hesketh, S., & Skrondal, A. (2016). Generalized linear latent and mixed modeling. In W. J. van der Linden (Ed.), *Handbook of item response theory: Vol. 1. Models* (pp. 503–526). Chapman and Hall/CRC. <https://doi.org/10.1201/9781315374512>. A handbook chapter on item response models as generalized linear latent and mixed models.

```{r}
#| output: asis
data_sources("explanatory-irt")
```

## For instructors

