---
title: "Factor analysis II: confirmatory models, omega, and IRT as factor analysis"
---

```{r}
#| output: asis
source("_course.R")
lesson_header("fa-confirmatory")
```

```{webr}
#| context: setup
{{< include code/_webr-setup.R >}}
suppressWarnings(webr::install("lavaan", quiet = TRUE))
# webR can't count CPU cores (detectCores() is NA), and lavaan's option checks then
# never finish; report one core. Not needed in local R.
local({
  ns <- asNamespace("parallel")
  unlockBinding("detectCores", ns)
  assign("detectCores", function(...) 1L, envir = ns)
  lockBinding("detectCores", ns)
})
```

```{ojs}
//| echo: false
import {grid, palette} from "./widgets/irt.js"
import {zeros, implied, fitCFA, fitIndices, omegas, pOgive, ogiveToIRT, p2plAB} from "./widgets/fa-confirmatory.js"
import {quiz, predict} from "./widgets/quiz.js"
```

## What this is for

The Grit-O scale was written around two facets of grit, perseverance of effort and consistency of interest. Does it measure one thing or two? [Factor analysis I](fa-exploratory.qmd) let the data propose a structure. This lesson states a structure first and asks whether the data agree: **confirmatory factor analysis** (CFA). A written-down structure also answers four more questions. How well does it fit? How reliable is a score built on it (**omega**)? What is it, seen from item response theory? And does it hold in two groups, so that their means can be compared (**measurement invariance**)? Along the way we settle whether the DASS-21 yields one score or three.

## Goals

By the end of this lesson you should be able to:

- write a confirmatory factor model in `lavaan` syntax, say how it is identified, and count its degrees of freedom;
- read χ², CFI, TLI, RMSEA and SRMR, say what each responds to, and compare nested models;
- compute omega and omega hierarchical, and say when omega equals alpha;
- convert an ordinal factor loading and threshold into a slope and a difficulty, and check the conversion against `mirt`;
- test configural, metric and scalar invariance across two groups, and say which one a comparison of means needs.

## Core ideas

### Stating the structure first

A confirmatory model is the common factor model with most loadings fixed at zero before the data are seen. For grit, perseverance items load on one factor and consistency items on the other, with each item's other loading fixed at 0. [Jöreskog (1969)](https://doi.org/10.1007/bf02289343) showed how to fit such a model by maximum likelihood and test it. We use the `lavaan` package ([Rosseel, 2012](https://doi.org/10.18637/jss.v048.i02)), whose syntax reads almost like a path diagram. `=~` means "is measured by", and an item left off a line has its loading on that factor fixed at zero:

```r
two <- "per =~ g1 + g4 + g6 + g9 + g10 + g12
        con =~ g2 + g3 + g5 + g7 + g8 + g11"
fit <- cfa(two, data = grit)
```

The factors may correlate (lavaan's default), and each item keeps a unique variance.

::: {.callout-tip appearance="simple"}
**Recall** from [Factor analysis I](fa-exploratory.qmd): with two or more factors, any rotation reproduces the correlations exactly as well as any other. The data fix the correlations, not the frame of the factors.
:::

The zeros pin the frame, since rotating would move them. One indeterminacy is left: a factor has no natural unit. The model is identified once each factor's scale is set, by fixing one loading per factor to 1 (lavaan's default) or each factor's variance to 1 (`std.lv = TRUE`). The two choices fit identically and report loadings in different units, as the grit output shows.

What can the data test? They supply $p(p+1)/2$ variances and covariances. The model spends one parameter on each free loading, unique variance and factor correlation. The **degrees of freedom** are the difference, and each one is a constraint the data could violate.

```{ojs}
//| echo: false
quiz({
  q: "Twelve items, two factors, each item loading on one factor only, factor variances fixed at 1 and the factors correlated. How many degrees of freedom does the model have?",
  options: [
    {text: "53", correct: true, why: "There are $12 \\times 13/2 = 78$ variances and covariances. The model estimates 12 loadings, 12 unique variances and 1 factor correlation, 25 in all: $78 - 25 = 53$."},
    {text: "54", correct: false, why: "That is the one-factor model, with 24 parameters. The factor correlation costs one more."},
    {text: "66", correct: false, why: "That counts only the 66 covariances and forgets the 12 variances the data also supply."}
  ]
})
```

### Fit indices, and what they can't tell you

Every fit index compares the covariance matrix the model implies with the one the data supply. The likelihood-ratio **χ²** tests exact fit: it is $N - 1$ times the discrepancy, so a large sample rejects any model that is slightly wrong, and every model is. The other indices ask how wrong. **RMSEA** is the misfit per degree of freedom ([Steiger, 1990](https://doi.org/10.1207/s15327906mbr2502_4)). **CFI** ([Bentler, 1990](https://doi.org/10.1037/0033-2909.107.2.238)) and **TLI** ([Tucker & Lewis, 1973](https://doi.org/10.1007/BF02291170)) compare the model's misfit with that of a baseline in which the items are uncorrelated. **SRMR** is the root mean square residual correlation.

[Hu and Bentler (1999)](https://doi.org/10.1080/10705519909540118) found, in simulations of maximum-likelihood estimation, that cutoffs near 0.95 for CFI and TLI, 0.06 for RMSEA and 0.08 for SRMR were needed to call a fit good. [Marsh, Hau and Wen (2004)](https://doi.org/10.1207/s15328007sem1103_2) argue against reading them as fixed rules, and with ordinal estimators the same indices tell a different story ([Xia & Yang, 2019](https://doi.org/10.3758/s13428-018-1055-2)). They describe how far off a model is; the residuals show where.

#### Try it: what misfit does to fit indices

Six items, two factors correlated 0.3. The population also has things the model leaves out: item 3 loads on the second factor, and items 5 and 6 share an error. The widget fits the two-factor model by maximum likelihood.

```{ojs}
//| echo: false
viewof fs_cross = Inputs.range([0, 0.5], {value: 0.3, step: 0.05, label: "Item 3's loading on factor 2 (not in the model)"})
viewof fs_cerr = Inputs.range([0, 0.3], {value: 0, step: 0.05, label: "Error covariance, items 5 and 6 (not in the model)"})
viewof fs_n = Inputs.range([100, 5000], {value: 500, step: 100, label: "Sample size N"})
```

```{ojs}
//| echo: false
fs = {
  const lam = [0.7, 0.6, 0.5, 0.7, 0.6, 0.5];
  const L = lam.map((v, i) => i < 3 ? [v, 0] : [0, v]); L[2][1] = fs_cross;
  const P = [[1, 0.3], [0.3, 1]];
  const common = implied(L, P, zeros(6));
  const Th = zeros(6).map((r, i) => r.map((_, j) => i === j ? 1 - common[i][i] : 0));
  Th[4][5] = Th[5][4] = fs_cerr;
  const S = implied(L, P, Th);
  const fit = fitCFA(S, [[1, 0], [1, 0], [1, 0], [0, 1], [0, 1], [0, 1]]);
  return {S, fit, idx: fitIndices(S, fit, 13, fs_n)};
}
fs_res = fs.S.flatMap((r, i) => r.map((v, j) => ({row: `Item ${i + 1}`, col: `Item ${j + 1}`, res: v - fs.fit.Sigma[i][j]})))
Plot.plot({
  width: 400, height: 380, padding: 0.04, x: {label: null, axis: "top"}, y: {label: null},
  color: {type: "diverging", domain: [-0.15, 0.15], range: [palette.contrast, palette.white, palette.main], legend: true, label: "Residual correlation (data − model)"},
  marks: [Plot.cell(fs_res, {x: "col", y: "row", fill: "res"}),
          Plot.text(fs_res, {x: "col", y: "row", text: d => d.res.toFixed(2), fontSize: 12, fill: palette.ink})]
})
```

```{ojs}
//| echo: false
md`Expected χ² **${fs.idx.chisq.toFixed(1)}** on ${fs.idx.df} df (the 5% critical value is 15.5) · CFI **${fs.idx.cfi.toFixed(3)}** · RMSEA **${fs.idx.rmsea.toFixed(3)}** · SRMR **${fs.idx.srmr.toFixed(3)}** · estimated factor correlation **${fs.fit.Phi[0][1].toFixed(2)}** (true 0.3)`
```

At the start, the omitted cross-loading gives an RMSEA of 0.080 and a CFI of 0.944, and the factor correlation comes out at 0.48: the model routes item 3's link to factor 2 through the correlation. The residuals point at item 3. Move N from 500 to 2,000: χ² rises from 34 to 111, and the other indices stay put. Now set the cross-loading to 0 and the error covariance to 0.2. CFI is 0.994 and RMSEA 0.026, inside the cutoffs, and the residual for items 5 and 6 is almost zero: the model raised their loadings to absorb the shared error, and the strain shows up between item 4 and items 1–3. A residual marks where the model strains, which isn't always where the cause is.

Nested models can be compared directly: the one-factor model is the two-factor model with the factor correlation fixed at 1, so the difference in χ² tests that constraint. Robust estimators need a scaled difference ([Satorra & Bentler, 2001](https://doi.org/10.1007/BF02296192)), which `lavTestLRT()` computes. Prediction offers another view: [Zhang et al. (2026)](https://doi.org/10.1080/00273171.2026.2645212) carry the IMV, which [Model fit and out-of-sample prediction](fit-prediction.qmd) teaches, over to CFA with binary items.

```{ojs}
//| echo: false
quiz({
  q: "A CFA fitted to 20,000 respondents has χ² = 450 on 100 degrees of freedom ($p < .001$) and CFI = 0.98. What should you make of the χ²?",
  options: [
    {text: "The model should be rejected", correct: false, why: "It is rejected as *exactly* true, but no model is exactly true, and with 20,000 respondents χ² detects tiny misfit. The misfit per respondent is small: RMSEA $= \\sqrt{(450 - 100)/(100 \\times 19{,}999)} \\approx 0.013$."},
    {text: "It says the model isn't exactly right, which with this N was always going to happen; how wrong, and where, is the question", correct: true, why: "χ² grows with $N$ for a fixed misfit (the widget shows it). RMSEA here is about 0.013 and CFI 0.98; the residuals show where the misfit is."},
    {text: "χ² is invalid with a sample this large", correct: false, why: "It is valid; it answers a question (exact fit) that large samples always answer 'no'."}
  ]
})
```

### Omega: reliability from the model

::: {.callout-tip appearance="simple"}
**Recall** from [Classical test theory and reliability](ctt-reliability.qmd): alpha is a lower bound on reliability, with equality under essential tau-equivalence. [Factor analysis I](fa-exploratory.qmd) read tau-equivalence as a one-factor model with equal loadings.
:::

A factor model makes the bound unnecessary. Under a one-factor model with loadings $\lambda_i$ and unique variances $\psi_i$, the sum score has variance $(\sum_i \lambda_i)^2 + \sum_i \psi_i$, of which $(\sum_i\lambda_i)^2$ comes from the factor. Their ratio is **omega** ([McDonald, 1999](https://doi.org/10.4324/9781410601087)):
$$\omega = \frac{(\sum_i \lambda_i)^2}{(\sum_i \lambda_i)^2 + \sum_i \psi_i}.$$
When the loadings are equal, omega equals alpha (problem 1). When they aren't, alpha is smaller, usually by a little ([Revelle & Zinbarg, 2009](https://doi.org/10.1007/s11336-008-9102-z)).

Omega raises a sharper question when a scale has subscales. In a **bifactor** model ([Holzinger & Swineford, 1937](https://doi.org/10.1007/BF02287965); [Reise, 2012](https://doi.org/10.1080/00273171.2012.715555)), every item loads on a general factor and on one specific factor for its subscale, with all factors uncorrelated. **Omega total** counts the general and specific factors as signal. **Omega hierarchical**, $\omega_h$, counts only the general factor: it is the share of the total score's variance that one common construct accounts for ([Rodriguez, Reise & Haviland, 2016](https://doi.org/10.1037/met0000045)).

#### Try it: alpha, omega and omega hierarchical

Nine items in three subscales of three.

```{ojs}
//| echo: false
viewof om_g = Inputs.range([0.3, 0.6], {value: 0.6, step: 0.05, label: "Average general-factor loading"})
viewof om_sp = Inputs.range([0, 0.3], {value: 0, step: 0.05, label: "Spread of general loadings (± within each subscale)"})
viewof om_s = Inputs.range([0, 0.4], {value: 0, step: 0.05, label: "Specific-factor loading"})
```

```{ojs}
//| echo: false
om = {
  const off = [-1, 0, 1, -1, 0, 1, -1, 0, 1];
  return omegas(off.map(o => om_g + om_sp * o), off.map(() => om_s), [1, 1, 1, 2, 2, 2, 3, 3, 3]);
}
om_rows = [{k: "Alpha", v: om.alpha}, {k: "Omega total", v: om.omegaT}, {k: "Omega hierarchical", v: om.omegaH}]
Plot.plot({
  width: 560, height: 170, marginLeft: 150, x: {domain: [0, 1], label: "Coefficient"}, y: {label: null, domain: om_rows.map(d => d.k)},
  marks: [Plot.barX(om_rows, {y: "k", x: "v", fill: d => d.k === "Alpha" ? palette.contrast : palette.main}),
          Plot.text(om_rows, {y: "k", x: "v", text: d => d.v.toFixed(3), dx: 22})]
})
```

With equal loadings and no specific factors, all three are 0.835. Spread the loadings to ±0.2 and alpha falls a little below omega (0.833 against 0.841): the lower bound at work. Set the spread to 0 and the specific loading to 0.4: omega total rises to 0.886 and omega hierarchical falls to 0.771, because the subscales now carry reliable variance of their own. Alpha sits between the two.

```{ojs}
//| echo: false
quiz({
  q: "A scale's bifactor model gives $\\omega_{\\text{total}} = 0.90$ and $\\omega_h = 0.80$. How is the variance of its total score divided?",
  options: [
    {text: "80% general factor, 10% specific factors, 10% unique", correct: true, why: "$\\omega_h$ is the general factor's share; $\\omega_{\\text{total}} - \\omega_h = 0.10$ is the specific factors' share; the remaining $1 - 0.90$ is unique variance."},
    {text: "90% general factor, 10% unique", correct: false, why: "That would make $\\omega_h = \\omega_{\\text{total}}$. The 0.10 gap between them belongs to the specific factors."},
    {text: "80% general factor, 20% unique", correct: false, why: "The specific factors are reliable variance too; omega total counts them, which is why it is 0.90."}
  ]
})
```

### Item response theory as factor analysis

::: {.callout-tip appearance="simple"}
**Recall** from [From the 1PL to the 4PL](1pl-to-4pl.qmd): the 2PL gives each item a slope $a$ and a difficulty $b$; `mirt` writes $a\theta + d$, so $b = -d/a$.
:::

[Factor analysis I](fa-exploratory.qmd) treated an ordered response as a coarsened continuous one and factored polychoric correlations. Write that down for one item: a latent response $y^* = \lambda\eta + \varepsilon$, with $\text{var}(y^*) = 1$, and $x = 1$ when $y^*$ passes a threshold $\tau$. Then
$$\Pr(x = 1 \mid \eta) = \Phi\!\left(\frac{\lambda\eta - \tau}{\sqrt{1 - \lambda^2}}\right) = \Phi\big(a(\eta - b)\big), \quad a = \frac{\lambda}{\sqrt{1-\lambda^2}},\; b = \frac{\tau}{\lambda}.$$
That is the normal-ogive 2PL. The factor model for dichotomous items *is* an item response model ([Takane & de Leeuw, 1987](https://doi.org/10.1007/bf02294363)), and with more categories each threshold becomes a category boundary of the graded response model ([Samejima, 1969](https://doi.org/10.1007/BF03372160); [Kamata & Bauer, 2008](https://doi.org/10.1080/10705510701758406)). For the logistic metric of `mirt` ([Chalmers, 2012](https://doi.org/10.18637/jss.v048.i06)) and this course, multiply the slope by about 1.7 ([Camilli, 1994](https://doi.org/10.3102/10769986019003293), on where the constant comes from); this is the only place in the course it appears.

#### Try it: loading to slope

```{ojs}
//| echo: false
viewof ls_l = Inputs.range([0.1, 0.95], {value: 0.7, step: 0.05, label: "Loading λ"})
viewof ls_t = Inputs.range([-2, 2], {value: 0.35, step: 0.05, label: "Threshold τ"})
```

```{ojs}
//| echo: false
ls = {
  const irt = ogiveToIRT(ls_l, ls_t);
  const g = grid(-4, 4, 161);
  const maxDiff = Math.max(...g.map(e => Math.abs(pOgive(e, ls_l, ls_t) - p2plAB(e, irt.a, irt.b))));
  return {irt, maxDiff, rows: g.flatMap(e => [
    {eta: e, p: pOgive(e, ls_l, ls_t), curve: "Factor model (normal ogive)"},
    {eta: e, p: p2plAB(e, irt.a, irt.b), curve: "Logistic 2PL, a = 1.702 × normal slope"}])};
}
Plot.plot({
  width: 700, height: 300, x: {label: "Factor η (= θ)"}, y: {label: "Pr(x = 1)", domain: [0, 1]},
  color: {legend: true, domain: ["Factor model (normal ogive)", "Logistic 2PL, a = 1.702 × normal slope"], range: [palette.main, palette.contrast]},
  marks: [Plot.ruleX([ls.irt.b], {stroke: palette.guide, strokeDasharray: "4,3"}),
          Plot.line(ls.rows, {x: "eta", y: "p", stroke: "curve", strokeWidth: 2.5})]
})
```

```{ojs}
//| echo: false
md`Normal-ogive slope **${ls.irt.aNormal.toFixed(2)}**, logistic slope **${ls.irt.a.toFixed(2)}**, difficulty b = τ/λ = **${ls.irt.b.toFixed(2)}** (dashed line). Largest gap between the two curves: **${ls.maxDiff.toFixed(3)}**.`
```

At a loading of 0.7 and a threshold of 0.35, the slope is 0.98 (normal ogive) or 1.67 (logistic), the difficulty is 0.5, and the curves never differ by more than 0.01, wherever the sliders sit. Push the loading to 0.9 and the slope passes 3.5; at 0.95 it passes 5. Loadings stop at 1 and slopes don't.

::: {.callout-note collapse="true"}
## Go deeper: ordinal factor analysis is the graded response model

**Assumptions:** an item with ordered categories $0, \dots, K$; a latent response $y^* = \lambda^\top \eta + \varepsilon$ with $\eta$ multivariate normal (variances 1), $\varepsilon \sim N(0, 1 - h^2)$ independent of $\eta$ and of the other items' errors, where $h^2 = \text{var}(\lambda^\top\eta)$ is the communality, so $\text{var}(y^*) = 1$. The response is $x = k$ when $\tau_k < y^* \le \tau_{k+1}$, with $\tau_0 = -\infty$ and $\tau_{K+1} = \infty$.

**Boundary curves.** For $k = 1, \dots, K$,
$$\Pr(x \ge k \mid \eta) = \Pr(\varepsilon > \tau_k - \lambda^\top\eta) = \Phi\!\left(\frac{\lambda^\top\eta - \tau_k}{\sqrt{1 - h^2}}\right).$$
With one factor this is $\Phi\big(a(\eta - b_k)\big)$ with $a = \lambda/\sqrt{1-\lambda^2}$ and $b_k = \tau_k/\lambda$: one slope per item and one boundary per category, which is Samejima's (1969) normal-ogive graded response model. Category probabilities are differences of adjacent boundary curves. With several factors, the slope on factor $m$ is $a_m = \lambda_m/\sqrt{1 - h^2}$.

**Local independence.** The errors are independent across items, so given $\eta$ the responses are independent. Item response theory's local independence is the factor model's uncorrelated unique parts.

**Logistic metric.** $\sup_z |\Phi(z) - \text{logit}^{-1}(1.702z)| < 0.01$ (the widget's gap), so a logistic slope is about $1.702a$; in `mirt`'s form, $d_k = -1.702\,a\,b_k$.

**Why the numbers differ in practice.** The model is the same; the estimators are not. `lavaan`'s WLSMV estimates thresholds and polychoric correlations, which use the data one or two items at a time ([Muthén, 1984](https://doi.org/10.1007/BF02294210); [Flora & Curran, 2004](https://doi.org/10.1037/1082-989X.9.4.466)). `mirt` maximizes the likelihood of whole response patterns. When the model holds they converge on the same values as the sample grows; in a finite sample they differ by sampling error, by the logistic approximation, and wherever the model misfits, since the two estimators weigh the misfit differently.
:::

When can Likert items be treated as numbers? In simulations by [Rhemtulla, Brosseau-Liard and Savalei (2012)](https://doi.org/10.1037/a0029315), robust maximum likelihood did about as well as the categorical estimator with five or more roughly symmetric categories, and the categorical estimator did better with fewer. Grit has five categories with no pile-up at either end; the DASS-21 has four, most responses at zero. We treat them accordingly.

### Does the model hold in both groups?

Do women and men with the same level of anxiety answer an anxiety item the same way? If not, a difference in their scores is partly a difference in the items. **Measurement invariance** ([Meredith, 1993](https://doi.org/10.1007/BF02294825)) asks this of a factor model fitted to two or more groups at once, in steps that add constraints ([Vandenberg & Lance, 2000](https://doi.org/10.1177/109442810031002); [Millsap, 2011](https://doi.org/10.4324/9780203821961); [Putnick & Bornstein, 2016](https://doi.org/10.1016/j.dr.2016.06.004)):

- **configural**: the same items load on the same factors in every group;
- **metric**: the loadings are equal, so a unit of the factor means the same thing in each group;
- **scalar**: the intercepts (for ordinal items, the thresholds) are equal too, so the same factor level predicts the same response.

Comparing factor correlations across groups needs metric invariance; comparing means needs scalar. When one item fails, **partial invariance** frees its parameters and keeps the rest equal. In item response language, an unequal loading is non-uniform differential item functioning (DIF) and an unequal intercept is uniform DIF, which [Differential item functioning](dif.qmd) takes up item by item. With ordinal items, which parameters are identified at each step is subtler than with continuous ones ([Wu & Estabrook, 2016](https://doi.org/10.1007/s11336-016-9506-0)).

#### Try it: one item that works differently

Four items measure one factor in groups A and B (loadings 0.7, intercepts 0, factor SD 1). Item 4 can behave differently in group B.

```{ojs}
//| echo: false
viewof inv_k = Inputs.range([-1, 1], {value: 0.5, step: 0.05, label: "True factor mean, B minus A (SDs)"})
viewof inv_nu = Inputs.range([-0.5, 0.5], {value: 0, step: 0.05, label: "Item 4's intercept, B minus A"})
viewof inv_l = Inputs.range([-0.4, 0.4], {value: 0, step: 0.05, label: "Item 4's loading, B minus A"})
```

```{ojs}
//| echo: false
inv = {
  const sdA = Math.sqrt((4 * 0.7) ** 2 + 4 * (1 - 0.49));
  return {gap: (inv_nu + (3 * 0.7 + 0.7 + inv_l) * inv_k) / sdA, fair: (4 * 0.7 * inv_k) / sdA,
          level: Math.abs(inv_l) > 1e-9 ? "metric invariance fails (non-uniform DIF), and so does scalar" :
                 Math.abs(inv_nu) > 1e-9 ? "metric invariance holds, but scalar fails (uniform DIF)" : "scalar invariance holds"};
}
Plot.plot({
  width: 700, height: 290, x: {label: "Factor η", domain: [-3, 3]}, y: {label: "Expected response to item 4"},
  color: {legend: true, domain: ["Group A", "Group B"], range: [palette.main, palette.contrast]},
  marks: [Plot.ruleX([0], {stroke: palette.main, strokeDasharray: "4,3"}), Plot.ruleX([inv_k], {stroke: palette.contrast, strokeDasharray: "4,3"}),
          Plot.line(grid(-3, 3, 61).flatMap(e => [{e, y: 0.7 * e, g: "Group A"}, {e, y: inv_nu + (0.7 + inv_l) * e, g: "Group B"}]),
            {x: "e", y: "y", stroke: "g", strokeWidth: 2.5})]
})
```

```{ojs}
//| echo: false
md`Here ${inv.level}. Sum-score gap, B minus A: **${inv.gap.toFixed(2)}** SDs of group A's sum score; if item 4 were invariant it would be **${inv.fair.toFixed(2)}**. (Dashed lines: each group's factor mean.)`
```

With both item-4 controls at 0 the lines coincide, and the sum-score gap (0.45 SD) is the latent 0.5 SD shrunk by unreliability. Set the true gap to 0 and item 4's intercept to 0.3: the sum scores differ by 0.10 SD between groups that don't differ at all. A loading difference makes the lines cross, so the item favours one group at low levels and the other at high levels.

## Simulate

The code simulates 1,000 respondents answering eight four-category items from two correlated factors, fits the CFA with `lavaan` (items ordered) and the graded model with `mirt` (one factor at a time, to keep it quick), and puts both on the logistic IRT scale. The first run takes 20–30 seconds while the packages load.

With the seed as written, every slope from either route is within 0.3 of the truth, and the steepest items miss by the most; `lavaan` puts the factor correlation at 0.36, against a true 0.4. Set `np <- 300` and the slopes scatter; with 5,000 respondents (locally) they settle on the truth.

```{webr}
{{< include code/fa-confirmatory-sim.R >}}
```

[Download this code](code/fa-confirmatory-sim.R).

## With real data

```{r}
#| include: false
knitr::read_chunk("code/fa-confirmatory-irw.R")
options(digits = 7)  # R's default, so the page prints what a fresh R session prints
```

### Grit: one factor or two?

Our main example is [grit](https://itemresponsewarehouse.org/tables/grit/): the 12 items of the Grit-O ([Duckworth, Peterson, Matthews & Kelly, 2007](https://doi.org/10.1037/0022-3514.92.6.1087)), answered online through [Open Psychometrics](https://openpsychometrics.org/_rawdata/). Its job is to test a designed two-facet structure. [Download the code](code/fa-confirmatory-irw.R) to run every chunk.

```{r fetch-grit}
#| echo: true
#| code-fold: show
```

On keying: the source coded 1 = "Very much like me" to 5 = "Not like me at all" (its [codebook](https://openpsychometrics.org/_rawdata/duckworth-grit-scale-data.zip), in the data download), and the IRW's processing script reversed the six consistency items, which are worded against grit. So every item in the table runs toward *less* grit, and reversing all twelve gives higher = more grit. No inter-item correlation is left below −0.03. Because the consistency items are exactly the ones worded against grit, facet and wording direction coincide here (problem 3).

::: {.callout-tip appearance="simple"}
**Recall** from [Factor analysis I](fa-exploratory.qmd): a one-factor model is what "the items are unidimensional" means, and it is a claim the data can reject.
:::

```{r grit-fit}
#| echo: true
#| code-fold: show
```

The data reject it: one factor gives a CFI of 0.758 and an RMSEA of 0.128. The designed facets improve both (0.897 and 0.084, SRMR 0.066), and the difference test is decisive, though by the conventional cutoffs two factors still fall short. The facets correlate 0.61.

```{r grit-identify}
#| echo: true
#| code-fold: show
```

The two ways of setting the scale give the same χ² on 53 degrees of freedom; only the loadings' units change.

```{r grit-omega}
#| echo: true
#| code-fold: show
```

Omega and alpha nearly agree (0.777 and 0.772 for perseverance): the loadings run from 0.44 to 0.86, so the items aren't tau-equivalent, but the gap is small, as the widget suggested.

Now the items as ordered categories, and the same model as a graded response model in `mirt`:

```{r grit-ordinal}
#| echo: true
#| code-fold: show
```

```{r grit-mirt}
#| echo: true
#| code-fold: show
```

The ordinal fit tells the same story in different numbers (CFI 0.906, RMSEA 0.112). The converted slopes track `mirt`'s (correlation 0.94), though single items differ by up to 0.33. The category difficulties agree closely (correlation 1.00, largest gap 0.29), and the factor correlation is 0.62 by one route and 0.61 by the other. Item 11 has the weakest loading, 0.46; [Dimensionality and multidimensional IRT](dimensionality.qmd) meets it again.

### The DASS-21: three scales or one?

Our contrast table is [neurodegenerative_huizinga_2019_dass](https://itemresponsewarehouse.org/tables/neurodegenerative_huizinga_2019_dass/): the Dutch DASS-21, answered by a community sample of adults recruited by [Huizinga et al. (2020)](https://doi.org/10.1371/journal.pone.0232232) for a study of visual complaints. Its job: a structure whose designed scales may not be distinct, and then invariance. The DASS-21 scores depression, anxiety and stress with seven items each ([Lovibond & Lovibond, 1995](https://doi.org/10.1016/0005-7967(94)00075-u); [Henry & Crawford, 2005](https://doi.org/10.1348/014466505x29657)). Every item is worded toward more distress, and the scales follow the DASS [scoring template](https://www2.psy.unsw.edu.au/dass/Download%20files/Dass_template.pdf).

```{r fetch-dass}
#| echo: true
#| code-fold: show
```

We keep the 1,246 respondents who answered all 21 items. Most responses are 0 and only 181 of 26,166 are 3, so the items are declared ordered.

```{ojs}
//| echo: false
predict({
  q: "The DASS-21 is scored as three scales, and a three-factor model fits these data well (CFI 0.975). How strongly will the three factors correlate?",
  choices: ["About 0.3", "About 0.5", "About 0.7", "Above 0.8"],
  answer: "Above 0.8",
  reveal: "Between 0.84 and 0.87. In this sample the three scales are close to one construct: a respondent high on one tends to be high on all three."
})
```

```{r dass-three}
#| echo: true
#| code-fold: show
```

```{r dass-compare}
#| echo: true
#| code-fold: show
```

The factors correlate 0.84 to 0.87. One factor fits a little worse (CFI 0.948 against 0.975), the bifactor model best (0.988). How much of the total score is one general distress factor?

```{r dass-omega}
#| echo: true
#| code-fold: show
```

Omega total is 0.944 and omega hierarchical 0.879: one general factor accounts for almost all of the total score's reliable variance. Within each scale, the specific factor's share of the scale score is 0.24 for depression, 0.23 for anxiety and 0.002 for stress; the stress items carry almost nothing beyond general distress.

Should a clinic report three DASS scores or one? The subscales add information only through their specific factors. In these data those are small, and the DASS authors describe the total as a sensible composite ([DASS FAQ, no. 26](https://www2.psy.unsw.edu.au/dass/DASSFAQ.htm)). For a community sample like this one, I would report the total, and the depression and anxiety scores only alongside it. A clinical sample could give a different answer.

### Invariance: women and men

The table's `cov_sex` codes are unlabelled in the processing script. The paper reports 785 women among its 1,461 participants (its Table 1), and code 2 holds exactly 785 of the table's 1,461 respondents, so we read 2 as women: a match of counts, not a codebook. We fit the three-factor model to both groups with robust maximum likelihood (MLR), items as numbers; the ordinal version follows.

```{r dass-invariance}
#| echo: true
#| code-fold: show
```

Equal loadings cost almost nothing (a scaled Δχ² of 13.5 on 18 df, $p = 0.76$), and neither do equal intercepts (24.6 on 18 df, $p = 0.14$). CFI is 0.931 or 0.932 at every step. The configural fit is modest, but the question is whether the constraints make it worse, and they don't.

```{r dass-constraints}
#| echo: true
#| code-fold: show
```

The constraint under most strain is item 16's intercept ("I was unable to become enthusiastic about anything", in the wording of the [English DASS-21](https://www2.psy.unsw.edu.au/dass/Download%20files/Dass21.pdf); Lovibond & Lovibond, 1995), with a score test of 6.9 on 1 df. That is not unusual among 39 tests: its $p$-value is 0.009, and with 39 independent tests at least one that small turns up by chance about three times in ten ($1 - 0.991^{39} \approx 0.30$).

```{r dass-means}
#| echo: true
#| code-fold: show
```

With scalar invariance the factor means can be compared: women sit 0.16 SD above men on anxiety, 0.18 SD on stress, and 0.09 SD on depression.

```{r dass-ordinal-check}
#| echo: true
#| code-fold: show
#| warning: false
```

As ordered categories, the verdict is the same: CFI from 0.978 to 0.987 and RMSEA at most 0.050 at every step. (Some top categories hold a single response in one group, and with data this sparse the scaled χ² behaves erratically, so we don't report difference tests for this version.) The DASS-21 passes. Problem 4 gives a scale that doesn't.

## Problems

1. **Omega and alpha.** Show that under a one-factor model with equal loadings $\lambda$ and equal unique variances, omega equals alpha. Then show that with unequal loadings (and uncorrelated errors) alpha is smaller.
2. **The probit link.** Starting from $y^* = \lambda\eta + \varepsilon$ with $\text{var}(y^*) = 1$, derive the normal-ogive slope and difficulty of a dichotomized item. Use your result on grit: which facet's items discriminate better, and does `mirt` agree?
3. **A patch or a finding?** Run `modindices()` on the two-factor grit model. Among the largest suggestions is a correlated error between items 3 and 11. Add it. How much does fit improve? Given the codebook's stems ("My interests...", "I become..."), is it a finding about the construct or a patch? Can any model of these data separate consistency of interest from wording direction ([From construct map to items](instrument-building.qmd))?
4. **Where invariance breaks.** Fit a one-factor model to the 20-item CES-D ([Radloff, 1977](https://doi.org/10.1177/014662167700100306)) in [alexandrowicz_2018_cesd](https://itemresponsewarehouse.org/tables/alexandrowicz_2018_cesd/) ([Alexandrowicz, Jahn & Wancata, 2018](https://doi.org/10.1371/journal.pone.0197908)) across `cov_sex`, and climb the ladder. Where does it fail? Find the constraint with the largest score test, free it, and compare the latent means with and without partial invariance. (The paper reports 264 women among 518 respondents; the table has 263 coded 2 among 514, so 2 is probably women, but the processing script doesn't say.)
5. **Design.** You want to compare DASS-21 scores across the education levels in `cov_education`. Which level of invariance do you need to compare means, and which to compare the depression–anxiety correlation? What would you do if one item's intercept failed in one group? Check the group sizes first.
6. **Challenge (open).** [Credé, Tynan and Harms (2017)](https://doi.org/10.1037/pspp0000102) argue from a meta-analysis that grit overlaps heavily with conscientiousness. What confirmatory model would test that, and what data would you need? Given that fit indices miss some kinds of misfit, how would you guard against that? I don't know of a settled answer.

```{r}
#| output: asis
ask_claude("fa-confirmatory", "It covers confirmatory factor analysis with lavaan (model syntax, identification, degrees of freedom), fit indices (chi-square, CFI, TLI, RMSEA, SRMR) and nested model comparison, omega and omega hierarchical in bifactor models, ordinal factor analysis as the normal-ogive graded response model (converting loadings and thresholds to slopes and difficulties, compared with mirt), and multigroup measurement invariance (configural, metric, scalar), with the Grit-O and DASS-21 as examples.")
```

## Going further

- Jöreskog, K. G. (1969). A general approach to confirmatory maximum likelihood factor analysis. *Psychometrika*, 34(2), 183–202. <https://doi.org/10.1007/bf02289343>. Where CFA starts.
- Rosseel, Y. (2012). lavaan: An R package for structural equation modeling. *Journal of Statistical Software*, 48(2). <https://doi.org/10.18637/jss.v048.i02>. The package and its syntax; [Structural equation modeling with lavaan](sem.qmd) adds paths between factors.
- Hu, L., & Bentler, P. M. (1999). Cutoff criteria for fit indexes in covariance structure analysis: Conventional criteria versus new alternatives. *Structural Equation Modeling*, 6(1), 1–55. <https://doi.org/10.1080/10705519909540118>. Where the cutoffs come from, and under which conditions.
- Marsh, H. W., Hau, K.-T., & Wen, Z. (2004). In search of golden rules. *Structural Equation Modeling*, 11(3), 320–341. <https://doi.org/10.1207/s15328007sem1103_2>. Argues against reading the cutoffs as rules.
- Xia, Y., & Yang, Y. (2019). RMSEA, CFI, and TLI in structural equation modeling with ordered categorical data: The story they tell depends on the estimation methods. *Behavior Research Methods*, 51(1), 409–428. <https://doi.org/10.3758/s13428-018-1055-2>. Why the grit indices changed with the estimator.
- McDonald, R. P. (1999). *Test theory: A unified treatment*. Erlbaum. <https://doi.org/10.4324/9781410601087>. Omega, and test theory built on the factor model.
- Revelle, W., & Zinbarg, R. E. (2009). Coefficients alpha, beta, omega, and the glb: Comments on Sijtsma. *Psychometrika*, 74(1), 145–154. <https://doi.org/10.1007/s11336-008-9102-z>. Omega hierarchical beside the other coefficients.
- Reise, S. P. (2012). The rediscovery of bifactor measurement models. *Multivariate Behavioral Research*, 47(5), 667–696. <https://doi.org/10.1080/00273171.2012.715555>. What a bifactor model can say about subscales.
- Rodriguez, A., Reise, S. P., & Haviland, M. G. (2016). Evaluating bifactor models: Calculating and interpreting statistical indices. *Psychological Methods*, 21(2), 137–150. <https://doi.org/10.1037/met0000045>. Omega hierarchical and its relatives, computed and read.
- Green, S. B., & Yang, Y. (2009). Reliability of summed item scores using structural equation modeling: An alternative to coefficient alpha. *Psychometrika*, 74(1), 155–167. <https://doi.org/10.1007/s11336-008-9099-3>. Omega for ordinal items, on the scale of the sum score.
- Takane, Y., & de Leeuw, J. (1987). On the relationship between item response theory and factor analysis of discretized variables. *Psychometrika*, 52(3), 393–408. <https://doi.org/10.1007/bf02294363>. The equivalence, proved.
- Kamata, A., & Bauer, D. J. (2008). A note on the relation between factor analytic and item response theory models. *Structural Equation Modeling*, 15(1), 136–153. <https://doi.org/10.1080/10705510701758406>. The conversions under each parameterization.
- Rhemtulla, M., Brosseau-Liard, P. É., & Savalei, V. (2012). When can categorical variables be treated as continuous? A comparison of robust continuous and categorical SEM estimation methods under suboptimal conditions. *Psychological Methods*, 17(3), 354–373. <https://doi.org/10.1037/a0029315>. The five-category guideline.
- Flora, D. B., & Curran, P. J. (2004). An empirical evaluation of alternative methods of estimation for confirmatory factor analysis with ordinal data. *Psychological Methods*, 9(4), 466–491. <https://doi.org/10.1037/1082-989X.9.4.466>. How WLSMV behaves.
- Vandenberg, R. J., & Lance, C. E. (2000). A review and synthesis of the measurement invariance literature. *Organizational Research Methods*, 3(1), 4–70. <https://doi.org/10.1177/109442810031002>. The ladder, step by step.
- Putnick, D. L., & Bornstein, M. H. (2016). Measurement invariance conventions and reporting: The state of the art and future directions for psychological research. *Developmental Review*, 41, 71–90. <https://doi.org/10.1016/j.dr.2016.06.004>. How invariance is tested and reported in practice.
- Chen, F. F. (2007). Sensitivity of goodness of fit indexes to lack of measurement invariance. *Structural Equation Modeling*, 14(3), 464–504. <https://doi.org/10.1080/10705510701301834>. Which index notices which kind of non-invariance.
- Wu, H., & Estabrook, R. (2016). Identification of confirmatory factor analysis models of different levels of invariance for ordered categorical outcomes. *Psychometrika*, 81(4), 1014–1045. <https://doi.org/10.1007/s11336-016-9506-0>. Invariance with ordinal items.
- Millsap, R. E. (2011). *Statistical approaches to measurement invariance*. Routledge. <https://doi.org/10.4324/9780203821961>. The full treatment, factor models and IRT together.
- From the IRW: the [CFA walkthrough with lavaan](https://itemresponsewarehouse.org/vignettes/cfa.html), and the [dimensionality vignette](https://itemresponsewarehouse.org/vignettes/dimensionality.html) on how often one factor holds.

```{r}
#| output: asis
data_sources("fa-confirmatory")
```

## For instructors

- **Source:** new material. EDUC 252 did not cover confirmatory factor analysis; the lesson was added after the landscape analysis (`notes/landscape-2026-09-24.md`), where CFA appeared in 6 of 11 first-course syllabi. Multigroup invariance moved here from [Differential item functioning](dif.qmd) (Ben, 09-24).
- **Timing:** about 60 minutes for the core ideas and widgets; the real-data section is a 30-minute lab (the two-dimensional graded model takes under a minute locally).
- **Placement:** after [From the 1PL to the 4PL](1pl-to-4pl.qmd), although it sits in the factor-analysis module, because the IRT connection needs the 2PL. `lavaan` syntax is introduced here; [Structural equation modeling with lavaan](sem.qmd) builds on it.
- **Held back:** paths between latent variables and multigroup SEM ([Structural equation modeling with lavaan](sem.qmd)); multidimensional IRT ([Dimensionality and multidimensional IRT](dimensionality.qmd)); the graded response model in its own right ([Models for polytomous responses](polytomous.qmd)); invariance item by item with a matching variable ([Differential item functioning](dif.qmd)); omega for reported scores ([What does a score mean?](score-meaning.qmd)).
- **Data:** the DASS is in the public domain ([DASS FAQ, no. 3](https://www2.psy.unsw.edu.au/dass/DASSFAQ.htm)), so one item is quoted. The Grit-O items are described, not reproduced. Which `cov_sex` code is women in the DASS table rests on a count matching the paper's Table 1, not on a codebook. The invariance ladder is fitted with items as numbers (MLR); the ordinal ladder is shown as a check only, because its difference tests misbehave with sparse top categories.
- **Solutions:** held back in `solutions/fa-confirmatory.qmd` (not published).
