---
title: "Many sources of error: generalizability theory"
---

```{r}
#| output: asis
source("_course.R")
lesson_header("g-theory")
```

```{webr}
#| context: setup
webr::install("lme4", quiet = TRUE)
suppressPackageStartupMessages(library(lme4))
```

```{ojs}
//| echo: false
import {palette, grid} from "./widgets/irt.js"
import {gCoef, coefBars, presets, twoRaters} from "./widgets/g-theory.js"
import {quiz, predict} from "./widgets/quiz.js"
```

## What this is for

Classical test theory gives a score one error term. Suppose 202 people list unusual uses for a paperclip, a garbage bag and a rope, and five raters judge how clever each list is. A respondent might have done better with other objects, and would have got a different score from other raters. Which source of error is larger decides whether to add raters or objects. **Generalizability theory** (G theory) starts from a question CTT leaves implicit: what should the score generalize over? Tasks, raters, both? It then splits the error into those sources and prices each one ([Cronbach, Rajaratnam & Gleser, 1963](https://doi.org/10.1111/j.2044-8317.1963.tb00206.x)). In the cleverness ratings the answer is *more tasks*; in a second table, teachers rating essays, it is *more raters*.

## Goals

By the end of this lesson you should be able to:

- name the facets of a measurement design, say which are crossed, and decide which the score should generalize over (random) and which not (fixed);
- estimate variance components for a persons × tasks × raters design with `lme4`, and say which ones are error for which decision;
- compute relative and absolute G coefficients, and show that the one-facet relative coefficient is alpha;
- use a D study to decide whether to add tasks or raters;
- compute consistency and agreement intraclass correlations, and say when each is the one to report.

## Core ideas

### One error term, many sources

::: {.callout-tip appearance="simple"}
**Recall** from [Classical test theory and reliability](ctt-reliability.qmd): an observed score is $X = T + E$, reliability is $\sigma^2_T/(\sigma^2_T + \sigma^2_E)$, alpha estimates it from the items of one administration, and the Spearman–Brown formula says what lengthening the test does.
:::

Where does $E$ come from? The items behind alpha are one source; [From construct map to items](instrument-building.qmd) added another, the rater who scores a constructed response; occasions and forms are more. G theory calls each such source a **facet**, and each of its values (this rater, that task) a **condition** ([Cronbach, Gleser, Nanda & Rajaratnam, 1972](https://doi.org/10.1126/science.178.4067.1275), a book with no DOI, linked here through Bock's review in *Science*).

::: {.callout-tip appearance="simple"}
**Recall** from [Constructs and construct maps](constructs.qmd): a test built from a blueprint supports a claim about a domain, and its items can be seen as a sample from a larger pool.
:::

G theory applies that view to every facet. The tasks we used are a sample from the tasks we could have used; the raters, from the raters we would accept. Together these make up the **universe of admissible observations**, and the investigator, not the data, decides what belongs in it. The score we would like is a respondent's average over that universe, the **universe score** (G theory's counterpart of the true score). The question that runs through this lesson is how well an average over the few conditions we used stands in for it ([Shavelson, Webb & Rowley, 1989](https://doi.org/10.1037/0003-066X.44.6.922)). A facet we want to generalize over is **random**; a facet we don't (these three objects and no others) is **fixed**, and then what is specific to its conditions stops being error ([Brennan, 2001](https://doi.org/10.1007/978-1-4757-3456-0)).

### Variance components

Take the cleverness design: every person $p$ does every task $t$, and every rater $r$ rates every response, so all three are **crossed**. A rating is a sum of effects:
$$X_{ptr} = \mu + \nu_p + \nu_t + \nu_r + \nu_{pt} + \nu_{pr} + \nu_{tr} + \nu_{ptr,e}.$$
Each $\nu$ is a random effect with mean zero and its own variance, and the effects are uncorrelated, so the variance of one rating splits into seven **variance components**, $\sigma^2_p + \sigma^2_t + \dots + \sigma^2_{ptr,e}$ ([Brennan, 2001](https://doi.org/10.1007/978-1-4757-3456-0)). $\sigma^2_p$ is how much people differ, the signal; $\sigma^2_r$, how much raters differ in severity; $\sigma^2_{pt}$, how much people's standing changes from task to task; $\sigma^2_{pr}$, raters disagreeing about who is better. The last term mixes the three-way interaction with whatever else the design can't separate from it, hence the $e$.

How do we estimate them? Classically, by setting the mean squares of an analysis of variance equal to their expected values, a reasonable choice for a balanced design ([Brennan, 2001](https://doi.org/10.1007/978-1-4757-3456-0)). This lesson fits the model directly: each term is a random intercept in a mixed model. With `lme4` ([Bates, Mächler, Bolker & Walker, 2015](https://doi.org/10.18637/jss.v067.i01)) and a long table with one row per rating:

```r
lmer(resp ~ 1 + (1 | id) + (1 | item) + (1 | rater) +
       (1 | id:item) + (1 | id:rater) + (1 | item:rater), data = ratings)
```

The residual variance is $\sigma^2_{ptr,e}$. Why this route? `lmer` estimates the variances by restricted maximum likelihood (REML). Plain maximum likelihood treats the fixed effects as known and so underestimates each variance, by more the fewer levels a facet has; REML accounts for the degrees of freedom the fixed effects use and removes that bias ([Patterson & Thompson, 1971](https://doi.org/10.1093/biomet/58.3.545); [Harville, 1977](https://doi.org/10.1080/01621459.1977.10480998)). With facets of three and five levels, that matters. In a balanced design REML reproduces the ANOVA estimates whenever none is negative ([Searle, Casella & McCulloch, 1992](https://doi.org/10.1002/9780470316856)), and it also handles the missing ratings and sparse designs the second table has.

### Two G coefficients

Why are some components error only some of the time? It depends on the decision. For a **relative** decision (who ranks above whom), only the components that change a person's standing count: those that interact with $p$. For a score averaged over $n_t$ tasks and $n_r$ raters, the relative error variance is
$$\sigma^2_\delta = \frac{\sigma^2_{pt}}{n_t} + \frac{\sigma^2_{pr}}{n_r} + \frac{\sigma^2_{ptr,e}}{n_t n_r}.$$
For an **absolute** decision (is this score above 3? how clever is this list, on the rating scale?), a harsh rater or a hard task moves the score itself, so the main effects count too:
$$\sigma^2_\Delta = \sigma^2_\delta + \frac{\sigma^2_t}{n_t} + \frac{\sigma^2_r}{n_r} + \frac{\sigma^2_{tr}}{n_t n_r}.$$
Each coefficient sets the universe-score variance against one of them ([Brennan, 2001](https://doi.org/10.1007/978-1-4757-3456-0)): the **generalizability coefficient** $E\rho^2 = \sigma^2_p/(\sigma^2_p + \sigma^2_\delta)$ and the **dependability coefficient** $\Phi = \sigma^2_p/(\sigma^2_p + \sigma^2_\Delta)$. Since $\sigma^2_\Delta \ge \sigma^2_\delta$, $\Phi$ is never the larger. Error near a cut score is taken up in [What does a score mean?](score-meaning.qmd).

Both have the form of reliability in [Classical test theory and reliability](ctt-reliability.qmd), $\sigma^2_T/(\sigma^2_T + \sigma^2_E)$. The "true" variance is universe-score variance, $\sigma^2_p$: what stays with a person across every condition we generalize over. The error is what changes from one sample of conditions to another: $\sigma^2_\delta$ for rankings, $\sigma^2_\Delta$ for levels. Fix a facet and the split moves. With the tasks fixed, $\sigma^2_{pt}/n_t$ belongs to the universe score rather than the error ([Brennan, 2001](https://doi.org/10.1007/978-1-4757-3456-0)); the interrater section uses this.

#### Try it: the two coefficients side by side

Each bar is one coefficient's denominator: the universe score and the error that coefficient counts. The light pieces (task, rater, task × rater) count only for Φ. The sliders start at the cleverness ratings' estimates from *With real data*; switch to the essays to see raters that differ in severity. (For the essays, "task" is a rating criterion.)

```{ojs}
//| echo: false
viewof gpre = Inputs.radio(Object.keys(presets), {value: "Cleverness ratings", label: "Start from"})
```

```{ojs}
//| echo: false
viewof wv = Inputs.form({
  p: Inputs.range([0, 1], {value: presets[gpre].p, step: 0.001, label: "person"}),
  t: Inputs.range([0, 1], {value: presets[gpre].t, step: 0.001, label: "task"}),
  r: Inputs.range([0, 1], {value: presets[gpre].r, step: 0.001, label: "rater"}),
  pt: Inputs.range([0, 1], {value: presets[gpre].pt, step: 0.001, label: "person × task"}),
  pr: Inputs.range([0, 1], {value: presets[gpre].pr, step: 0.001, label: "person × rater"}),
  tr: Inputs.range([0, 1], {value: presets[gpre].tr, step: 0.001, label: "task × rater"}),
  res: Inputs.range([0, 1], {value: presets[gpre].res, step: 0.001, label: "residual"})
})
viewof wn = Inputs.form({
  nt: Inputs.range([1, 10], {value: presets[gpre].nt, step: 1, label: "Tasks averaged over"}),
  nr: Inputs.range([1, 10], {value: presets[gpre].nr, step: 1, label: "Raters averaged over"})
})
```

```{ojs}
//| echo: false
cb = coefBars(wv, wn.nt, wn.nr)
Plot.plot({
  width: 700, height: 190, marginLeft: 150,
  x: {label: "Variance (universe score + error)", grid: true}, y: {label: null, domain: [cb.rel, cb.abs]},
  color: {legend: true, domain: ["universe score", "relative and absolute error", "absolute error only"],
          range: [palette.main, palette.contrast, palette.light]},
  marks: [
    Plot.barX(cb.rows, {x: "value", y: "coef", fill: "role", stroke: palette.white, title: d => `${d.name}: ${d.value.toFixed(3)}`}),
    Plot.text(cb.rows, Plot.stackX({x: "value", y: "coef", text: d => d.value > 0.05 ? d.name : "", fill: d => d.role === "absolute error only" ? palette.ink : palette.white, fontSize: 10}))
  ]
})
```

```{ojs}
//| echo: false
md`Averaged over ${wn.nt} task${wn.nt > 1 ? "s" : ""} and ${wn.nr} rater${wn.nr > 1 ? "s" : ""}: relative error **${cb.g.rel.toFixed(3)}**, absolute error **${cb.g.abs.toFixed(3)}**; Eρ² **${cb.g.erho2.toFixed(2)}**, Φ **${cb.g.phi.toFixed(2)}**.`
```

The coefficients agree when raters and tasks don't differ in harshness or difficulty (cleverness: 0.71 and 0.71) and part when they do (essays, one teacher: 0.47 and 0.40). So use Φ when the decision is absolute (a cut) or raters vary across respondents, and Eρ² for ranking in a fully crossed design, the rule of this lesson's verdict (end of the interrater section).

```{ojs}
//| echo: false
quiz({
  q: "In a persons × raters design, raters differ a lot in severity (large $\\sigma^2_r$) but agree closely on who is better (small $\\sigma^2_{pr}$). Every respondent is scored by the same two raters, and scores are used only to rank respondents. Does rater severity add error to the ranking?",
  options: [
    {text: "No: a harsh rater lowers everyone's score by the same amount", correct: true, why: "When every respondent has the same raters, severity shifts all scores equally and leaves the order alone. It counts for absolute decisions, or when respondents get different raters."},
    {text: "Yes: $\\sigma^2_r$ is large, so the scores are unreliable", correct: false, why: "A large $\\sigma^2_r$ moves every score together here. What would disturb the ranking is $\\sigma^2_{pr}$, raters disagreeing about who is better, and that is small."},
    {text: "Only if the raters were trained differently", correct: false, why: "Training might explain why severity differs, but the question is what severity does to a ranking when raters are crossed with respondents: nothing."}
  ]
})
```


What does the simplest case look like? Keep one facet, items, in a persons × items design. Then $\sigma^2_\delta = \sigma^2_{pi,e}/n_i$ and
$$E\rho^2 = \frac{\sigma^2_p}{\sigma^2_p + \sigma^2_{pi,e}/n_i},$$
which is Cronbach's alpha ([Cronbach, 1951](https://doi.org/10.1007/BF02310555); [Hoyt, 1941](https://doi.org/10.1007/BF02289270), estimated reliability from the same analysis of variance). Alpha is the one-facet relative G coefficient; the callout shows why.

::: {.callout-note collapse="true"}
## Go deeper: why the one-facet coefficient is alpha

**Assumptions:** a persons × items design with $n_p$ persons and $n_i$ items, one score per cell, $X_{pi} = \mu + \nu_p + \nu_i + \nu_{pi,e}$, with uncorrelated random effects ([Brennan, 2001](https://doi.org/10.1007/978-1-4757-3456-0)).

In a two-way analysis of variance without replication, $E(MS_p) = \sigma^2_{pi,e} + n_i\sigma^2_p$ and $E(MS_{pi,e}) = \sigma^2_{pi,e}$, so $\hat\sigma^2_p = (MS_p - MS_{pi,e})/n_i$, $\hat\sigma^2_{pi,e} = MS_{pi,e}$, and
$$\widehat{E\rho^2} = \frac{\hat\sigma^2_p}{\hat\sigma^2_p + \hat\sigma^2_{pi,e}/n_i} = 1 - \frac{MS_{pi,e}}{MS_p}.$$

**Why it is alpha.** Write $X_p = \sum_i X_{pi}$ for a person's total and $X_{\cdot i}$ for item $i$'s column, and use sample variances with divisor $n_p - 1$. The persons sum of squares is $n_i$ times the sum of squared deviations of person means, so $MS_p = \text{Var}(X_p)/n_i$. The residual sum of squares is what is left of the total after the person and item parts, $(n_p - 1)\big[\sum_i \text{Var}(X_{\cdot i}) - \text{Var}(X_p)/n_i\big]$, on $(n_p-1)(n_i-1)$ degrees of freedom. So
$$\frac{MS_{pi,e}}{MS_p} = \frac{n_i\sum_i \text{Var}(X_{\cdot i}) - \text{Var}(X_p)}{(n_i - 1)\,\text{Var}(X_p)},$$
and
$$1 - \frac{MS_{pi,e}}{MS_p} = \frac{n_i}{n_i - 1}\left(1 - \frac{\sum_i \text{Var}(X_{\cdot i})}{\text{Var}(X_p)}\right) = \alpha. \qquad\blacksquare$$
The equality holds exactly in any sample, not only in expectation. Problem 1 asks you to fill in the sums of squares.
:::

```{ojs}
//| echo: false
quiz({
  q: "A persons × items G study of a 12-item test gives $\\sigma^2_p = 0.5$ and $\\sigma^2_{pi,e} = 3.0$. What is the relative G coefficient of the 12-item test?",
  options: [
    {text: "0.67, the number alpha would give", correct: true, why: "$0.5/(0.5 + 3.0/12) = 0.5/0.75 = 0.67$. In a one-facet design the relative G coefficient is alpha."},
    {text: "0.14", correct: false, why: "That is one item: $0.5/(0.5 + 3.0)$. The test averages over 12 items, so the interaction is divided by 12."},
    {text: "0.67, but it is a test-retest reliability, not alpha", correct: false, why: "Occasions aren't a facet of this design. The only facet is items, so the coefficient is the internal-consistency one, alpha."}
  ]
})
```

### D studies: what would another design buy?

A **decision study** (D study) uses the G study's components to ask what a different design would give: change $n_t$ and $n_r$ in the formulas and recompute. With one facet this is Spearman–Brown (problem 1); with two, the components say which lever moves the coefficient more.

#### Try it: a D study

The components are the ones set in the previous widget. Each line is a number of raters; the horizontal axis is the number of tasks.

```{ojs}
//| echo: false
viewof dcoef = Inputs.radio(["relative", "absolute"], {value: "relative", label: "Coefficient"})
```

```{ojs}
//| echo: false
d_lines = [1, 2, 3, 5].flatMap(nr => grid(1, 10, 10).map(nt => {
  const g = gCoef(wv, nt, nr);
  return {nt, nr: `${nr} rater${nr > 1 ? "s" : ""}`, g: dcoef === "relative" ? g.erho2 : g.phi};
}))
Plot.plot({
  width: 700, height: 300, x: {label: "Tasks averaged over", ticks: 10}, y: {label: `${dcoef === "relative" ? "Relative" : "Absolute"} G coefficient`, domain: [0, 1]},
  color: {legend: true, domain: ["1 rater", "2 raters", "3 raters", "5 raters"], range: [palette.light, palette.main, palette.contrast, palette.guide]},
  marks: [Plot.line(d_lines, {x: "nt", y: "g", stroke: "nr", strokeWidth: 2.5}), Plot.dot(d_lines, {x: "nt", y: "g", fill: "nr", r: 3}),
          Plot.ruleY([0.8], {stroke: palette.guide, strokeDasharray: "4 3"})]
})
```

```{ojs}
//| echo: false
md`Ten tasks with one rater: **${gCoef(wv, 10, 1).erho2.toFixed(2)}** (relative). One task with ten raters: **${gCoef(wv, 1, 10).erho2.toFixed(2)}**. *The cleverness data had 3 tasks and 5 raters, the essays 4 criteria and 1 to 7 teachers per essay: points beyond those are projections, which assume new tasks and raters are exchangeable with the ones observed.*`
```

With the cleverness components, ten tasks with one rater reach 0.74, one task with ten raters 0.49: person × task (0.284) dwarfs person × rater (0.053). Switch the previous widget to the essays and raters become the lever. Every curve flattens: past a point only the *other* facet helps.

What are we assuming when a D study adds raters we never had, ten when the data had five? That new raters are exchangeable draws from the same universe (same pool, same training); that the variance components hold, so a new rater disagrees with the others as much as the old ones did; and that the design stays crossed, every new rater scoring every response. If the new raters come from a different pool, $\sigma^2_r$ and $\sigma^2_{pr}$ change and the curve is wrong; if each respondent gets different raters, severity joins the relative error (next section). And the components themselves were estimated from a handful of raters, so a point far beyond the observed design is a projection that carries that uncertainty (Simulate shows how noisy a facet with three levels is).

### Interrater reliability as a G study

In a persons × raters design, the setting for **interrater reliability**, the intraclass correlations of [Shrout and Fleiss (1979)](https://doi.org/10.1037/0033-2909.86.2.420) and [McGraw and Wong (1996)](https://doi.org/10.1037/1082-989X.1.1.30) are G coefficients. The **consistency** ICC for one rater, ICC(C,1) in McGraw and Wong's notation, is $\sigma^2_p/(\sigma^2_p + \sigma^2_{pr,e})$, the relative coefficient. The **absolute agreement** ICC, ICC(A,1), is $\sigma^2_p/(\sigma^2_p + \sigma^2_r + \sigma^2_{pr,e})$, the dependability coefficient. Averaged over $k$ raters, they are the D study for $k$ raters.

When raters score the same tasks, what counts as true? If the question is only whether another rater would score the same response the same way, the tasks are fixed. [Brennan (2001)](https://doi.org/10.1007/978-1-4757-3456-0) then puts $\sigma^2_{pt}$ in the universe score: for one rater on one task the coefficient is $(\sigma^2_p + \sigma^2_{pt})/(\sigma^2_p + \sigma^2_{pt} + \sigma^2_{pr} + \sigma^2_{ptr,e})$. It answers a narrower question than a coefficient that generalizes over tasks too.

#### Try it: consistency and agreement

Forty simulated respondents, two raters. Rater B is harsher by a constant; both add rating noise.

```{ojs}
//| echo: false
viewof ic_h = Inputs.range([0, 2], {value: 1, step: 0.1, label: "How much harsher rater B is"})
viewof ic_n = Inputs.range([0.1, 1.5], {value: 0.5, step: 0.05, label: "Rating noise (SD)"})
```

```{ojs}
//| echo: false
ic = twoRaters({seed: 76, n: 40, harsh: ic_h, noise: ic_n})
Plot.plot({
  width: 420, height: 420, aspectRatio: 1,
  x: {label: "Rater A", domain: [-5, 5]}, y: {label: "Rater B", domain: [-5, 5]},
  marks: [Plot.line([[-5, -5], [5, 5]], {stroke: palette.guide, strokeDasharray: "4 3"}),
          Plot.dot(ic.pts, {x: "a", y: "b", r: 3.5, fill: palette.main, fillOpacity: 0.7})]
})
```

```{ojs}
//| echo: false
md`Consistency ICC(C,1): **${ic.consistency.toFixed(2)}**. Absolute agreement ICC(A,1): **${ic.agreement.toFixed(2)}**. Pearson correlation: ${ic.r.toFixed(2)}. Rater B scores ${ic.meanGap.toFixed(2)} lower on average.`
```

At the starting values consistency is 0.84 and agreement 0.62: the points lie along a line, below the dashed line of equal ratings. At harshness 0 the two meet; at 2 agreement falls to 0.35 while consistency stays at 0.84. Agreement charges for the gap.

```{ojs}
//| echo: false
quiz({
  q: "Rater B gives every essay exactly one point less than rater A. What are the two single-rater ICCs?",
  options: [
    {text: "Consistency 1; agreement below 1", correct: true, why: "One rater's scores are a shift of the other's, so $\\sigma^2_{pr,e} = 0$ and consistency is perfect. The shift is rater variance, which only the agreement ICC counts."},
    {text: "Both 1", correct: false, why: "The raters never give the same score, so they don't agree. Agreement counts the one-point gap as error."},
    {text: "Consistency below 1; agreement 1", correct: false, why: "It is the other way round: agreement is the stricter of the two, since its error term includes rater severity."}
  ]
})
```

Which should you report? The choice rests on two facts about the use. If every respondent is scored by the same raters and only the order matters, severity cancels and consistency is the coefficient that describes the scores. If scores are compared with a fixed standard, or respondents are scored by different raters (each essay by whichever teacher it was assigned to), a harsh rater lowers some respondents' scores and not others', and rater variance is error even for ranking ([Brennan, 2001](https://doi.org/10.1007/978-1-4757-3456-0), treats these nested designs). I report the absolute coefficient, $\Phi$ or the agreement ICC, whenever raters vary across respondents or a cut is involved, and the consistency coefficient only for a fully crossed design whose scores are used to rank.

## Simulate

The code generates a persons × tasks × raters design with known variance components (150 persons, 3 tasks, 3 raters, with components loosely like the cleverness ratings below), fits the crossed random-effects model with `lme4`, and compares the estimates and the G coefficients with the truth. Loading `lme4` takes 10 to 20 seconds the first time the page runs code; the fit itself takes about two.

With the seed as written, the person components come back close (person 0.346 against 0.30; person × task 0.317 against 0.30), and so does the relative coefficient (0.692 against 0.668). The task component does not (0.065 against 0.02): with three tasks, its variance is estimated from three numbers, however many persons there are. Set `nt <- 10` and the coefficient rises to 0.842, though the task component is still off (0.067). Put `nt` back and set `nr <- 10`: the coefficient rises only to 0.723, since person × rater is small.

```{webr}
{{< include code/g-theory-sim.R >}}
```

[Download this code](code/g-theory-sim.R).

## With real data

```{r}
#| include: false
knitr::read_chunk("code/g-theory-irw.R")
```

```{r helpers}
#| echo: true
#| code-fold: true
#| message: false
```

### Cleverness ratings: tasks or raters?

The main example is [Forthmann-2024-cleverness_ratings](https://itemresponsewarehouse.org/tables/forthmann-2024-cleverness_ratings/), released by [Forthmann and Myszkowski (2024)](https://osf.io/a9qnc) for a special issue of the *Journal of Intelligence* (CC BY 4.0; the project page points to [Forthmann et al., 2017](https://doi.org/10.1080/10400419.2017.1360059), as an earlier analysis of these data). Respondents listed unusual uses for three objects (the alternate uses task), and five raters rated each respondent's list for each object for cleverness, on a scale stored as 1 to 5, with higher meaning cleverer. ([Silvia et al., 2008](https://doi.org/10.1037/1931-3896.2.2.68), study the reliability of such ratings.) The table's job: a fully crossed persons × tasks × raters G study. [Download the code](code/g-theory-irw.R); it needs `lme4`.

```{r fetch-clever}
#| echo: true
#| code-fold: show
```

Ten ratings are missing, all for one respondent; `lmer` handles the imbalance.

```{r fit-clever}
#| echo: true
#| code-fold: show
```

People differ (person, 32.1% of the variance of one rating), but person × task is nearly as large (29.4%): being clever with a rope says only so much about being clever with a paperclip. Raters account for little: 2.1% for severity and 5.5% for disagreement about who is cleverer. The task component is estimated at zero, so the three objects are about equally hard to be clever with. How much did REML matter? Refit by plain maximum likelihood:

```{r reml-ml}
#| echo: true
#| code-fold: show
```

With five raters, maximum likelihood puts rater severity at about four-fifths of the REML value, the downward bias REML removes.

```{ojs}
//| echo: false
predict({
  q: "The design as run uses 3 tasks and 5 raters: 15 ratings per respondent. Suppose instead each respondent did 6 tasks, each scored by one rater: 6 ratings. What happens to the relative G coefficient?",
  choices: ["It falls a lot", "It falls a little", "It rises"],
  answer: "It falls a little",
  reveal: "From 0.71 to 0.68, with fewer than half the ratings. Person × task (29.4% of a single rating's variance) is far larger than person × rater (5.5%), so tasks buy more than raters."
})
```

```{r dstudy-clever}
#| echo: true
#| code-fold: show
```

```{r plot-clever}
#| echo: true
#| code-fold: true
#| fig-width: 7
#| fig-height: 4.5
```

The design as run, 3 tasks and 5 raters, has a relative G coefficient of 0.71. One rater gives 0.56, but one rater with 6 tasks gives 0.68: 6 ratings per respondent instead of 15, for nearly the same coefficient. No number of raters takes 3 tasks past 0.77, the ceiling set by person × task. For this construct and these raters, the money goes on tasks. The absolute coefficient equals the relative one to two decimals (0.71), since the task and rater components are small.

If the tasks are fixed instead, as in the interrater coefficient of [Brennan (2001)](https://doi.org/10.1007/978-1-4757-3456-0), person × task joins the universe score:

```{r interrater}
#| echo: true
#| code-fold: show
```

A second rater would score the same list much as the first did (0.64), but a respondent's cleverness with one object says much less about their cleverness in general (0.33). Same ratings, two universes.

### Essays: when raters matter

The contrast is a table where raters are a large source of error: [teacherjudgements_lohmann_2026_essayratings](https://itemresponsewarehouse.org/tables/teacherjudgements_lohmann_2026_essayratings/), from a study in which preservice English teachers rated essays by learners of English as a second language ([Lohmann et al., 2026](https://doi.org/10.1037/edu0000969); CC BY 4.0). Each teacher rated five essays on four criteria (language quality, structure, content, and a holistic rating), from 1 (very low quality) to 7 (very high quality). "Person" in the output now means essay, and criteria take the place of tasks. We set aside the table's expert benchmark scores; the question is about the teachers.

```{r fetch-essays}
#| echo: true
#| code-fold: show
```

This design is sparse. Every teacher rated 5 essays, but 435 of the 881 essays were rated by a single teacher, and only 50 by four or more. Most essay–teacher pairs were never observed; the components rest on the essays more than one teacher rated, and on each teacher's five essays.

```{r fit-essays}
#| echo: true
#| code-fold: show
```

The picture reverses. Teachers differ in severity (rater, 9.7% of the variance, against 2.1% in the cleverness ratings), and essay × rater is 25.2%, against the essays' own 29.7%: teachers disagree about which essays are better. The criteria matter less (essay × criterion, 9.6%). Since each teacher rated an essay on four criteria, essay × rater is an impression that carries across all four, the pattern a halo effect (named in Lohmann et al.'s title) would produce. On the rating scale, teachers' severity has an SD of 0.50 points, against 0.87 for essays' universe scores.

```{r dstudy-essays}
#| echo: true
#| code-fold: show
```

With one teacher and four criteria, the relative coefficient is 0.47 and the absolute 0.40. A second teacher brings them to 0.62 and 0.56; five, to 0.78 and 0.73. One teacher on unlimited criteria reaches only 0.54: the lever is raters. Since each essay had its own teachers, the absolute coefficient describes these scores.

The two tables call for different designs, and only their variance components could say so.

## Problems

1. **Derivation.** Fill in the sums of squares in the Go deeper callout to show that $1 - MS_{pi,e}/MS_p$ is alpha. Then show that the D-study coefficient for $m n_i$ items is the Spearman–Brown formula with multiplier $m$.
2. **Real data with a twist.** From the cleverness components, find the combinations of tasks and raters that give a relative G coefficient of at least 0.8. Is there more than one? If a task costs a respondent five minutes and a rating costs a rater one minute per respondent, which design is cheapest for 200 respondents? Does the answer change for $\Phi$?
3. **Judgment: a sparse design.** Refit the essay model using only essays with at least two teachers. Which components change? Suppose every essay had one teacher: which components could the data no longer separate, and what would that do to your D study?
4. **Design.** Plan a G study for a performance assessment you know. What should the score generalize over? Name the facets, random or fixed, crossed or nested, and choose sample sizes. Which component do you expect to be largest?
5. **Real data: two raters.** In the cleverness data, take raters 1 and 2 and the paperclip task. Compute the consistency and agreement ICCs from a `lmer` fit with random intercepts for respondent and rater. Which pair of raters agrees least?
6. **Challenge (open).** G theory treats a harsh rater as a random effect to be averaged away. When should severity be estimated and adjusted for instead, and what would you need to know about the essays' teachers to decide? ([Rater models](rater-models.qmd) takes this up with the many-facet Rasch model.)

```{r}
#| output: asis
ask_claude("g-theory", "It covers generalizability theory: facets and universe scores, variance components for a persons x tasks x raters design estimated as crossed random effects with lme4, relative and absolute error and the two G coefficients (with the one-facet coefficient equal to Cronbach's alpha and an expected-mean-squares derivation), D studies, and interrater reliability as a G study (consistency and agreement ICCs), with cleverness ratings of alternate-uses responses and teachers' ratings of essays as examples.")
```

## Going further

- Brennan, R. L. (2001). *Generalizability theory*. Springer. <https://doi.org/10.1007/978-1-4757-3456-0>. The reference: designs, nested and fixed facets, estimation and D studies, at book length.
- Shavelson, R. J., & Webb, N. M. (1991). *Generalizability theory: A primer*. Sage. No DOI; see [Sundre's (1993) review](https://doi.org/10.1177/109821409301400219). A gentler book-length introduction.
- Shavelson, R. J., Webb, N. M., & Rowley, G. L. (1989). Generalizability theory. *American Psychologist*, 44(6), 922–932. <https://doi.org/10.1037/0003-066X.44.6.922>. A short overview.
- Brennan, R. L. (2000). Performance assessments from the perspective of generalizability theory. *Applied Psychological Measurement*, 24(4), 339–353. <https://doi.org/10.1177/01466210022031796>. Which facets matter for performance assessments, and how popular reliability estimates for them, interrater reliability among them, look from G theory.
- Brennan, R. L. (1992). Generalizability theory. *Educational Measurement: Issues and Practice*, 11(4), 27–34. <https://doi.org/10.1111/j.1745-3992.1992.tb00260.x>. An instructional module on G and D studies.
- Cronbach, L. J., Rajaratnam, N., & Gleser, G. C. (1963). Theory of generalizability: A liberalization of reliability theory. *British Journal of Statistical Psychology*, 16(2), 137–163. <https://doi.org/10.1111/j.2044-8317.1963.tb00206.x>. Where the theory starts.
- McGraw, K. O., & Wong, S. P. (1996). Forming inferences about some intraclass correlation coefficients. *Psychological Methods*, 1(1), 30–46. <https://doi.org/10.1037/1082-989X.1.1.30>. The ICCs, sorted by design and by consistency versus agreement.
- Bates, D., Mächler, M., Bolker, B., & Walker, S. (2015). Fitting linear mixed-effects models using lme4. *Journal of Statistical Software*, 67(1). <https://doi.org/10.18637/jss.v067.i01>. How `lmer` specifies and fits crossed random effects.
- Eckes, T. (2011). *Introduction to many-facet Rasch measurement*. Peter Lang. <https://doi.org/10.3726/978-3-653-04844-5>. Raters as a facet in an item response model, which estimates each rater's severity and adjusts scores for it; [Rater models](rater-models.qmd) develops it on this lesson's two tables.
- De Boeck, P. (2008). Random item IRT models. *Psychometrika*, 73(4), 533–559. <https://doi.org/10.1007/s11336-008-9092-x>. Items as random effects in an item response model, an idea [What is an item?](explanatory-irt.qmd) takes further.

```{r}
#| output: asis
data_sources("g-theory")
```

## For instructors

- **Source:** new material, not in EDUC 252. The course's landscape analysis (`notes/landscape-2026-09-24.md`, gap 2) put G theory and interrater reliability into one lesson.
- **Default, to be revised (#76):** the main line estimates variance components with `lme4`, and the ANOVA / expected-mean-squares route is a pointer plus a short Go deeper (the one-facet coefficient is alpha). A colleague consulted on #76 preferred `lmer` (REML) and asked for more on what the score generalizes over, the assumptions behind D-study projections, and Brennan's interrater coefficient with tasks fixed; those are in this version.
- **Timing:** about 50 minutes for the core ideas and widgets; the two real-data analyses work as a 25-minute lab (the essay fit takes about half a minute in local R).
- **Held back:** nested designs are named, not developed, and fixed facets appear only in the interrater coefficient (see Brennan, 2001); multivariate G theory is not covered. Raters in item response models (many-facet Rasch) are left to [Rater models](rater-models.qmd); random item effects to [What is an item?](explanatory-irt.qmd); error near a cut to [What does a score mean?](score-meaning.qmd); occasions as a facet to [Measurement invariance under treatment and life events](invariance-experience.qmd).
- **Solutions:** held back in `solutions/g-theory.qmd` (not published). Problem 6 is open.
