---
title: "Guessing and priors"
---

```{r}
#| output: asis
source("_course.R")
lesson_header("guessing-priors")
```

```{webr}
#| context: setup
{{< include code/_webr-setup.R >}}
```

```{ojs}
//| echo: false
import {grid, palette} from "./widgets/irt.js"
import {p4pl} from "./widgets/1pl-to-4pl.js"
import {se3pl, shareNearFloor, slopePosterior} from "./widgets/guessing-priors.js"
import {quiz, predict} from "./widgets/quiz.js"
```

## What this is for

On the Reading the Mind in the Eyes Test, where chance is one in four, the 3PL's median lower asymptote was 0.01; on a reading test, the 3PL predicted new responses no better than the 2PL. Yet respondents do guess. So what should we do about it? On tests like the ones in this lesson, I wouldn't estimate $c$ from the data alone. Where the chance rate is known, I'd fix it; where it isn't, I'd fit the 2PL; either way I'd let held-out responses decide whether the choice earned its place. This lesson makes the case: it asks what the lower asymptote learns from, fixes it at chance on a two-button reading task, finds the guessing there concentrated in a few respondents, and uses priors to keep a small-sample 3PL in bounds.

## Goals

By the end of this lesson you should be able to:

- say which respondents the data use to estimate a lower asymptote, and predict when its estimate will be imprecise;
- fit a Rasch model with the lower asymptote fixed at chance, and say what it assumes;
- use response times to find rapid responses, and explain why guessing can be a property of respondents rather than items;
- choose lognormal and beta priors for slopes and lower asymptotes, and say what they do to the estimates;
- judge a guessing correction or a prior by out-of-sample prediction.

## Core ideas

### What does the lower asymptote learn from?

::: {.callout-tip appearance="simple"}
**Recall** from [From the 1PL to the 4PL](1pl-to-4pl.qmd): the lower asymptote $c$ describes respondents far below an item. On the RMET only 84 of 17,680 respondents (0.48%) scored at or below chance, and the 3PL's median $c$ was 0.01 although chance is 0.25.
:::

Why so little? A response carries information about $c$ only where $c$ matters to its probability: where the logistic part of the curve is near zero. A respondent well above an item answers it correctly whatever $c$ is. So the precision of $\hat c$ is set by how many respondents find the item too hard, not by how many respondents there are.

#### Try it: where the lower asymptote is learned

One 3PL item ($a = 1.2$, $b = 0$), answered by $n$ respondents whose abilities are normal around a mean you choose, relative to the item. The curve is the standard error of the estimated $c$, from the Fisher information for $(a, b, c)$ with the abilities known. With estimated abilities it would be larger.

```{ojs}
//| echo: false
viewof lc_mu = Inputs.range([-2, 2], {value: 1.5, step: 0.1, label: "Mean ability, relative to the item"})
viewof lc_n = Inputs.range([100, 20000], {value: 1000, step: 100, label: "Respondents (n)"})
viewof lc_c = Inputs.range([0, 0.4], {value: 0.25, step: 0.01, label: "True lower asymptote c"})
```

```{ojs}
//| echo: false
lc_curve = grid(-2, 2, 81).map(mu => ({mu, se: se3pl(1.2, 0, lc_c, mu, lc_n).c}))
lc_now = se3pl(1.2, 0, lc_c, lc_mu, lc_n)
lc_share = shareNearFloor(1.2, 0, lc_mu)
Plot.plot({
  width: 700, height: 300,
  x: {label: "Mean ability, relative to the item"},
  y: {label: "Standard error of ĉ", domain: [0, Math.min(1, Math.max(0.3, ...lc_curve.map(d => d.se)))]},
  marks: [
    Plot.ruleY([0], {stroke: palette.rule}),
    Plot.line(lc_curve, {x: "mu", y: "se", stroke: palette.main, strokeWidth: 2.5}),
    Plot.dot([{mu: lc_mu, se: lc_now.c}], {x: "mu", y: "se", r: 6, fill: palette.contrast})
  ]
})
```

```{ojs}
//| echo: false
md`${(100 * lc_share).toFixed(1)}% of respondents sit where the item's logistic part is below 0.1, so that their correct answers are mostly chance. The standard error of ĉ is **${lc_now.c.toFixed(3)}**, so a 95% interval runs from ${Math.max(0, lc_c - 1.96 * lc_now.c).toFixed(2)} to ${Math.min(1, lc_c + 1.96 * lc_now.c).toFixed(2)}. For comparison, the slope's standard error is ${lc_now.a.toFixed(2)}.`
```

As the widget loads, the item sits 1.5 below the mean ability, as many RMET items did, and almost nobody is near the floor: with 1,000 respondents the standard error of $\hat c$ is 0.27, and the interval runs from 0 to well past chance. Slide the mean down to −1 and a fifth of the respondents are near the floor; the standard error falls to 0.05. Back at 1.5, even 20,000 respondents leave it at about 0.06. The slope's standard error barely moves with the mean.

::: {.callout-tip appearance="simple"}
**Recall** from [Model fit and out-of-sample prediction](fit-prediction.qmd): the IMV turns two sets of predictions of held-out responses into the expected return on a fair bet, which makes predictive gains comparable across datasets. On a reading comprehension test, the 3PL's IMV over the 2PL was 0.0001 out of sample and 0.0035 in-sample.
:::

[Domingue et al. (2024, §3)](https://doi.org/10.1007/s11336-024-09977-2) asked the question by simulation. They generated responses from a 3PL with lower asymptotes drawn between 0 and 0.3, fitted the 1PL, 2PL and 3PL, and computed IMVs on held-out responses. The average IMV of the 3PL over the 2PL never exceeded 0.001, including when the 3PL generated the data, while the 2PL gained 0.01 or more over the 1PL when slopes varied substantially. They relate this to known difficulties in identifying the 3PL's parameters ([Maris & Bechger, 2009](https://doi.org/10.1080/15366360903070385); [Haberman, 2005](https://doi.org/10.1002/j.2333-8504.2005.tb02001.x); [von Davier, 2009](https://doi.org/10.1080/15366360903117079)). The widget suggests why: a curve that differs from a 2PL only where few respondents are changes few predictions. The data can't see guessing, which isn't the same as no one guessing. The 3PL remains a reasonable choice for multiple-choice tests with many respondents near chance, and in large calibration samples.

```{ojs}
//| echo: false
quiz({
  q: "Two samples take the same 40-item test. Sample A has 5,000 respondents, nearly all well above most items' difficulty. Sample B has 800 respondents, a third of whom find most items too hard. From which sample are the lower asymptotes better estimated?",
  options: [
    {text: "Sample B", correct: true, why: "Information about $c$ comes from respondents for whom the item is too hard. B has hundreds of them on most items; A has almost none, however large it is."},
    {text: "Sample A, because it is six times larger", correct: false, why: "Size helps only through the respondents near the floor. In the widget, a thousand respondents 1.5 above the item leave $c$ almost undetermined."},
    {text: "Neither: $c$ is set by the number of options, not the data", correct: false, why: "That is a choice we can make (the next section), but a fitted 3PL estimates $c$ from the responses, and how well depends on who gave them."}
  ]
})
```

### Fix guessing from the item's structure

If we know where the floor should be, why not put it there? When an item offers $k$ equally attractive options, a respondent choosing at random is right with probability $1/k$. The simplest model with guessing sets every lower asymptote to that value and estimates nothing new:
$$\Pr(x_{ij} = 1) = \frac{1}{k} + \Big(1 - \frac{1}{k}\Big)\frac{\exp(\theta_j - b_i)}{1 + \exp(\theta_j - b_i)}.$$
It is the Rasch model with a fixed floor, with exactly the Rasch model's parameters. It is the first of three ways to add guessing that [San Martín, del Pino and De Boeck (2006)](https://doi.org/10.1177/0146621605282773) list; the others estimate one floor for all items, or one per item, as the 3PL does. [Han (2012)](https://doi.org/10.7275/f0gz-kc87) reports that fixing $c = 1/k$ gave more stable item estimates than a free 3PL without a loss of fit, on simulated and real multiple-choice data.

What does fixing the floor assume? That a respondent who doesn't know the answer picks at random. That doesn't hold if some wrong options attract more than others, if respondents can rule some out (San Martín et al. let the chance of a correct guess rise with ability for this reason), or if not knowing pushes respondents toward one answer, as in the real data below.

#### Try it: what a correct answer is worth

The same item ($b = 0$) under the Rasch model and with its floor fixed at chance. Choose the chance rate, and how far below the item a respondent sits.

```{ojs}
//| echo: false
viewof fg_c = Inputs.radio(new Map([["0.25 (four options)", 0.25], ["0.5 (two options)", 0.5]]), {value: 0.5, label: "Fixed lower asymptote"})
viewof fg_gap = Inputs.range([0.5, 4], {value: 3, step: 0.1, label: "Respondent's distance below the item"})
```

```{ojs}
//| echo: false
Plot.plot({
  width: 700, height: 300, x: {label: "Ability θ (item at 0)"}, y: {label: "P(correct)", domain: [0, 1]},
  color: {legend: true, domain: ["Rasch", "Rasch, c fixed"], range: [palette.light, palette.main]},
  marks: [
    Plot.ruleY([fg_c], {stroke: palette.guide, strokeDasharray: "4,3"}),
    Plot.ruleX([0, -fg_gap], {stroke: palette.rule}),
    Plot.line(grid(-5, 5).flatMap(t => [
      {theta: t, p: p4pl(t, 0), curve: "Rasch"},
      {theta: t, p: p4pl(t, 0, 1, fg_c), curve: "Rasch, c fixed"}]), {x: "theta", y: "p", stroke: "curve", strokeWidth: 2.5})
  ]
})
```

```{ojs}
//| echo: false
fg_r0 = p4pl(0, 0) / p4pl(-fg_gap, 0)
fg_r1 = p4pl(0, 0, 1, fg_c) / p4pl(-fg_gap, 0, 1, fg_c)
md`A correct answer is **${fg_r0.toFixed(1)}** times as likely at θ = 0 as at θ = −${fg_gap.toFixed(1)} under the Rasch model, but only **${fg_r1.toFixed(2)}** times as likely with the floor at ${fg_c}. That likelihood ratio is how strongly one correct answer favours the higher ability.`
```

With two options and a gap of 3, one correct answer favours the higher ability by a factor of 10.5 under the Rasch model and by 1.43 with the floor at 0.5. A coin would give that answer half the time. With a floor of 0.25 its worth falls less (a factor of 2.2). Wrong answers lose nothing: the floor multiplies the probability of a wrong answer by the same $1 - c$ at every ability, so the ratio between two abilities is unchanged. Problem 1 asks you to show it.

```{ojs}
//| echo: false
quiz({
  q: "With the lower asymptote fixed at 0.5, an item that 55% of respondents answer correctly is, compared with its Rasch difficulty,",
  options: [
    {text: "much harder: 55% is barely above what random pressing gives", correct: true, why: "Under the fixed floor the logistic part only has to explain the 5 points above 0.5, so the item sits well above most respondents. The real data below show this in the extreme."},
    {text: "about as hard: fixing $c$ doesn't touch difficulties", correct: false, why: "It does. The difficulty is where the logistic part crosses one half, and with a floor at 0.5 the whole curve is squeezed between 0.5 and 1."},
    {text: "easier, because some of the correct answers are guesses", correct: false, why: "Guesses explain some of the correct answers, which leaves fewer to be explained by ability, so the item looks harder, not easier."}
  ]
})
```

### Guessing may belong to respondents

The 3PL and the fixed floor both give guessing to items: every respondent far below an item has the same chance of a lucky answer. But a respondent who has stopped trying, or is racing a clock, may answer items they could solve, quickly and at chance. [Schnipke and Scrams (1997)](https://doi.org/10.1111/j.1745-3984.1997.tb00516.x) modelled response times as a mixture of rapid guesses and slower solution behaviour; [Wise and Kong (2005)](https://doi.org/10.1207/s15324818ame1802_2) measured engagement by how many of a respondent's responses come too quickly to reflect reading the item; and [Wise (2017)](https://doi.org/10.1111/emip.12165) reviews how rapid guesses are identified and argues that they shouldn't count toward the score.

::: {.callout-tip appearance="simple"}
**Recall** from [Item response data and the IRW](irw-data.qmd): when a table records response times, they sit in an `rt` column, in seconds, beside `resp`.
:::

If guessing belongs to respondents, an item's lower asymptote spreads a few respondents' behaviour over everyone. [Response time and the speed-accuracy tradeoff](response-time.qmd) models times and responses together.

### Priors keep item estimates in bounds

::: {.callout-tip appearance="simple"}
**Recall** from [Estimating abilities](ability-estimation.qmd): with a prior $g(\theta)$, the posterior is proportional to the likelihood times the prior. The EAP estimate, the posterior mean, is pulled toward the prior's mean (shrinkage), most where the likelihood is flattest: at the ends of the scale and on short tests.
:::

Nothing in that argument needs $\theta$ to be an ability. An item parameter can have a prior too, and its posterior is again likelihood times prior ([Mislevy, 1986](https://doi.org/10.1007/bf02293979); [Swaminathan & Gifford, 1986](https://doi.org/10.1007/BF02295598); [Harwell & Baker, 1991](https://doi.org/10.1177/014662169101500409)). The usual choices respect each parameter's range. A slope is positive, so $\log a$ gets a normal prior: a **lognormal** on $a$. A lower asymptote lies between 0 and 1, so it gets a **beta** prior; beta($\alpha$, $\beta$) has mean $\alpha/(\alpha + \beta)$ and counts for roughly $\alpha + \beta$ responses, so beta(2, 18) says "about 0.1, worth about 20 responses". Domingue et al. (2024) used a lognormal(0, 1) and a beta(2, 17) to keep their small-sample fits converging. Where data are plentiful the prior hardly matters; where they are thin, which for $c$ is nearly always, the estimate sits near the prior. That makes priors a reasonable choice whenever a sample is small for the number of item parameters. `mirt` finds the posterior mode; a full Bayesian treatment is in [Johnson and Sinharay (2017)](https://doi.org/10.1201/b19166-13).

#### Try it: prior × likelihood = posterior for a slope

One item ($b = 0$, no floor), answered by $n$ respondents with known abilities, from a fixed sample of 2,000. The prior is lognormal: $\log a$ is normal with mean 0 and the SD you choose. Each curve is scaled to a peak of 1.

```{ojs}
//| echo: false
viewof sp_n = Inputs.range([5, 500], {value: 20, step: 5, label: "Respondents (n)"})
viewof sp_a = Inputs.range([0.5, 3], {value: 1.5, step: 0.1, label: "True slope a"})
viewof sp_sd = Inputs.range([0.25, 2], {value: 1, step: 0.05, label: "Prior SD of log a"})
```

```{ojs}
//| echo: false
sp = slopePosterior(sp_n, sp_a, sp_sd)
Plot.plot({
  width: 700, height: 300, x: {label: "Slope a", domain: [0, 6]}, y: {label: "Scaled to a peak of 1", domain: [0, 1.05]},
  color: {legend: true, domain: ["Likelihood", "Prior", "Posterior"], range: [palette.light, palette.guide, palette.main]},
  marks: [
    Plot.ruleX([sp_a], {stroke: palette.rule}),
    Plot.line(sp.as.flatMap((a, k) => a > 6 ? [] : [
      {a, y: sp.L[k], curve: "Likelihood"}, {a, y: sp.prior[k], curve: "Prior"}, {a, y: sp.post[k], curve: "Posterior"}]),
      {x: "a", y: "y", stroke: "curve", strokeWidth: 2.5})
  ]
})
```

```{ojs}
//| echo: false
md`With ${sp_n} respondents: MLE **${Number.isFinite(sp.mle) ? sp.mle.toFixed(2) : "none (the likelihood keeps rising)"}**, posterior mode ${sp.map.toFixed(2)}, posterior mean **${sp.eap.toFixed(2)}**; the true slope is ${sp_a.toFixed(1)} (grey line).`
```

With 20 respondents the maximum-likelihood slope is 3.17, more than twice the truth, and the posterior mode is 2.16: the prior pulls the estimate back toward 1. With 5 respondents the likelihood never turns down, as for a perfect response pattern, and the posterior still has a peak (1.23). With 300, the three estimates agree within a few hundredths. Tighten the prior's SD to 0.25 and at 20 respondents the prior wins: the posterior mode is 1.15, now below the truth. Loosen it to 2 and the mode moves most of the way back toward the MLE (2.47). A prior is an assumption, and a tight one in the wrong place pulls every estimate toward the wrong place. This is shrinkage again, now for items.

```{ojs}
//| echo: false
quiz({
  q: "A 3PL is fitted with a beta(2, 18) prior on every lower asymptote to a test on which nearly every respondent is far above every item. What will the estimated $c$s look like?",
  options: [
    {text: "Close to one another and to the prior, whatever the respondents did", correct: true, why: "The likelihood is nearly flat in $c$ when no one is near the floor, so the posterior is nearly the prior. The vocabulary test below shows this."},
    {text: "Close to 0, because nobody guessed", correct: false, why: "The data can't tell whether anybody guessed: nobody is where guessing would show. Without a prior, estimates often do land near 0; with one, they land near the prior."},
    {text: "Scattered from 0 to 0.5, as without a prior", correct: false, why: "Scatter is what a flat likelihood can give without a prior. The prior is what removes it."}
  ]
})
```

### Judge it by prediction

Fixing $c$ and adding priors are both assumptions, which the model's own fit statistics take for granted. The check that doesn't is the one from [Model fit and out-of-sample prediction](fit-prediction.qmd): fit with and without the assumption, and compute the IMV on held-out responses.

## Simulate

The code below simulates a five-option test from the 3PL (20 items, every $c = 0.2$) for 300 respondents, then fits the 2PL, the 3PL without priors, and the 3PL with a lognormal(0, 1) prior on the slopes and a beta(4, 16) prior on the lower asymptotes (mean 0.2, the chance rate). It reports each model's error (RMSE) against the truth. The first run takes 20–30 seconds while R and `mirt` load, and the fits up to a minute more.

With the seed as written, the 3PL without priors has a slope as large as 6.8 (the true slopes are around 1), a slope error of 1.74 and a median $c$ of 0.29. With the priors, the slope error is 0.31, the difficulty error falls from 1.04 to 0.51, and the median $c$ is 0.17. The 2PL's difficulties are off by 1.28. All three give abilities that correlate 0.76 to 0.78 with the truth. With `np <- 3000` (in your own R session) the 3PL's slope error falls to 0.32 on its own, the priors still help (0.22), and the abilities (0.82 for all three) still hardly notice the model. These priors are centred on the true $c$, which flatters them; try `expbeta, 2, 18` (mean 0.1) to see what a prior in the wrong place costs.

```{webr}
{{< include code/guessing-priors-sim.R >}}
```

[Download this code](code/guessing-priors-sim.R) to run it in your own R session.

## With real data

```{r}
#| include: false
knitr::read_chunk("code/guessing-priors-irw.R")
options(digits = 7)  # R's default, so the page prints what a fresh R session prints
```

### A two-button reading task

Our main example is [roar_lexical](https://itemresponsewarehouse.org/tables/roar_lexical/), from the lexical decision task of the Rapid Online Assessment of Reading ([Yeatman et al., 2021](https://doi.org/10.1038/s41598-021-85907-x)): children and adolescents saw a string of letters and pressed one of two buttons to say whether it was a real word, under a time limit. Its job is to be a test whose chance rate is known exactly: 0.5. A response is 1 when the string was classified correctly, and every response has a time. [Download the code](code/guessing-priors-irw.R) to run every chunk yourself; it needs `mirt` ([Chalmers, 2012](https://doi.org/10.18637/jss.v048.i06)).

```{r fetch-roar}
#| echo: true
#| code-fold: show
```

There are 123 respondents, each classifying the same 500 strings (half real words), with no missing responses and one wave. Overall accuracy is 0.80.

```{ojs}
//| echo: false
predict({
  q: "Chance on this task is 0.5. If we fix every lower asymptote at 0.5 instead of 0, the number of parameters stays the same. How closely will the two models' ability estimates agree?",
  choices: ["Correlation near 1: they rank respondents almost identically", "Correlation around 0.7", "Correlation near 0: fixing chance reorders respondents"],
  answer: "Correlation near 1: they rank respondents almost identically",
  reveal: "The abilities correlate 0.97. What changes is the items: the SD of the difficulties rises from 0.8 to 2.0, and the log likelihood improves by about 200 with no extra parameters. Check the output below."
})
```

```{r fit-roar}
#| echo: true
#| code-fold: show
```

The two models rank respondents almost identically (the abilities correlate 0.97), and the fixed floor fits better, with a log likelihood of −25,380 against −25,579 for the same number of parameters. The scale spreads out: the SD of ability goes from 0.92 to 1.67, and of difficulty from 0.8 to 2.0, since the logistic part now explains only the correct answers above 0.5.

```{r items-roar}
#| echo: true
#| code-fold: show
#| fig-width: 6
#| fig-height: 4
```

The hardest items show where the spread comes from. Seventeen strings, 11 real words and 6 made-up ones, are classified correctly by fewer than half the respondents, below what random pressing would give; the real word *napery* by 14%. A respondent who doesn't know a word doesn't press at random: they say it isn't a word, and a made-up string that looks like a word gets called one. A floor at 0.5 can't describe an item below chance, so the fixed-floor model sends the five hardest to difficulties between 5.4 and 14.4. That is the model's assumption, that not knowing means choosing at random, meeting items for which it doesn't hold.

```{r cv-roar}
#| echo: true
#| code-fold: show
```

Out of sample, the fixed floor beats the Rasch model in all five folds, by an IMV of 0.0023 on average: a fifth of the 0.01 benchmark for the 2PL over the Rasch model, with no extra parameters. Among held-out responses from the quarter of respondents with the lowest accuracy, the IMV is 0.057; among everyone else it is slightly negative (−0.0017). It helps with a few respondents and costs a little with the rest.

```{r rt-roar}
#| echo: true
#| code-fold: show
```

The response times say who those respondents are. Of all responses, 3.1% came in under 0.3 seconds, with an accuracy of 0.51 against 0.81 for the rest. They aren't spread evenly: 11 of the 123 respondents gave more than 10% of their responses that fast, and the four with the most gave 40–47%, with overall accuracies of 0.50 to 0.55. Across respondents, the share of rapid responses and accuracy have a rank correlation of −0.66. On this task, guessing looks more like something a handful of respondents did than something every item invites.

So for a test like this one, with a known chance rate, I'd fix the floor rather than estimate it: it costs no parameters and predicts new responses better. But I'd read it as a rough description of a few respondents, check the items that fall below chance, and look at response times before trusting it. Problem 2 asks what changes when the rapid responses are set aside.

### Priors when the sample is small

Our contrast table is [vocabulary_iq](https://itemresponsewarehouse.org/tables/vocabulary_iq/), the Vocabulary IQ Test from Open Psychometrics ([Open Psychometrics, 2018](http://openpsychometrics.org/_rawdata/)). Each of its 45 items lists five words, and the respondent picks the two with the same meaning, so a random pair is right one time in ten. According to the test's codebook, respondents were told not to guess, that a wrong answer cost 0.35 points, and that they could answer "don't know". Its job is to show what priors do for a 3PL fitted to a small sample. We drop items 46–75, a personality survey collected alongside the test. A response is 1 for the correct pair; "don't know" is missing, and we keep the respondents who never chose it.

```{r fetch-viqt}
#| echo: true
#| code-fold: show
```

"Don't know" accounts for 13% of responses, and 2,802 of the 12,173 respondents never chose it. For them the items are easy: the median item is answered correctly by 90%, and the hardest by half.

```{r full-viqt}
#| echo: true
#| code-fold: show
```

Fitted to all 2,802, the 3PL puts every $c$ between 0 and 0.09, with a median of 0.000: on an easy test that discouraged guessing, almost nobody is where a floor would show. With the priors (lognormal(0, 1) on the slopes, beta(2, 18) on $c$, centred on the chance rate of 0.1), every $c$ lands between 0.045 and 0.066. Even with 2,802 respondents, the lower asymptotes are mostly the prior.

```{r sub-viqt}
#| echo: true
#| code-fold: show
#| fig-width: 5
#| fig-height: 4.5
```

Fitted to 300 respondents without priors, the slopes have an SD of 2.07 and three exceed 5, while the full-sample slopes run from 0.44 to 4.74. With the priors, the SD is 0.79 and none exceeds 5, and the slopes' error against the full-sample 3PL drops from 1.75 to 0.94. The difficulties go the other way (0.37 to 0.69), shifting to match $c$s that the prior moved from about 0 to about 0.05.

```{r repeat-viqt}
#| echo: true
#| code-fold: show
```

Over five more subsamples the pattern holds (medians above): the priors tame the slopes and cost something on the difficulties. The reference is itself a 3PL without priors, so "error" means distance from what 2,802 respondents give, not from the truth.

## Problems

1. **What an answer is worth.** Under the Rasch model with $c$ fixed at $1/k$, write the log likelihood for $\theta$ from one correct response and from one incorrect response, and differentiate each. Show that a correct response's derivative is smaller than under the Rasch model at every $\theta$ and an incorrect one's is unchanged. What does that mean for a respondent who gets most items wrong?
2. **Real data with a twist.** In `roar_lexical`, set responses faster than 0.3 seconds to missing and refit the Rasch model and the fixed-floor model. How do the difficulty SDs and the out-of-sample IMV change, and does the fixed floor still help the lowest-accuracy quarter?
3. **Real data: the items below chance.** Refit both models without the 17 `roar_lexical` items answered correctly by fewer than half the respondents and compare the out-of-sample IMV with the lesson's. What would a lower asymptote below 0.5 mean for those items? (PS4#3.)
4. **Judgment.** A four-option test gives 3PL lower asymptotes between 0.00 and 0.05. A colleague concludes that nobody guesses on it. What would you check before agreeing?
5. **Design.** You must fit a 3PL to 200 respondents on a 30-item, four-option test. Choose priors for the slopes and lower asymptotes, defend them, and check them by simulation. How would you know if your prior on $c$ were in the wrong place? (PS4#4.)
6. **Challenge (open): guessing as a respondent trait.** Write down a model with two classes of respondent: engaged (the Rasch model) and rapid-guessing (correct with probability $1/k$). What data would identify the classes: responses, response times, or both? I don't know of a single answer; [Response time and the speed-accuracy tradeoff](response-time.qmd) and [Nominal response and multiple-choice models](nominal.qmd) have pieces of one.

```{r}
#| output: asis
ask_claude("guessing-priors", "It covers the lower asymptote (guessing parameter) of the 3PL and why it is hard to estimate, fixing the lower asymptote at the chance rate in a Rasch model, rapid guessing and response times, lognormal and beta priors on item slopes and lower asymptotes in mirt, and judging these choices by out-of-sample prediction with the IMV, with R analyses of a lexical decision task (roar_lexical) and a vocabulary test (vocabulary_iq).")
```

## Going further

- Domingue, B. W., Kanopka, K., Kapoor, R., Pohl, S., Chalmers, R. P., Rahal, C., & Rhemtulla, M. (2024). The InterModel Vigorish as a lens for understanding (and quantifying) the value of item response models for dichotomously coded items. *Psychometrika*, 89(3), 1034–1054. <https://doi.org/10.1007/s11336-024-09977-2>. §3 and Table 1: how much the 3PL adds to prediction over the 2PL in simulation, including when it generated the data.
- Maris, G., & Bechger, T. (2009). On interpreting the model parameters for the three parameter logistic model. *Measurement: Interdisciplinary Research and Perspectives*, 7(2), 75–88. <https://doi.org/10.1080/15366360903070385>. What the 3PL's parameters can and can't be read as.
- von Davier, M. (2009). Is there need for the 3PL model? Guess what? *Measurement: Interdisciplinary Research and Perspectives*, 7(2), 110–114. <https://doi.org/10.1080/15366360903117079>. A commentary in the same issue.
- Haberman, S. J. (2005). Identifiability of parameters in item response models with unconstrained ability distributions. *ETS Research Report Series*, 2005(2), i–22. <https://doi.org/10.1002/j.2333-8504.2005.tb02001.x>. Identification in general form.
- Han, K. T. (2012). Fixing the c parameter in the three-parameter logistic model. *Practical Assessment, Research, and Evaluation*, 17, Article 1. <https://doi.org/10.7275/f0gz-kc87>. Fixing $c = 1/k$, on simulated and real multiple-choice data.
- San Martín, E., del Pino, G., & De Boeck, P. (2006). IRT models for ability-based guessing. *Applied Psychological Measurement*, 30(3), 183–203. <https://doi.org/10.1177/0146621605282773>. Ways to add guessing to the 1PL, and a model in which the chance of a correct guess rises with ability.
- Schnipke, D. L., & Scrams, D. J. (1997). Modeling item response times with a two-state mixture model: A new method of measuring speededness. *Journal of Educational Measurement*, 34(3), 213–232. <https://doi.org/10.1111/j.1745-3984.1997.tb00516.x>. Rapid guesses and solution behaviour as two states.
- Wise, S. L., & Kong, X. (2005). Response time effort: A new measure of examinee motivation in computer-based tests. *Applied Measurement in Education*, 18(2), 163–183. <https://doi.org/10.1207/s15324818ame1802_2>. Engagement measured from response times.
- Wise, S. L. (2017). Rapid-guessing behavior: Its identification, interpretation, and implications. *Educational Measurement: Issues and Practice*, 36(4), 52–61. <https://doi.org/10.1111/emip.12165>. How rapid guesses are found, and what to do with them.
- Mislevy, R. J. (1986). Bayes modal estimation in item response models. *Psychometrika*, 51(2), 177–195. <https://doi.org/10.1007/bf02293979>. Priors on item parameters as well as abilities.
- Swaminathan, H., & Gifford, J. A. (1986). Bayesian estimation in the three-parameter logistic model. *Psychometrika*, 51(4), 589–601. <https://doi.org/10.1007/BF02295598>. Priors that keep 3PL estimates inside the parameter space.
- Harwell, M. R., & Baker, F. B. (1991). The use of prior distributions in marginalized Bayesian item parameter estimation: A didactic. *Applied Psychological Measurement*, 15(4), 375–389. <https://doi.org/10.1177/014662169101500409>. Mislevy's procedure, step by step.
- Yeatman, J. D., Tang, K. A., Donnelly, P. M., Yablonski, M., Ramamurthy, M., Karipidis, I. I., Caffarra, S., Takada, M. E., Kanopka, K., Ben-Shachar, M., & Domingue, B. W. (2021). Rapid online assessment of reading ability. *Scientific Reports*, 11, 6396. <https://doi.org/10.1038/s41598-021-85907-x>. The lexical decision task behind `roar_lexical`.
- Chalmers, R. P. (2012). mirt: A multidimensional item response theory package for the R environment. *Journal of Statistical Software*, 48(6). <https://doi.org/10.18637/jss.v048.i06>. The package, including its `guess` argument and `PRIOR` syntax.
- From the IRW: the [guessing vignette](https://itemresponsewarehouse.org/vignettes/guessing.html).
- Johnson, M. S., & Sinharay, S. (2017). Bayesian estimation. In W. J. van der Linden (Ed.), *Handbook of item response theory: Vol. 2. Statistical tools* (pp. 237–258). Chapman and Hall/CRC. <https://doi.org/10.1201/b19166-13>. A handbook chapter on Bayesian estimation for item response models.
- Sinharay, S. (2017). Bayesian model fit and model comparison. In W. J. van der Linden (Ed.), *Handbook of item response theory: Vol. 2. Statistical tools* (pp. 379–394). Chapman and Hall/CRC. <https://doi.org/10.1201/b19166-19>. A handbook chapter on Bayesian checks of model fit and Bayesian model comparison.

```{r}
#| output: asis
data_sources("guessing-priors")
```

## For instructors

- **Source:** EDUC 252 class 4 (slides c4, 33 and 60–61), problem set 4 (#3 and #4), and the code in `ben-domingue/252`: `ps4/guessing.R`, `ps4/priors.R`.
- **Timing:** the core ideas and widgets take about 50 minutes; the real-data section works as a 30-minute lab. Locally, the `roar_lexical` cross-validation takes a few minutes and the `vocabulary_iq` subsamples several more; start them first.
- **Changed from 252:** PS4#3 fixes guessing on the LEVANTE vocabulary task, which isn't a tokenless IRW table; this lesson uses `roar_lexical`, the alternative PS4#3's code names, and adds the out-of-sample comparison (the PS4#3 bonus), the items below chance and the response times. PS4#4 puts priors on a 3PL for `gilbert_meta_2`; here they go on `vocabulary_iq` at 300 respondents, where they have more to do, and the design half of PS4#4 is problem 5.
- **Held back:** response times modelled jointly with accuracy ([Response time and the speed-accuracy tradeoff](response-time.qmd), which reuses `roar_lexical`); skipped and "don't know" responses as information ([IRTree models](irtrees.qmd)); the wrong options ([Nominal response and multiple-choice models](nominal.qmd)); EM with priors on item parameters ([Estimating item parameters](item-estimation.qmd)). Deep dive #22 (which guessing correction helps across the IRW's multiple-choice tables) is not built here.
- **Solutions:** held back in `solutions/guessing-priors.qmd` (not published).
