---
title: "Response time and the speed-accuracy tradeoff"
---

```{r}
#| output: asis
source("_course.R")
lesson_header("response-time")
```

## What this is for

*To be written.*

## Goals

## Core ideas

## Worked example

## Problems

## Going further

- van der Linden, W. J. (2016). Lognormal response-time model. In W. J. van der Linden (Ed.), *Handbook of item response theory: Vol. 1. Models* (pp. 261–282). Chapman and Hall/CRC. <https://doi.org/10.1201/9781315374512>. A handbook chapter on the lognormal model for response times.
- van der Linden, W. J., & Fox, J.-P. (2016). Joint hierarchical modeling of responses and response times. In W. J. van der Linden (Ed.), *Handbook of item response theory: Vol. 1. Models* (pp. 481–500). Chapman and Hall/CRC. <https://doi.org/10.1201/9781315374512>. A handbook chapter on modelling responses and response times together.
- van der Linden, W. J. (2017). Test speededness and time limits. In W. J. van der Linden (Ed.), *Handbook of item response theory: Vol. 3. Applications* (pp. 249–266). Chapman and Hall/CRC. <https://doi.org/10.1201/9781315117430>. A handbook chapter on test speededness and time limits.

```{r}
#| output: asis
data_sources("response-time")
```

## For instructors

