---
title: "Unfolding models"
---

```{r}
#| output: asis
source("_course.R")
lesson_header("unfolding")
```

```{webr}
#| context: setup
{{< include code/_webr-setup.R >}}
```

```{ojs}
//| echo: false
import {grid, dnorm, p2pl, rng, palette} from "./widgets/irt.js"
import {hcm, pcm3, best2PL, oneRun, shuffle} from "./widgets/unfolding.js"
import {quiz, predict} from "./widgets/quiz.js"
```

## What this is for

Two different accounts of how a respondent answers an item:

- **Dominance.** The more of the trait a respondent has, the more likely they are to agree or answer correctly: each item's curve only rises. Every model in this course so far has assumed this.
- **Ideal point.** A respondent agrees most with statements close to their own position, and less the farther a statement is from it, in either direction. A statement that stakes out a middle position is rejected from both sides, by respondents who find it too strong and by those who find it too weak, so its curve rises and then falls.

For tests of ability dominance is natural; for attitude statements it can fail. [Where CTT breaks](ctt-limits.qmd) found two such statements among eight about capital punishment and left them there. This lesson picks them up: what an unfolding response process leaves in the data, the models written for it (the hyperbolic cosine model, a folded partial credit model, and the GGUM), and what happens when we fit them to the same eight statements. The short answer: the patterns look very much like unfolding, and yet, with 54 respondents and most statements near the ends, a dominance model describes the responses just as well.

## Goals

By the end of this lesson you should be able to:

- say what distinguishes an ideal-point (unfolding) response process from a dominance process, and sketch the item curve each implies;
- say what unfolding leaves in the data: runs of agreement in the right order, negative correlations between the ends, and middle statements that correlate with nothing;
- derive the hyperbolic cosine model by folding a three-category partial credit model, and place the GGUM and `mirt`'s ideal-point item beside it;
- fit dominance and ideal-point models in R, and find a statement order with MUDFOLD;
- explain why the two kinds of model are hard to tell apart when statements sit near the ends of the scale, and design statements that could tell them apart.

## Core ideas

### Dominance and ideal points

::: {.callout-tip appearance="simple"}
**Recall** from [Where CTT breaks](ctt-limits.qmd): keyed toward support, six of the eight capital-punishment statements rose with the score on the others. The two middle statements did not. DONTBELIEV rose from 0.44 among the most opposed to 0.75 in the middle and fell back to 0.33 and 0.38; WISHNOTNEC reached 1.00 at a score of 4 and dropped to 0.85 among the most in favour. That item curves rise with the sum score was an assumption, and these two statements are where it gave way.
:::

What response process would produce a curve that rises and then falls? In a **dominance** process, a respondent agrees when their level exceeds what the statement asks for, as a student solves an item when their ability exceeds its difficulty. Every model since [The Rasch model](rasch.qmd) has had this form: the item curve rises with $\theta$. In an **ideal-point** process, a respondent agrees when the statement is close to their own position and disagrees when it is too far away, in either direction. [Thurstone (1928)](https://doi.org/10.1086/214483) scaled attitude statements by where they sit on a continuum, and [Coombs (1964)](https://openlibrary.org/works/OL1361275W) put respondents and stimuli on one scale, with preference falling off with distance. Each respondent's preferences are the scale folded at their own position; recovering the scale is **unfolding**.

Under an ideal-point process the item curve is single-peaked: highest for respondents at the statement's location $b$, lower on both sides. So a disagreement is ambiguous. Someone who rejects "I think capital punishment is necessary but I wish it were not" (WISHNOTNEC in [Andrich, 1988](https://doi.org/10.1177/014662168801200105), worded as in the `mudfold` help page) may oppose capital punishment outright, or may support it without regret. [Drasgow, Chernyshenko and Stark (2010)](https://doi.org/10.1111/j.1754-9434.2010.01273.x) argue that respondents answer many attitude and personality items this way.

#### Try it: dominance and ideal points

One statement, located at $b$. The dominance curve is a 2PL with slope 1.5; the ideal-point curve is the hyperbolic cosine model introduced below, at the same location. The filled markers are a respondent at the chosen $\theta$; the open marker is a respondent as far from the statement on the other side.

```{ojs}
//| echo: false
viewof dp_b = Inputs.range([-2, 2], {value: 0.5, step: 0.1, label: "Statement location b"})
viewof dp_th = Inputs.range([-4, 4], {value: -1.5, step: 0.1, label: "Respondent θ"})
```

```{ojs}
//| echo: false
dp_tau = 1.5
dp_grid = grid(-4, 4, 241)
dp_mirror = 2 * dp_b - dp_th
Plot.plot({
  width: 700, height: 280, x: {label: "θ", domain: [-4, 4]}, y: {label: "P(agree)", domain: [0, 1]},
  color: {legend: true, domain: ["Dominance (2PL)", "Ideal point"], range: [palette.main, palette.contrast]},
  marks: [
    Plot.ruleX([dp_b], {stroke: palette.guide, strokeDasharray: "4,3"}),
    Plot.line(dp_grid.flatMap(t => [{theta: t, p: p2pl(t, dp_b, 1.5), m: "Dominance (2PL)"}, {theta: t, p: hcm(t, dp_b, dp_tau), m: "Ideal point"}]),
      {x: "theta", y: "p", stroke: "m", strokeWidth: 2.5}),
    Plot.dot([{theta: dp_th, p: p2pl(dp_th, dp_b, 1.5)}, {theta: dp_th, p: hcm(dp_th, dp_b, dp_tau)}], {x: "theta", y: "p", r: 5, fill: palette.ink}),
    Plot.dot([{theta: dp_mirror, p: p2pl(dp_mirror, dp_b, 1.5)}, {theta: dp_mirror, p: hcm(dp_mirror, dp_b, dp_tau)}].filter(d => Math.abs(d.theta) <= 4),
      {x: "theta", y: "p", r: 5, stroke: palette.ink, fill: palette.white})
  ]
})
```

```{ojs}
//| echo: false
md`At θ = ${dp_th.toFixed(1)}, P(agree) is **${p2pl(dp_th, dp_b, 1.5).toFixed(2)}** under the dominance curve and **${hcm(dp_th, dp_b, dp_tau).toFixed(2)}** under the ideal-point curve. At θ = ${dp_mirror.toFixed(1)}, as far on the other side of the statement, it is **${p2pl(dp_mirror, dp_b, 1.5).toFixed(2)}** and **${hcm(dp_mirror, dp_b, dp_tau).toFixed(2)}**.`
```

Under the dominance curve the two respondents answer very differently (0.05 and 0.95 at the defaults); under the ideal-point curve they answer alike (0.37 each), since only the distance to $b$ matters. A disagreement doesn't say which side the respondent is on. This ideal-point curve never reaches 1: even respondents at the statement's location sometimes disagree.

```{ojs}
//| echo: false
quiz({
  q: "Under an ideal-point process, a respondent disagrees with a statement located at $b = 0.5$. What can you conclude about their $\\theta$?",
  options: [
    {text: "That it is probably some distance from 0.5, on one side or the other", correct: true, why: "The ideal-point curve is symmetric about $b$, so a disagreement is equally consistent with $\\theta$ well below and well above the statement. Other responses have to say which side."},
    {text: "That it is probably below 0.5", correct: false, why: "That is the dominance reading. Under an ideal-point process, respondents far above the statement disagree too."},
    {text: "Nothing at all", correct: false, why: "It does say something: respondents near $b$ agree most often, so a disagreement makes a $\\theta$ near 0.5 less likely."}
  ]
})
```

### What unfolding leaves in the data

Suppose statements and respondents sit on one line, and each respondent agrees with the statements near them. Put the statements in their order along the line. Then each respondent's agreements form one unbroken run: a respondent in the middle agrees with a few middle statements and rejects both ends. Stack the respondents in order and the runs move down the matrix as a diagonal band, a **parallelogram** in Coombs's (1964) terms. A dominance process leaves a different shape: in difficulty order, each respondent agrees with everything up to some point, a triangle, as in [Guttman's (1944)](https://doi.org/10.2307/2086306) scalogram.

Two more consequences follow. Statements at opposite ends are rarely both endorsed, since no respondent sits near both, so they correlate negatively. A statement in the middle is endorsed by respondents who are neither strongly for nor against, so it correlates with nothing.

#### Try it: the parallelogram

These are the 54 respondents and eight capital-punishment statements from the data section. Each row is a respondent, each column a statement; filled cells are agreements. Rows are sorted by the average position of each respondent's agreements in the chosen order.

```{ojs}
//| echo: false
viewof pg_order = Inputs.radio(["Shuffled", "Andrich's order"], {value: "Shuffled", label: "Order the statements"})
viewof pg_seed = Inputs.range([1, 50], {value: 1, step: 1, label: "Shuffle seed"})
```

```{ojs}
//| echo: false
pg_names = ["HIDEOUS", "LIFESACRED", "INEFFECTIV", "DONTBELIEV", "WISHNOTNEC", "MUSTHAVEIT", "DETERRENT", "CRIMDESERV"]
pg_ord = pg_order === "Andrich's order" ? pg_names : shuffle(pg_names, rng(pg_seed).unif)
pg_rows = transpose(mud_rows)
pg_pos = pg_rows.map((r, i) => {
  const hits = pg_ord.map((k, j) => r[k] === 1 ? j : null).filter(j => j !== null);
  return {i, centre: hits.reduce((u, v) => u + v, 0) / hits.length, run: oneRun(r, pg_ord)};
}).sort((u, v) => u.centre - v.centre)
pg_rank = new Map(pg_pos.map((d, k) => [d.i, k]))
pg_cells = pg_rows.flatMap((r, i) => pg_ord.map(k => ({row: pg_rank.get(i), st: k, x: r[k]})))
Plot.plot({
  width: 560, height: 520, marginLeft: 30, marginBottom: 90, padding: 0.04,
  x: {label: null, domain: pg_ord, tickRotate: -40}, y: {label: null, axis: null},
  color: {domain: [0, 1], range: [palette.white, palette.main]},
  marks: [Plot.cell(pg_cells, {x: "st", y: "row", fill: "x", stroke: palette.rule, strokeWidth: 0.3})]
})
```

```{ojs}
//| echo: false
md`In this order, **${pg_pos.filter(d => d.run).length}** of the 54 patterns are a single unbroken run of agreements.`
```

Shuffled (seed 1), the matrix looks scattered, and 5 patterns are single runs; across the 50 seeds most orders give fewer than 10 and none more than 25. In Andrich's order a band appears, and 39 of the 54 are single runs: respondents who oppose capital punishment agree with the first three or four statements, those in favour with the last three or four, and a few in between with the middle.

::: {.callout-tip appearance="simple"}
**Recall** from [Classical test theory and reliability](ctt-reliability.qmd): scales mix items worded in both directions, and each has to be reverse-keyed before the items are summed. One Mach IV item, m19, had an item-rest correlation of −0.05 keyed one way and 0.05 the other, so the data couldn't say which way it ran.
:::

Unfolding produces items like m19 by design. A statement in the middle of the scale is neither forward nor reversed: reversing it turns a peak into a trough, and neither correlates with the total. Keying presumes a dominance process.

```{ojs}
//| echo: false
quiz({
  q: "Six statements about a policy sit in order from strongly against (1) to strongly in favour (6), and responses follow an ideal-point process. Which pattern of agreement (1 = agree) is the model *least* likely to produce?",
  options: [
    {text: "1 0 0 0 0 1", correct: true, why: "Agreeing with both extremes and nothing between breaks the run: no single position is near both ends of the scale."},
    {text: "0 0 1 1 0 0", correct: false, why: "A single run in the middle: a respondent near statements 3 and 4."},
    {text: "1 1 1 0 0 0", correct: false, why: "A single run at one end: a respondent who is fairly strongly against."}
  ]
})
```

### Models for single-peaked curves

How do we write down a single-peaked curve? One route starts from a model we already have.

::: {.callout-tip appearance="simple"}
**Recall** from [Models for polytomous responses](polytomous.qmd): under the partial credit model, category $k$ has probability proportional to $\exp\big(\sum_{v \le k}(\theta - b_v)\big)$, and each adjacent pair of categories is a Rasch item. With three categories and ordered steps, the middle one is the most likely response between its two steps.
:::

A disagreement under an ideal-point process has two sources: disagreeing from below (the statement goes too far) and disagreeing from above (it doesn't go far enough). Treat these as categories 0 and 2 of a three-category PCM, with agreement as category 1 between them, and put the two steps at $b - \tau$ and $b + \tau$. We observe only agree or disagree, so categories 0 and 2 are merged, and what remains is
$$\Pr(\text{agree} \mid \theta) = \frac{e^{\tau}}{e^{\tau} + 2\cosh(\theta - b)}.$$
This is the **hyperbolic cosine model** (HCM) of [Andrich and Luo (1993)](https://doi.org/10.1177/014662169301700307); see also [Andrich (1997)](https://doi.org/10.1007/978-1-4757-2691-6_23). It is symmetric about $b$ and peaks there, at $e^{\tau}/(e^{\tau} + 2)$; $\tau$ sets how wide the region of agreement is. The Go deeper callout derives it. (Andrich and Luo's notation differs; we write $\tau$ to keep $\lambda$ for loadings.)

#### Try it: folding the partial credit model

```{ojs}
//| echo: false
viewof fo_b = Inputs.range([-2, 2], {value: 0, step: 0.1, label: "Location b"})
viewof fo_tau = Inputs.range([0, 3], {value: 1.5, step: 0.1, label: "τ (half the gap between the steps)"})
viewof fo_fold = Inputs.toggle({label: "Fold: merge the two disagree categories", value: false})
```

```{ojs}
//| echo: false
fo_th = grid(-5, 5, 241)
fo_rows = fo_fold
  ? fo_th.flatMap(t => { const p = pcm3(t, fo_b, fo_tau); return [{theta: t, p: p[1], c: "agree"}, {theta: t, p: p[0] + p[2], c: "disagree (either side)"}]; })
  : fo_th.flatMap(t => pcm3(t, fo_b, fo_tau).map((p, k) => ({theta: t, p, c: ["disagree from below", "agree", "disagree from above"][k]})))
Plot.plot({
  width: 700, height: 280, x: {label: "θ", domain: [-5, 5]}, y: {label: "Probability", domain: [0, 1]},
  color: {legend: true, domain: ["disagree from below", "agree", "disagree from above", "disagree (either side)"], range: [palette.light, palette.contrast, palette.main, palette.guide]},
  marks: [Plot.ruleX([fo_b - fo_tau, fo_b + fo_tau], {stroke: palette.guide, strokeDasharray: "4,3"}),
          Plot.line(fo_rows, {x: "theta", y: "p", stroke: "c", strokeWidth: 2.5})]
})
```

```{ojs}
//| echo: false
md`Peak probability of agreeing, at θ = ${fo_b.toFixed(1)}: **${hcm(fo_b, fo_b, fo_tau).toFixed(2)}** from the hyperbolic cosine formula, **${pcm3(fo_b, fo_b, fo_tau)[1].toFixed(2)}** from the three categories. The dashed lines are the PCM's steps.`
```

Unfolded, the three curves are an ordinary PCM item. Folded, the two outer curves add up to a trough, and agreement is the single-peaked curve that remains; both routes give the same peak (0.69 at the defaults). With $\tau = 0$ the peak is $1/3$; widen the steps and it rises toward 1.

The HCM has a location and a width for each statement, and no slope. The **generalized graded unfolding model** (GGUM; [Roberts, Donoghue & Laughlin, 2000](https://doi.org/10.1177/01466216000241001)) adds a discrimination for each statement and allows graded agreement (strongly disagree to strongly agree) as well as agree/disagree. It rests on the same fold: each observed category merges a response from below with one from above. The `GGUM` R package fits it ([Tendeiro & Castro-Alvarez, 2019](https://doi.org/10.1177/0146621618772290)). `mirt` offers a simpler single-peaked item, `itemtype = "ideal"`, following [Maydeu-Olivares, Hernández and McDonald (2006)](https://doi.org/10.1207/s15327906mbr4104_2): $\Pr(\text{agree}) = \exp\{-\tfrac{1}{2}a^2(\theta - b)^2\}$, a bell that reaches 1 at $\theta = b$. That peak is an assumption: every respondent located at a statement agrees with it. The HCM and GGUM let the peak sit lower.

::: {.callout-tip appearance="simple"}
**Recall** from [The Rasch model](rasch.qmd): the model assumed that a single $\theta$ drives every response and that responses are locally independent given $\theta$, and it paired these with an item curve that rises.
:::

The unfolding models keep the first two assumptions and drop the third. Given $\theta$, agreements are still independent and the likelihood is still a product over statements; only the shape of each curve changes.

There is also a nonparametric counterpart. **MUDFOLD** ([van Schuur, 1992](https://doi.org/10.1093/pan/4.1.41); the `mudfold` package, [Balafas, Krijnen, Post & Wit, 2020](https://doi.org/10.32614/rj-2020-002)) looks for the longest order of statements in which triples behave as unfolding predicts, and reports a scalability coefficient $H$: one minus the ratio of observed to expected violations, so that 1 means none. It asks for no curve shape, only an order, which makes it a reasonable first look when the question is whether the statements can be placed on one line at all.

::: {.callout-note collapse="true"}
## Go deeper: the hyperbolic cosine model is a folded PCM

**Assumptions:** one statement at location $b$; a respondent at $\theta$ gives one of three latent responses, disagree from below (0), agree (1) or disagree from above (2), following a partial credit model with steps $b - \tau$ and $b + \tau$; we observe only agree (1) or disagree (0 or 2).

**The PCM.** With $\Pr(x = k) \propto \exp\big(\sum_{v \le k}(\theta - b_v)\big)$, $b_1 = b - \tau$ and $b_2 = b + \tau$:
$$\Pr(0) \propto 1, \qquad \Pr(1) \propto e^{\theta - b + \tau}, \qquad \Pr(2) \propto e^{(\theta - b + \tau) + (\theta - b - \tau)} = e^{2(\theta - b)}.$$

**Fold.** The odds of agreeing against either kind of disagreement are
$$\frac{\Pr(1)}{\Pr(0) + \Pr(2)} = \frac{e^{\theta - b + \tau}}{1 + e^{2(\theta - b)}} = \frac{e^{\tau}}{e^{-(\theta - b)} + e^{\theta - b}} = \frac{e^{\tau}}{2\cosh(\theta - b)},$$
dividing top and bottom by $e^{\theta - b}$. So $\Pr(\text{agree}) = e^{\tau}/\big(e^{\tau} + 2\cosh(\theta - b)\big)$.

**Properties.** $\cosh$ is even and smallest at 0, so the curve is symmetric about $b$ and peaks there, at $e^{\tau}/(e^{\tau} + 2)$. As $|\theta - b|$ grows, $2\cosh(\theta - b) \approx e^{|\theta - b|}$, and the log odds of agreeing approach $\tau - |\theta - b|$: on each side, far from $b$, the curve is a Rasch curve with slope 1, facing the other way. That is why a dominance model can stand in for it when respondents are all on one side of a statement.

**What's lost.** Under the PCM the sum score is sufficient ([Models for polytomous responses](polytomous.qmd)). The fold merges categories scored 0 and 2, so the sum of the latent categories is no longer observed, and the number of agreements is not sufficient for $\theta$: which statements a respondent agrees with matters, not only how many.
:::

### Choosing between them is hard

If the curves are so different, why is the choice hard? Because the data only see a curve where there are respondents. Almost every respondent sits to the right of a statement near the "against" end, on the falling side, where the ideal-point curve is a decreasing S. A 2PL with a negative slope draws the same S, and one with a positive slope matches a statement near the "in favour" end. Only in the middle, with respondents on both sides, does the rise and fall show.

#### Try it: can a 2PL mimic an ideal-point curve?

The solid curve is a hyperbolic cosine item; the dashed curve is the 2PL that best matches it for respondents distributed as $N(0, 1)$ (shaded).

```{ojs}
//| echo: false
viewof mi_b = Inputs.range([-3, 3], {value: 2.5, step: 0.1, label: "Statement location b"})
viewof mi_tau = Inputs.range([0.5, 3], {value: 2, step: 0.1, label: "τ"})
```

```{ojs}
//| echo: false
mi_fit = best2PL(t => hcm(t, mi_b, mi_tau))
mi_gap = Math.max(...mi_fit.th.map((t, i) => Math.abs(t) <= 2 ? Math.abs(mi_fit.y[i] - mi_fit.fit[i]) : 0))
Plot.plot({
  width: 700, height: 280, x: {label: "θ", domain: [-4, 4]}, y: {label: "P(agree)", domain: [0, 1]},
  color: {legend: true, domain: ["Ideal point (HCM)", "Best 2PL"], range: [palette.contrast, palette.main]},
  marks: [
    Plot.areaY(mi_fit.th.map(t => ({theta: t, d: dnorm(t) / dnorm(0) * 0.5})), {x: "theta", y: "d", fill: palette.light, fillOpacity: 0.4}),
    Plot.line(mi_fit.th.map((t, i) => ({theta: t, p: mi_fit.y[i], m: "Ideal point (HCM)"})), {x: "theta", y: "p", stroke: "m", strokeWidth: 2.5}),
    Plot.line(mi_fit.th.map((t, i) => ({theta: t, p: mi_fit.fit[i], m: "Best 2PL"})), {x: "theta", y: "p", stroke: "m", strokeWidth: 2.5, strokeDasharray: "6,4"})
  ]
})
```

```{ojs}
//| echo: false
md`Best 2PL slope: **${mi_fit.a.toFixed(2)}**. Largest gap between the curves for θ between −2 and 2: **${mi_gap.toFixed(2)}**.`
```

At $b = 2.5$ the best 2PL (slope 0.93) stays within 0.03 of the ideal-point curve wherever respondents are. At $b = 0$ the gap is 0.22 and the best 2PL is flat: it can't bend. At $b = -2.5$ the 2PL fits again, with slope −0.93. A dominance model with free signs describes end statements about as well as an ideal-point model; the two separate on middle statements, when enough respondents sit on both sides. [Stark, Chernyshenko, Drasgow and Williams (2006)](https://doi.org/10.1037/0021-9010.91.1.25) make this comparison for personality items.

```{ojs}
//| echo: false
quiz({
  q: "You want to know whether a set of attitude statements follows an ideal-point process. Which design gives the data the best chance to say?",
  options: [
    {text: "Several statements written to sit in the middle of the scale, and many respondents near the middle", correct: true, why: "Middle statements with respondents on both sides are where the rise and fall show. At the ends a sloped 2PL matches the ideal-point curve."},
    {text: "Many strongly worded statements at both ends", correct: false, why: "End statements see mostly one side of their curve, which a 2PL with a positive or negative slope matches closely."},
    {text: "A larger sample, whatever the statements", correct: false, why: "More respondents sharpen the estimates, but if no statement has respondents on both sides, the two models still predict nearly the same responses."}
  ]
})
```

A single-peaked curve can only be told from a sloped one where there are respondents on both sides of the peak.

## Simulate

The code below simulates 500 respondents answering 20 statements from $b = -2.5$ to $2.5$ with `mirt`'s ideal-point curve ($a = 1.5$), fits the 2PL and the ideal-point model, and compares AIC ([Akaike, 1974](https://doi.org/10.1109/TAC.1974.1100705)). Then it keeps the four statements at each end and fits both again. The first run takes 20–30 seconds while R and `mirt` load, and the fits about another minute and a half in the browser (15 seconds in local R). Ideal-point fits can stop at a poor local maximum, so the code starts each statement at the average 2PL score of the respondents who agreed with it; remove `pars = sv` to see what happens without that start.

With the seed as written, the ideal-point model wins easily with all 20 statements (AIC 7,405 against 8,844) and recovers the locations (correlation 0.998). With only the end statements the two are one AIC point apart (2,296 and 2,297), the 2PL with positive slopes at one end and negative at the other. In the plot, the 2PL misses the middle statement and matches the end one. Try `np <- 200`, or `ends <- c(1:2, 9:12, 19:20)`, which puts four statements back in the middle.

```{webr}
{{< include code/unfolding-sim.R >}}
```

[Download this code](code/unfolding-sim.R) to run it in your own R session.

## With real data

```{r}
#| include: false
knitr::read_chunk("code/unfolding-irw.R")
options(digits = 7)  # R's default, so the page prints what a fresh R session prints
```

### Capital punishment

The main example returns from [Where CTT breaks](ctt-limits.qmd): [andrich_mudfold](https://itemresponsewarehouse.org/tables/andrich_mudfold/), in which 54 graduate students in an introductory measurement course agreed (1) or disagreed (0) with eight statements about capital punishment ([Andrich, 1988](https://doi.org/10.1177/014662168801200105)), as distributed with the `mudfold` package ([Balafas, Krijnen & Wit, 2017](https://doi.org/10.32614/CRAN.package.mudfold)). Its job is to be a set of attitude statements written to span the scale, on which we can ask whether an unfolding model is needed. Reading the statements is part of the analysis, so here are all eight, in Andrich's order from most against to most in favour, quoted from [Andrich (1988)](https://doi.org/10.1177/014662168801200105) in the wording of the `ANDRICH` help page of the `mudfold` package:

| Code | Statement |
|---|---|
| HIDEOUS | "Capital punishment is one of the most hideous practices of our time" |
| LIFESACRED | "The state cannot teach the sacredness of human life by destroying it" |
| INEFFECTIV | "Capital punishment is not an effective deterrent to crime" |
| DONTBELIEV | "I do not believe in capital punishment but I am not sure it is not necessary" |
| WISHNOTNEC | "I think capital punishment is necessary but I wish it were not" |
| MUSTHAVEIT | "Until we find a more civilized way to prevent crime we must have capital punishment" |
| DETERRENT | "Capital punishment is justified because it does act as a deterrent to crime" |
| CRIMDESERV | "Capital punishment gives the criminal what he deserves" |

The response is agreement, and there is nothing to key: which statements run "forward" is the question. With 54 respondents, every estimate below is noisy. [Download the code](code/unfolding-irw.R) to run every chunk yourself.

```{r fetch-mud}
#| echo: true
#| code-fold: show
```

Everyone agreed with between two and six statements. The correlations have the unfolding signature:

```{r cor-mud}
#| echo: true
#| code-fold: show
```

The ends correlate negatively (LIFESACRED and CRIMDESERV at −0.84), and DONTBELIEV correlates with nothing: its largest correlation with another statement, in absolute value, is 0.18. WISHNOTNEC, the other middle statement, correlates 0.85 with MUSTHAVEIT and −0.61 with LIFESACRED, so its agreers sit mostly on the "in favour" side. Now the runs, in Andrich's order and in every other:

```{r runs-mud}
#| echo: true
#| code-fold: show
```

In Andrich's order 39 of the 54 patterns are a single run. Across all 40,320 orders the average is 5.4, and the best any order does is 40: Andrich's order with CRIMDESERV and DETERRENT swapped, or with LIFESACRED and INEFFECTIV swapped as well. The data put the statements close to the order their content suggests.

```{r mudfold-mud}
#| echo: true
#| code-fold: show
```

MUDFOLD keeps all eight statements in nearly the same order, differing only in how it arranges the three statements against and the two in favour. $H = 0.64$: the triples show about a third as many violations of the unfolding order as independent responses would (131 observed against 363 expected, in the package's summary). So the order and the runs look like unfolding. Does a model built for unfolding fit better than one that isn't?

```{ojs}
//| echo: false
predict({
  q: "We fit the Rasch model, the 2PL, `mirt`'s ideal-point model and the GGUM to the eight statements. Which has the lowest AIC?",
  choices: ["The GGUM, clearly", "The ideal-point model, clearly", "The 2PL, tied with the GGUM", "The Rasch model"],
  answer: "The 2PL, tied with the GGUM",
  reveal: "AIC is 422.0 for the 2PL and 422.1 for the GGUM, which has eight more parameters; `mirt`'s ideal-point model is at 436.5 and the Rasch model at 594.6. Check the output below."
})
```

```{r models-mud}
#| echo: true
#| code-fold: show
```

The Rasch model is far behind: agreeing with LIFESACRED and agreeing with CRIMDESERV can't both rise with the same $\theta$ and slope. The 2PL and the GGUM tie on AIC, and BIC, which charges more per parameter, prefers the 2PL (453.8 against 469.8). `mirt`'s ideal-point model has as many parameters as the 2PL and fits worse. The 2PL stopped at its iteration limit rather than converging; its parameters show why.

```{r params-mud}
#| echo: true
#| code-fold: show
```

The 2PL does what the mimicry widget did: negative slopes for the three statements against (from −1.44 to −7.23) and positive slopes for the three in favour, reaching 23.81 for CRIMDESERV. That slope keeps growing because the item is close to a step, and a steeper step always fits a little better. DONTBELIEV gets a slope of −0.11, close to flat. The GGUM puts the statements in Andrich's groups: the three against at −2.32 to −2.30, DONTBELIEV at −0.11, WISHNOTNEC and MUSTHAVEIT at 0.85 and 0.92, and the two in favour at 2.37 and 2.54. Three discriminations sit at the package's upper bound of 10. `mirt`'s ideal-point model puts DONTBELIEV's peak at −48.98, far outside the range of respondents: its way of drawing a nearly flat curve.

```{r curves-mud}
#| echo: true
#| code-fold: show
#| fig-width: 8
#| fig-height: 4.5
```

For the six end statements the 2PL and GGUM curves nearly coincide. `mirt`'s bell, which must reach 1 at its peak, puts the peak inside the range (near −1 for LIFESACRED and INEFFECTIV) and falls away on the far side, which is part of why it fits worse. DONTBELIEV is nearly flat under the 2PL and `mirt`'s model, and a low, wide hill under the GGUM, peaking at about 0.5. For WISHNOTNEC and MUSTHAVEIT both ideal-point models let agreement fall off among the respondents most in favour, where the 2PL keeps rising. What if we give just one statement an ideal-point curve?

```{r mixed-mud}
#| echo: true
#| code-fold: show
```

No single statement's ideal-point curve brings the AIC below the all-2PL model's 422.0; the closest are DETERRENT (422.5) and WISHNOTNEC (422.6), then DONTBELIEV (423.4). The two models' scores for the respondents correlate at 0.97.

So which is it? The statements were written to span the scale, the order and the runs look very much like unfolding, and [Where CTT breaks](ctt-limits.qmd) found WISHNOTNEC turning over. Yet on AIC and BIC a dominance model with free signs does as well. Six of the eight statements sit near an end, where a sloped 2PL traces the ideal-point curve, and the two in the middle have too few respondents on their far side to show the fall. With 54 respondents the data can't tell the models apart statement by statement. What would? More statements written for the middle, and more respondents near them.

<!-- TODO (Ben, F32): the lesson's first-person verdict, on hold with Ben. Candidate (Claude's reading of PS9#3 and the AIC result): "If the statements weren't written to span the middle, the dominance model is enough." It belongs here, after the assumptions above; confirm or reword. -->

### Party activists

For contrast, [eurpar2_mudfold](https://itemresponsewarehouse.org/tables/eurpar2_mudfold/): 1,786 activists in European political parties, each of whom picked the two of six party families in the European Parliament they most preferred, in 1980 ([van Schuur, 1984](https://openlibrary.org/works/OL13368770W); [Post, 1992](https://openlibrary.org/works/OL3986205W)), also distributed with `mudfold`. Its job is to show unfolding data in which a sum score carries no information at all.

```{r fetch-eur}
#| echo: true
#| code-fold: show
```

Every activist picked exactly two parties, so everyone's sum score is 2, and whatever the data say about where an activist stands is in *which* two they chose. Under an ideal-point process an activist picks the two parties nearest their own position, which should be neighbours on the scale.

```{r pairs-eur}
#| echo: true
#| code-fold: show
```

```{r mudfold-eur}
#| echo: true
#| code-fold: show
```

MUDFOLD orders five of the six parties from left to right (communists, social democrats, liberals, Christian democrats, conservatives), with $H = 0.79$. The four commonest pairs are the four neighbouring pairs in that order (437 + 341 + 217 + 202 = 1,197 of the 1,786 choices); communists with conservatives came up 4 times. MUDFOLD leaves out the democratic progressives, who were paired with the liberals (124), the social democrats (106) and the conservatives (77) alike, so it finds no single place for them on the line. A second dimension would produce that pattern, and so would a party chosen for reasons other than left and right; these data can't say which. Pick-$k$ choices are close relatives of forced-choice questionnaires, which [Brown (2016)](https://doi.org/10.1007/s11336-014-9434-9) models under dominance and ideal-point processes alike.

## Problems

1. **Folding the PCM.** Starting from a three-category PCM with steps $b - \tau$ and $b + \tau$, derive $\Pr(\text{agree}) = e^{\tau}/(e^{\tau} + 2\cosh(\theta - b))$. Show that the curve peaks at $\theta = b$, find the height of the peak, and show that far from $b$ the log odds of agreeing fall by one for each unit of $\theta$.
2. **Read, then fit (PS9#3).** Read the eight capital-punishment statements and, for each, say before looking at any output whether you'd model it with a dominance curve or an ideal-point curve, and why. Then compare with the curves and the mixed-model AICs. Where did the data agree with your reading, and where couldn't they say?
3. **Judgment.** DONTBELIEV correlates with no other statement above 0.18, gets a 2PL slope of −0.11, and gets a GGUM location near 0. With 54 respondents, what could change your mind about whether it follows an ideal-point process? What would you look for in a new sample, and roughly how large would it need to be?
4. **Design.** Write eight statements about a policy you know well, for 300 respondents, so that a dominance model and an unfolding model could be told apart. Where on the scale must the statements sit, and where must the respondents be? Check your design in the simulation: set `delta` to your planned locations and `np` to 300, and see whether the ideal-point model wins.
5. **Order from pairs.** In `eurpar2_mudfold`, recover a left-to-right order of the six parties from the pair counts alone (for example, treat the counts as similarities and scale them in one dimension with `cmdscale`). Compare with MUDFOLD's order. Where do the democratic progressives land, and does that fit what the counts showed?
6. **Challenge (open).** Are some personality items ideal-point items? [Stark et al. (2006)](https://doi.org/10.1037/0021-9010.91.1.25) argue that some are. Pick an IRW personality table with Likert items and say how you would find out; then try it, fitting the GRM and the GGUM (`mirt`'s `itemtype = "ggum"`) and comparing the curves. Or model the rankings in [franco_2024_unfolding](https://itemresponsewarehouse.org/tables/franco_2024_unfolding/). I don't know the answer for either.

```{r}
#| output: asis
ask_claude("unfolding", "It covers unfolding (ideal-point) models for attitude statements: dominance versus ideal-point response processes, what unfolding leaves in the data (runs of agreement, the parallelogram, negative correlations between the ends), the hyperbolic cosine model as a folded partial credit model, the generalized graded unfolding model (GGUM), mirt's ideal-point item type, MUDFOLD and its H coefficient, and why a 2PL with negative slopes can mimic an ideal-point model when statements sit near the ends of the scale, with fits in R (mirt, GGUM, mudfold) to Andrich's capital-punishment statements and European party activists' pick-two-of-six choices.")
```

## Going further

- Thurstone, L. L. (1928). Attitudes can be measured. *American Journal of Sociology*, 33(4), 529–554. <https://doi.org/10.1086/214483>. Attitude statements placed on a continuum by where they sit.
- Coombs, C. H. (1964). *A theory of data*. Wiley. <https://openlibrary.org/works/OL1361275W>. Unfolding, the parallelogram and the geometry of preference. [Flament's (1966) review](https://doi.org/10.1007/bf02289474) summarizes it.
- Guttman, L. (1944). A basis for scaling qualitative data. *American Sociological Review*, 9(2), 139–150. <https://doi.org/10.2307/2086306>. The scalogram, the dominance pattern the parallelogram is set against.
- Andrich, D. (1988). The application of an unfolding model of the PIRT type to the measurement of attitude. *Applied Psychological Measurement*, 12(1), 33–51. <https://doi.org/10.1177/014662168801200105>. The capital-punishment statements and a model for them.
- Andrich, D., & Luo, G. (1993). A hyperbolic cosine latent trait model for unfolding dichotomous single-stimulus responses. *Applied Psychological Measurement*, 17(3), 253–276. <https://doi.org/10.1177/014662169301700307>. The hyperbolic cosine model.
- Andrich, D. (1997). A hyperbolic cosine IRT model for unfolding direct responses of persons to items. In W. J. van der Linden & R. K. Hambleton (Eds.), *Handbook of modern item response theory* (pp. 399–414). Springer. <https://doi.org/10.1007/978-1-4757-2691-6_23>. The model as a handbook chapter.
- Andrich, D. (2016). Hyperbolic cosine model for unfolding responses. In W. J. van der Linden (Ed.), *Handbook of item response theory: Vol. 1. Models* (pp. 353–367). Chapman and Hall/CRC. <https://doi.org/10.1201/9781315374512>. A handbook chapter on the hyperbolic cosine unfolding model, by its originator.
- Roberts, J. S., Donoghue, J. R., & Laughlin, J. E. (2000). A general item response theory model for unfolding unidimensional polytomous responses. *Applied Psychological Measurement*, 24(1), 3–32. <https://doi.org/10.1177/01466216000241001>. The GGUM.
- Roberts, J. S. (2016). Generalized graded unfolding model. In W. J. van der Linden (Ed.), *Handbook of item response theory: Vol. 1. Models* (pp. 369–390). Chapman and Hall/CRC. <https://doi.org/10.1201/9781315374512>. A handbook chapter on the generalized graded unfolding model.
- Tendeiro, J. N., & Castro-Alvarez, S. (2019). GGUM: An R package for fitting the generalized graded unfolding model. *Applied Psychological Measurement*, 43(2), 172–173. <https://doi.org/10.1177/0146621618772290>. The package used here.
- Maydeu-Olivares, A., Hernández, A., & McDonald, R. P. (2006). A multidimensional ideal point item response theory model for binary data. *Multivariate Behavioral Research*, 41(4), 445–472. <https://doi.org/10.1207/s15327906mbr4104_2>. A bell-shaped ideal-point item for binary data, the form of `mirt`'s `"ideal"` itemtype.
- van Schuur, W. H. (1992). Nonparametric unidimensional unfolding for multicategory data. *Political Analysis*, 4, 41–74. <https://doi.org/10.1093/pan/4.1.41>. MUDFOLD, and unfolding without a parametric curve.
- Balafas, S. E., Krijnen, W. P., Post, W. J., & Wit, E. C. (2020). mudfold: An R package for nonparametric IRT modelling of unfolding processes. *The R Journal*, 12(1), 49–75. <https://doi.org/10.32614/rj-2020-002>. The `mudfold` package, which distributes both data sets.
- Drasgow, F., Chernyshenko, O. S., & Stark, S. (2010). 75 years after Likert: Thurstone was right! *Industrial and Organizational Psychology*, 3(4), 465–476. <https://doi.org/10.1111/j.1754-9434.2010.01273.x>. The case for ideal-point models of attitude and personality items.
- Stark, S., Chernyshenko, O. S., Drasgow, F., & Williams, B. A. (2006). Examining assumptions about item responding in personality assessment: Should ideal point methods be considered for scale development and scoring? *Journal of Applied Psychology*, 91(1), 25–39. <https://doi.org/10.1037/0021-9010.91.1.25>. Dominance and ideal-point models compared on personality items (problem 6).
- Brown, A. (2016). Item response models for forced-choice questionnaires: A common framework. *Psychometrika*, 81(1), 135–160. <https://doi.org/10.1007/s11336-014-9434-9>. Choice formats under dominance and ideal-point processes.
- Chalmers, R. P. (2012). mirt: A multidimensional item response theory package for the R environment. *Journal of Statistical Software*, 48(6). <https://doi.org/10.18637/jss.v048.i06>. The package, whose `itemtype` list includes the ideal-point, GGUM and hyperbolic cosine items.
- Akaike, H. (1974). A new look at the statistical model identification. *IEEE Transactions on Automatic Control*, 19(6), 716–723. <https://doi.org/10.1109/TAC.1974.1100705>. The AIC.

```{r ojs-data}
#| echo: false
# The capital-punishment responses for the parallelogram widget.
ojs_define(mud_rows = mud)
```

```{r}
#| output: asis
data_sources("unfolding")
```

## For instructors

- **Source:** EDUC 252 class 9 (slides c9, 13–14, "The unfolding model"), problem set 9 (#3, "Dominant or unfolding?") and problem set 2 (#5c, now in [Where CTT breaks](ctt-limits.qmd)); code in `ben-domingue/252`: `ps9/unfold.R`. Slide 14's pointer to Andrich's handbook chapter is Andrich (1997).
- **Timing:** the core ideas and widgets take about 45 minutes; the two analyses work as a 30-minute lab. The real-data code takes about a minute in local R, most of it the GGUM fit.
- **Changed from 252:** PS9#3 compared the Rasch and ideal-point models item by item. The lesson adds the 2PL (so the dominance model can have negative slopes), the GGUM, MUDFOLD, and mixed models with one ideal-point statement at a time. The `GGUM` package doesn't run in the browser, so the GGUM is fitted only in the real-data section, at render time; Simulate uses `mirt`.
- **Reuse:** `andrich_mudfold` also appears in [Where CTT breaks](ctt-limits.qmd), where the two middle statements' curves turn over against a score on the other six.
- **Held back:** multidimensional unfolding; rankings and paired comparisons as unfolding data (`franco_2024_unfolding`, problem 6 only); ideal-point models for personality items (problem 6 only); graded (Likert) unfolding beyond a mention of the GGUM.
- **Item text:** the eight statements are quoted from Andrich (1988) in the wording of the `mudfold` package's `ANDRICH` help page, which writes "i" for "I" in DONTBELIEV and WISHNOTNEC; they have not been checked against the printed article. The quotation (about 110 words) is within SAGE's pre-approved reuse of short extracts from a journal article (at most 200 words or 10% of the article), with a full citation.
- **Solutions:** held back in `solutions/unfolding.qmd` (not published). Problem 6 is open.
